```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Getting to Know the Data

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what it means for a dataset to be "tabular"
* Explain what a "feature" is
* Explain what an "observation" is
* Explain what it means for a dataset to be "tidy"
* Describe some widely-used data types
* Convert columns to appropriate data types
:::

<!--
:::{.callout-important}
We teach this lecture with [a slide deck][slides]. This chapter does not
necessarily cover everything covered in lecture.
:::

[slides]: https://docs.google.com/presentation/d/1e_LcUT1tU5WsUYnrTQBjTIOL6zIrcCsXBjUigTBKA6w/edit?usp=sharing
-->

This chapter explains where to start when you explore a dataset. Whether it's
data you collected yourself or a new and unfamiliar dataset, the steps to go
through and details to check at the beginning of an exploration are the same.
Some of these are qualitative: now is a good time to think about how and why
the data were collected, and whether they're appropriately representative to
address the questions you want to answer. Others are computational, such as
whether the dataset's structure matches expectations and is convenient for
analysis. The chapter also briefly discusses fixes for common problems, with
references to where you can learn more.


## Be a Data Biographer

The first step in getting to know a dataset is understanding who collected it
and why they did so.

* Who collected the dataset?
* Why was the dataset collected?

Next, think about the concrete details of how the dataset was collected. What
was the collection procedure?

From which population is the dataset drawn? Does the dataset include the entire
population? If the dataset is a sample of the population, how was the sample
taken? Is it possible that subgroups within the population were undersampled or
left out, either by coincidence, due to natural constraints, or intentionally?

Is the dataset also associated with other populations?

Consider whether any special privacy or security precautions are necessary for
the dataset.

Finally, think about the impact of the dataset and analyses based on it. Who
benefits? Who, if anyone, might be harmed?

:::{.callout-important}
Keep these questions in mind throughout your time working with the dataset.
When you arrive at results you want to share, revisit these questions before
doing so.
:::

A **data biography** is a document that attempts to answer these kinds of
questions and provide any other background and context relevant to a dataset.
When collecting a dataset, providing a biography as one part of the
documentation is a best practice. If the dataset you're working with doesn't
include a biography, try to write one yourself! Doing so will force you to
confront the questions above, and give you a document you can use to help
explain the results of your analyses. In some cases, sharing the data biography
might help coworkers or colleagues who also want to work with the dataset.


## Structure Matters

The structure of a dataset---its shape and organization---has enormous
influence on how difficult it will be to analyze.

Datasets can take many different forms, or **structures**, depending on what
they represent and the information they contain. For instance, a dataset that
represents:

<!-- TODO: pictures! or links to data? e.g., night heron range -->
<!-- Same dataset, different structures? -->

* People in a family, such as the Roosevelts, might be organized as a (family)
  tree, with connections from parents to children.
* Monthly population estimates for a fish species in the Sacramento River might
  be a series of numbers.
* Relationships between cats in a small town might be a graph (or network),
  with a node for each cat and an edge (connection) between each pair of cats
  that have met.
* Individual consumer reviews for several different vacuums might be a table
  with a row for each review and columns to identify the individual, vacuum
  model, and review score.
* Bike paths in Davis, CA might be a collection of geospatial coordinates that
  describe points, lines, and shapes.

<!-- FIXME: this is pretty abstract, can we be more concrete? -->
Different structures have tradeoffs.

Structure matters because it affects which avenues of exploration and analysis
are appropriate and realistic. It can also affect how you access and query data
in the dataset.

Sometimes, transforming a dataset to a different structure
makes analysis easier.

A **tabular** dataset is one that's structured as a table, with rows and
columns. This reader focuses on tabular datasets, since they're common in
practice and present the fewest programming challenges. Here's an example of a
tabular dataset:

| Fruit  | Quantity | Price
| :----  | -------: | ----:
| apple  | 32       | 1.49
| banana | 541      | 0.79
| pear   | 10       | 1.99

Data scientists think about datasets in terms of two components:

* A **feature** (also called a **covariate**) is measurement of something,
  usually across multiple subjects. For example, we might decide to measure the
  heights of everyone in the class. Each person in the class is a subject, and
  the height measurement is a feature. Features don't have to be quantitative.
  If we also asked each person their favorite color, then favorite color would
  be another feature in our dataset. Features are usually, but not always, the
  columns in a tabular dataset.

* An **observation** is a set of features measured for a single subject or at a
  single time. So in the preceding example, the combined height and favorite
  color measurement for one student is one observation. Observations are
  usually, but not always, the rows in a tabular dataset.

Every dataset has features and observations, regardless of whether the dataset
is tabular.


A **data dictionary** is documentation that describes each feature in a dataset
in detail. For a tabular dataset, a data dictionary typically describes the
unit of observation (in other words, what a subject is) and what each column
measures. The description explain the range of realistic values, regardless of
whether the feature is quantitative or qualitative. For quantitative features,
it should also specify the units of measurement.

<!--
## Tidy Datasets {#sec-tidy-data}

In 2014, [Hadley Wickham][hadley] refined and formalized the conventions for
tabular datasets by [introducing the concept of tidy datasets][tidy], which
have a specific structure. A tabular dataset is **tidy** if and only if:

[hadley]: https://hadley.nz/
[tidy]: https://vita.had.co.nz/papers/tidy-data.html

1. Each observation has its own row.
2. Each feature has its own column.
3. Each value has its own cell.

These rules ensure that all of the values in a dataset are visually organized
and are easy to access in programming languages. They're specific enough to
make tidiness a convenient standard for commands that operate on tabular
datasets. They also reflect the way statisticians conventionally arrange
tabular datasets.

:::{.callout-note}
R's [Tidyverse][tidyverse] packages (see @sec-the-tidyverse) are designed from
the ground up for working with tidy datasets. Tidy datasets have also been
adopted as a standard in other software, including various packages for Python
and Julia.

[tidyverse]: https://www.tidyverse.org/
:::

When you first look at a dataset, think about what the observations are and
what the features are. If the dataset comes with documentation, it may help you
figure this out. Since this dataset is a tidy dataset, we already know each row
is an observation and each column is a feature.

@sec-reshaping-data gives examples of tidy and untidy data, as well as
explanations of how to make untidy data tidy.
-->


## Dataset: CA Least Terns {#sec-ca-least-terns}

The California least tern is a endangered subspecies of seabird that nests
along the coast of California and Mexico. The California Department of Fish and
Wildlife (CDFW) monitors least tern nesting sites across the state to estimate
breeding pairs, fledglings, and predator activity in each annual breeding
season.

![
  A California least tern. Original photo by [Mark Pavelka, U.S. Fish &
  Wildlife Service][pavelka] ([CC BY 2.0][]).
](/images/exploring-data/ca_least_tern_usfws_pacificsw.jpg){
  height=24em
  fig-alt="A gray bird with a white belly, black head, and orange beak sitting
  on a clutch of eggs."
}

[pavelka]: https://www.flickr.com/photos/usfws_pacificsw/5704890596/
[CC BY 2.0]: https://creativecommons.org/licenses/by/2.0/

The CDFW publishes most of the data it collects to the [California Open Data
portal][data.ca.gov]. The examples in this and subsequent chapters use a
cleaned 2000-2023 version of the California least tern data.

[data.ca.gov]: https://data.ca.gov/

:::{.callout-important}
[Click here][ca-least-tern] to view the 2000-2023 California least tern
dataset.

[ca-least-tern]: https://ucdavis.box.com/s/m2w5l2ebp2rey2do5lnn38y1e3u31pin
:::

:::{.callout-note
  title="Documentation for 2000-2023 California Least Tern Dataset"
  collapse=false
}

Each row in the dataset contains measurements from one year-site combination.

| Column                | Description
| --------------------: | :----------
| `year`                | Year of the breeding season
| `site_name`           | Site name
| `site_name_2013_2018` | Site name from 2013-2018
| `site_name_1988_2001` | Site name from 1988-2001
| `site_abbr`           | Abbreviated site name
| `region_3`            | Region of state: S.F. Bay, Central, or Southern (includes Ventura)
| `region_4`            | Region of state: S.F. Bay, Central, Ventura, or Southern
| `event`               | Climate events
| `bp_min`              | Reported minimum breeding pairs
| `bp_max`              | Reported maximum breeding pairs
| `fl_min`              | Reported minimum fledges
| `fl_max`              | Reported maximum fledges
| `total_nests`         | Total reported nests (maximum if a range was reported)
| `nonpred_eggs`        | Total non-predator-related mortalities of eggs
| `nonpred_chicks`      | Total non-predator-related mortalities of chicks
| `nonpred_fl`          | Total non-predator-related mortalities of fledges
| `nonpred_ad`          | Total non-predator-related mortalities of adults
| `pred_control`        | Site predator control (yes/no)
| `pred_eggs`           | Total predator-related mortalities of eggs
| `pred_chicks`         | Total predator-related mortalities of chicks
| `pred_fl`             | Total predator-related mortalities of fledges
| `pred_ad`             | Total predator-related mortalities of adults
| `pred_pefa`           | Predation by peregrine falcons (yes/no)
| `pred_coy_fox`        | Predation by coyotes or foxes (yes/no)
| `pred_meso`           | Predation by other mesocarnivores: dogs, cats, skunks, opossums, raccoons, weasels, etc. (yes/no)
| `pred_owlspp`         | Predation by owls (yes/no)
| `pred_corvid`         | Predation by corvids: ravens or crows (yes/no)
| `pred_other_raptor`   | Predation by raptors other than peregrine falcons and owls (yes/no)
| `pred_other_avian`    | Predation by birds other than raptors and corvids (yes/no)
| `pred_misc`           | Predation by other animals (yes/no)
| `total_pefa`          | Total mortalities due to peregrine falcons
| `total_coy_fox`       | Total mortalities due to coyotes and foxes
| `total_meso`          | Total mortalities due to other mesocarnivores
| `total_owlspp`        | Total mortalities due to owls
| `total_corvid`        | Total mortalities due to ravens and crows
| `total_other_raptor`  | Total mortalities due to other raptors
| `total_other_avian`   | Total mortalities due to other birds
| `total_misc`          | Total mortalities due to other animals
| `first_observed`      | Date CA least terns first observed at site
| `last_observed`       | Date CA least terns last observed at site
| `first_nest`          | Date first egg observed at site
| `first_chick`         | Date first chick observed at site
| `first_fledge`        | Date first fledge observed at site

: {.striped .hover tbl-colwidths="[25, 75]"}

The messy source dataset (with more years and more columns) is available
[here][ca-least-tern-source].

[ca-least-tern-source]: https://data.ca.gov/dataset/california-least-tern-monitoring-sites-generalized-cdfw-ds3146
:::


## Structural Summaries

When you start working with a dataset, examining its structure is one of the
first things you should do. This way you can identify and correct potential
problems, and prepare the dataset for the analyses you'd like to carry out.

Whenever you load a dataset into R, your next step should be to investigate the
data's structure. This step is important because it can help you identify
whether:

* The data was loaded correctly
* There are structural problems with the data that will make it difficult to
  use if they aren't fixed

R provides several functions for getting structural summaries of data. Some of
these are:

* `str` to get a detailed structural summary
* `head`, `tail` to preview the data
* `nrow`, `ncol`, `dim`, `length` to get dimension information
* `names`, `colnames`, `rownames` to get element names
* `class`, `typeof` to get classes and types

Let's try a few of these on the California Least Terns dataset
(@sec-ca-least-terns). For instance, the `str` function shows the classes of
the columns:

```{r}
str(terns)
```

:::{.callout-note title="See Also"}
For a reminder of R's syntax and commands, see DataLab's [R Basics][r-basics]
workshop reader.

[r-basics]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/
:::


## Types of Data {#sec-types-of-data}

Different types of data require different approaches and analysis methods. For
example, if your data are numbers, you can summarize them by computing a sum or
an average. On the other hand, if your data are categories, like genres of
music (rock, pop, classical, ...), you can't use these summarization methods
and have to do something else.

Data scientists categorize data into **types** based on sets of shared
characteristics. This makes it easier to reason about how to transform,
analyze, and present the data. Some widely-understood data types are:

* **Categorical**
    + **Nominal** - data separated into specific categories, with no order. For
      example, hair color (red, brown, blonde, ...) is categorical.
    + **Ordinal** - data separated into specific categories, with an order. For
      example, school level (elementary, middle, high, college) is ordinal.
* **Numerical**
    + **Discrete** - integers, or a finite set of decimal numbers with no
      values in between. Sometimes discrete values can also be treated as
      ordinal. For example, month as a number (1, 2, ..., 12) is discrete.
    + **Continuous** - decimal numbers. There are no specific categories, but
      there is an order. For example, height in inches is numerical.

<!--
Of course, other types of data, like graphs (networks) and natural language
(books, speech, and so on), are also possible.
-->
Categorizing data this way is useful for reasoning about which methods to apply
to which data.

:::{.callout-tip}
Because a feature measures something, we often think and talk about types of
features rather than types of individual values. For example, for a feature
that measures temperature in degrees Celsius, all of the data values will be
numeric (and probably continuous). So we can say the feature is numeric.

In a tidy dataset, each feature has its own column, so it also makes sense to
talk about the types of the columns.
:::


## Correcting Data Types

Often when you load a new dataset, some of the columns won't have the correct
data type (or class) for what you want to do. For instance, in the least terns
dataset, the `site_name`, `region_3`, and `event` columns all contain
categorical data, so they should be factors.

You can convert these columns to factors with R's `factor` function:

```{r}
terns$site_name = factor(terns$site_name)
terns$region_3 = factor(terns$region_3)
terns$event = factor(terns$event)
```

:::{.callout-note title="See Also"}
For more about factors, see the [Factors][r-basics-factors] section of
DataLab's R Basics workshop reader.

[r-basics-factors]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/data-types-structures/data-types-structures.html#sec-factors
:::

:::{.callout-tip}
There's another way we could've done this that uses only two lines of code, no
matter how many columns there are:

```{r, eval = FALSE}
cols = c("site_name", "region_3", "event")
terns[cols] = lapply(terns[cols], factor)
```

You can learn more about the `lapply` function in the [The purrr
Package][r-basics-purrr] section of DataLab's R Basics workshop reader.

[r-basics-purrr]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/organizing-code/control-flow.html#sec-the-purrr-package

You can use whichever approach is more convenient and makes more sense to you.
If there were other columns to convert, we'd go through the same steps with the
appropriate conversion function.
:::

R provides `as.` functions to convert to the most common data types. For
instance, `as.character` converts an object to a string:

```{r}
x = 3.1
class(x)

y = as.character(x)
y
class(y)
```

The `read.csv` function does a good job at identifying columns of numbers, so
it's rarely necessary to convert columns of numbers manually. However, you may
have to do this for data you got some other way (rather than loading a file).
For instance, it's common to make these conversions when scraping data from the
web.

It's also a good idea to convert categorical columns into factors with the
`factor` function and to convert columns of dates into dates. For more about
the latter, see DataLab's [Intermediate R: Cleaning & Reshaping Data][int-r]
workshop reader.

[int-r]: https://ucdavisdatalab.github.io/dwr_workshop_intermediate_r/chapters/cleaning-reshaping/01_dates-times.html

