```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Getting to Know the Data

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what it means for a dataset to be "tabular"
* Explain what a "feature" is
* Explain what an "observation" is
* Explain what it means for a dataset to be "tidy"
* Describe some widely-used data types
* Convert columns to appropriate data types
:::

<!--
:::{.callout-important}
We teach this lecture with [a slide deck][slides]. This chapter does not
necessarily cover everything covered in lecture.
:::

[slides]: https://docs.google.com/presentation/d/1e_LcUT1tU5WsUYnrTQBjTIOL6zIrcCsXBjUigTBKA6w/edit?usp=sharing
-->

This chapter explains where to start when you explore a dataset. Whether it's
data you collected yourself or a new and unfamiliar dataset, the steps to go
through and details to check at the beginning of an exploration are the same.
Some of these are qualitative: now is a good time to think about how and why
the data were collected, and whether they're appropriately representative to
address the questions you want to answer. Others are computational, such as
whether the dataset's structure matches expectations and is convenient for
analysis. The chapter also briefly discusses fixes for common problems, with
references to where you can learn more.


## Be a Data Biographer

The first step in getting to know a dataset is to understand why and how it was
collected. This will help you identify:

* Who or what the dataset represents;
* What kinds of questions you can use it to answer;
* What kinds of analysis methods might be appropriate;
* What biases or limitations it might have.

We recommend taking some time to write a brief description of the dataset in
your own words, called a **data biography**. Do this even if the dataset's
documentation already includes a description or biography, to help you organize
and clarify your understanding of the dataset.

Use the documentation and related publications as you write. If the dataset
doesn't have documentation or the documentation is incomplete, you might need
to do some sleuthing with the dataset itself (see the remainder of this
reader), online, or by contacting the people who created the dataset directly.
Keep track of any unresolved questions about the dataset as you write the
biography.

A data biography serves as a reference as you analyze the dataset and
communicate results. If you collected the dataset yourself and plan to share it
with others, the data biography is also essential documentation for helping the
audience make sense of the dataset. Even if you didn't collect the dataset, the
biography and notes you write during analysis might be helpful to colleagues or
collaborators who also want to work with the dataset.

The following subsections break down some of the kinds of questions to ask as
you write a data biography.

:::{.callout-important}
Keep the biography questions in mind throughout your time working with the
dataset. When you arrive at results you want to share, revisit the questions
before doing so.
:::

<!--
A **data biography** is a document that attempts to answer these kinds of
questions and provide any other background and context relevant to a dataset.
When collecting a dataset, providing a biography as one part of the
documentation is a best practice. If the dataset you're working with doesn't
include a biography, try to write one yourself! Doing so will force you to
confront the questions above, and give you a document you can use to help
explain the results of your analyses. In some cases, sharing the data biography
might help coworkers or colleagues who also want to work with the dataset.


* Context and Motivation
* Collection Strategy
* Precautions and Impact
-->

### Context and Motivation

First, think about where the dataset came from---its **provenance**. Ask:

* [ ] Who collected the dataset?
* [ ] Why did they collect the dataset?
* [ ] What do they claim the dataset represents?

These questions can help you identify potential biases or limitations of the
dataset. Consider whether the person or group who collected the dataset has any
conflicts of interest. While conflicts of interest are not necessarily
indicative of a problem with a dataset, they do suggest it deserves extra
scrutiny for biases.

As an example, consider a dataset collected by a light bulb manufacturer about
the operating characteristics of their bulbs. Most likely, their motivation is
to convince potential customers that their bulbs are superior to competitors,
although it might also be to assess how they can improve their bulbs. In other
words, they have an obvious conflict of interest. As a result, they might use a
testing process that doesn't replicate typical household conditions,
selectively include results from their tests, or aggregate the results in a way
that makes the bulbs appear to have especially long lifespans or low power
consumption.

When thinking about what a dataset represents, data scientists (and
statisticians) usually think in terms of the **population** from which the data
were drawn. The population might consist of people, but it could also consist
of other animals, plants, inanimate objects, or something else. The members of
the population are called **subjects** or sometimes **units of observation**.
Some datasets represent or are associated with multiple populations.


### Collection Strategy

Next, think about how the dataset was collected. Ask:

* [ ] What was the collection procedure?
* [ ] When and where did they collect the dataset?
* [ ] Does the dataset consist of the entire population? If not, how were
  subjects selected?

The question about population gets at whether the dataset is a **sample**: a
subset of subjects selected from a population. When it's not possible to
collect data about an entire population---whether due to cost, difficulty,
ethical concerns, or something else---we might instead take a sample. The hope
is that the sample is representative, meaning results for the sample generalize
to the entire population.

How well a sample represents a population depends on how it was collected and
how large it is, among other things. A **simple random sample**, where subjects
are selected from the population randomly and independently, often generalizes
well because random sampling preserves most statistical properties of the
population. In experiments, researchers sometimes use more sophisticated
sampling procedures to control for various biases.

As you think about the population and sample procedure, it's especially
important to consider whether there are subgroups within the population that
might be underrepresented or left out entirely, whether by coincidence, due to
natural constraints, or intentionally. These can limit the generalizability of
results from the dataset.


### Precautions and Impact

Finally, think about how using the dataset might affect other people or things.
Ask:

* [ ] Who is impacted by use or analysis of the dataset? What are the potential
  consequences?
* [ ] Are any special privacy or security precautions necessary?

These questions are intended to help you reflect on potential ethical and legal
concerns when working with the dataset, as well as your own responsibilities to
the associated population(s) and sampled subjects.

The **CARE Principles** [@carroll2020], which were originally developed to
protect Indigenous peoples' rights and interests with respect to Indigenous
data, provide a basis for reasoning about the impacts of data work on
communities more generally. The principles are:

* Collective benefit: data and analyses should provide benefit to the community
  as a whole and reflect community values.
* Authority to control: community members have a right to and should be
  actively involved in the stewardship and governance of their data.
* Responsibility: data workers have a responsibility to engage and cultivate
  respectful relationships with community for the benefit of the community.
* Ethics: protecting the rights and well-being of community members should be
  the highest priority; community members must be allowed to assess benefits
  and harms themselves.

The CARE Principles have been adopted by organizations such as the [Research
Data Alliance][rda] and the [Smithsonian Institution][smithsonian].

[rda]: https://www.rd-alliance.org/
[smithsonian]: https://www.si.edu/

:::{.callout-note title="See Also"}
_The Turing Way_ [-@turing-way], an online textbook developed by data
scientists and other researchers from all over the world, incorporates a wide
variety of perspectives about best practices for collecting, analying, and
communicating results about data. [This chapter][turing-way-care] discusses the
CARE Principles.

[turing-way-care]: https://book.the-turing-way.org/reproducible-research/rdm/rdm-care/

_Data Feminism_ [@dignazio2020] is a textbook that explores how to approach
data work in an ethical and equitable way by drawing from feminist principles.
The publisher provides [a free, online version of the book][data-feminism] in
addition to selling print and eBook versions.

[data-feminism]: https://data-feminism.mitpress.mit.edu/
:::


## Structure Matters

The shape and organization of a dataset, together called **structure**, govern
how you can access and query its data. Depending on the structure, some
analyses are easier to carry out than others. Most datasets can be stored in
and transformed between a variety of structures, each with different advantages
and disadvantages.

When you start working with a dataset, identify the structure and think about
whether it's suitable for the analyses you want to carry out. If it isn't,
transform to a more suitable structure. The focus of this reader is
exploration, so we won't cover how to transform datasets.

:::{.callout-note title="See Also"}
If you want to learn more about how to transform a dataset with R, see
DataLab's [Intermediate R: Cleaning and Reshaping Data][int-r-clean] workshop
reader.

[int-r-clean]: https://ucdavisdatalab.github.io/dwr_workshop_intermediate_r/chapters/cleaning-reshaping/01_dates-times.html
:::

A **tabular** dataset is one that's structured as a table, with rows and
columns. This reader focuses on tabular datasets, since they're common in
practice and present the fewest programming challenges. Here's an example of a
tabular dataset:

| Fruit  | Quantity | Price
| :----  | -------: | ----:
| apple  | 32       | 1.49
| banana | 541      | 0.79
| pear   | 10       | 1.99

: {.striped .hover}

<!-- TODO: pictures! or links to data? e.g., night heron range -->
<!-- Same dataset, different structures? -->

Some other examples of datasets and structures are:

* People in a family, such as the Roosevelts, might be organized as a (family)
  tree, with connections from parents to children.
* Monthly population estimates for a fish species in the Sacramento River might
  be a series of numbers.
* Relationships between cats in a small town might be a graph (or network),
  with a node for each cat and an edge (connection) between each pair of cats
  that have met.
* Bike paths in Davis, CA might be a collection of geospatial coordinates that
  describe points, lines, and shapes.
* Individual consumer reviews for several different vacuums might be a table
  with a row for each review and columns to identify the individual, vacuum
  model, and review score.


### Features and Observations

Data scientists think about datasets in terms of two components:

* A **feature** (also called a **covariate**) is measurement of something,
  usually across multiple subjects. For example, we might decide to measure the
  heights of everyone in the class. Each person in the class is a subject, and
  the height measurement is a feature. Features don't have to be quantitative.
  If we also asked each person their favorite color, then favorite color would
  be another feature in our dataset. In tabular datasets, features are usually,
  but not always, the columns.

* An **observation** is a set of features measured for a single subject or at a
  single time. So in the preceding example, the combined height and favorite
  color measurement for one student is one observation. In tabular datasets,
  observations are usually, but not always, the rows.

Every dataset has features and observations, regardless of whether the dataset
is tabular.

A **data dictionary** is documentation that describes each feature in a dataset
in detail. For a tabular dataset, a data dictionary typically describes the
unit of observation (in other words, what a subject is) and what each column
measures. The description explain the range of realistic values, regardless of
whether the feature is quantitative or qualitative. For quantitative features,
it should also specify the units of measurement.

We'll see an example of a data dictionary in @sec-ca-least-terns.


## Dataset: CA Least Terns {#sec-ca-least-terns}

The California least tern is a endangered subspecies of seabird that nests
along the coast of California and Mexico. The California Department of Fish and
Wildlife (CDFW) monitors least tern nesting sites across the state to estimate
breeding pairs, fledglings, and predator activity in each annual breeding
season.

![
  A California least tern. Original photo by [Mark Pavelka, U.S. Fish &
  Wildlife Service][pavelka] ([CC BY 2.0][]).
](/images/exploring-data/ca_least_tern_usfws_pacificsw.jpg){
  height=24em
  fig-alt="A gray bird with a white belly, black head, and orange beak sitting
  on a clutch of eggs."
}

[pavelka]: https://www.flickr.com/photos/usfws_pacificsw/5704890596/
[CC BY 2.0]: https://creativecommons.org/licenses/by/2.0/

The CDFW publishes most of the data it collects to the [California Open Data
portal][data.ca.gov]. The examples in this and subsequent chapters use a
cleaned 2000-2023 version of the California least tern data.

[data.ca.gov]: https://data.ca.gov/

:::{.callout-important}
[Click here][ca-least-tern] to view the 2000-2023 California least tern
dataset.

[ca-least-tern]: https://ucdavis.box.com/s/m2w5l2ebp2rey2do5lnn38y1e3u31pin
:::

:::{.callout-note
  title="Documentation for 2000-2023 California Least Tern Dataset"
  collapse=false
}

Each row in the dataset contains measurements from one year-site combination.

| Column                | Description
| --------------------: | :----------
| `year`                | Year of the breeding season
| `site_name`           | Site name
| `site_name_2013_2018` | Site name from 2013-2018
| `site_name_1988_2001` | Site name from 1988-2001
| `site_abbr`           | Abbreviated site name
| `region_3`            | Region of state: S.F. Bay, Central, or Southern (includes Ventura)
| `region_4`            | Region of state: S.F. Bay, Central, Ventura, or Southern
| `event`               | Climate events
| `bp_min`              | Reported minimum breeding pairs
| `bp_max`              | Reported maximum breeding pairs
| `fl_min`              | Reported minimum fledges
| `fl_max`              | Reported maximum fledges
| `total_nests`         | Total reported nests (maximum if a range was reported)
| `nonpred_eggs`        | Total non-predator-related mortalities of eggs
| `nonpred_chicks`      | Total non-predator-related mortalities of chicks
| `nonpred_fl`          | Total non-predator-related mortalities of fledges
| `nonpred_ad`          | Total non-predator-related mortalities of adults
| `pred_control`        | Site predator control (yes/no)
| `pred_eggs`           | Total predator-related mortalities of eggs
| `pred_chicks`         | Total predator-related mortalities of chicks
| `pred_fl`             | Total predator-related mortalities of fledges
| `pred_ad`             | Total predator-related mortalities of adults
| `pred_pefa`           | Predation by peregrine falcons (yes/no)
| `pred_coy_fox`        | Predation by coyotes or foxes (yes/no)
| `pred_meso`           | Predation by other mesocarnivores: dogs, cats, skunks, opossums, raccoons, weasels, etc. (yes/no)
| `pred_owlspp`         | Predation by owls (yes/no)
| `pred_corvid`         | Predation by corvids: ravens or crows (yes/no)
| `pred_other_raptor`   | Predation by raptors other than peregrine falcons and owls (yes/no)
| `pred_other_avian`    | Predation by birds other than raptors and corvids (yes/no)
| `pred_misc`           | Predation by other animals (yes/no)
| `total_pefa`          | Total mortalities due to peregrine falcons
| `total_coy_fox`       | Total mortalities due to coyotes and foxes
| `total_meso`          | Total mortalities due to other mesocarnivores
| `total_owlspp`        | Total mortalities due to owls
| `total_corvid`        | Total mortalities due to ravens and crows
| `total_other_raptor`  | Total mortalities due to other raptors
| `total_other_avian`   | Total mortalities due to other birds
| `total_misc`          | Total mortalities due to other animals
| `first_observed`      | Date CA least terns first observed at site
| `last_observed`       | Date CA least terns last observed at site
| `first_nest`          | Date first egg observed at site
| `first_chick`         | Date first chick observed at site
| `first_fledge`        | Date first fledge observed at site

: {.striped .hover tbl-colwidths="[25, 75]"}

The messy source dataset (with more years and more columns) is available
[here][ca-least-tern-source].

[ca-least-tern-source]: https://data.ca.gov/dataset/california-least-tern-monitoring-sites-generalized-cdfw-ds3146
:::


## Structural Summaries

When you start working with a dataset, examining its structure is one of the
first things you should do. This way you can identify and correct potential
problems, and prepare the dataset for the analyses you'd like to carry out.

Whenever you load a dataset into R, your next step should be to investigate the
data's structure. This step is important because it can help you identify
whether:

* The data was loaded correctly
* There are structural problems with the data that will make it difficult to
  use if they aren't fixed

R provides several functions for getting structural summaries of data. Some of
these are:

* `str` to get a detailed structural summary
* `head`, `tail` to preview the data
* `nrow`, `ncol`, `dim`, `length` to get dimension information
* `names`, `colnames`, `rownames` to get element names
* `class`, `typeof` to get classes and types

Let's try a few of these on the California Least Terns dataset
(@sec-ca-least-terns). For instance, the `str` function shows the classes of
the columns:

```{r}
str(terns)
```

:::{.callout-note title="See Also"}
For a reminder of R's syntax and commands, see DataLab's [R Basics][r-basics]
workshop reader.

[r-basics]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/
:::


## Types of Data {#sec-types-of-data}

Different types of data require different approaches and analysis methods. For
example, if your data are numbers, you can summarize them by computing a sum or
an average. On the other hand, if your data are categories, like genres of
music (rock, pop, classical, ...), you can't use these summarization methods
and have to do something else.

Data scientists categorize data into **types** based on sets of shared
characteristics. This makes it easier to reason about how to transform,
analyze, and present the data. Some widely-understood data types are:

* **Categorical**
    + **Nominal** - data separated into specific categories, with no order. For
      example, hair color (red, brown, blonde, ...) is categorical.
    + **Ordinal** - data separated into specific categories, with an order. For
      example, school level (elementary, middle, high, college) is ordinal.
* **Numerical**
    + **Discrete** - integers, or a finite set of decimal numbers with no
      values in between. Sometimes discrete values can also be treated as
      ordinal. For example, month as a number (1, 2, ..., 12) is discrete.
    + **Continuous** - decimal numbers. There are no specific categories, but
      there is an order. For example, height in inches is numerical.

<!--
Of course, other types of data, like graphs (networks) and natural language
(books, speech, and so on), are also possible.
-->
Categorizing data this way is useful for reasoning about which methods to apply
to which data.

:::{.callout-tip}
Because a feature measures something, we often think and talk about types of
features rather than types of individual values. For example, for a feature
that measures temperature in degrees Celsius, all of the data values will be
numeric (and probably continuous). So we can say the feature is numeric.

In a tidy dataset, each feature has its own column, so it also makes sense to
talk about the types of the columns.
:::


## Correcting Data Types

Often when you load a new dataset, some of the columns won't have the correct
data type (or class) for what you want to do. For instance, in the least terns
dataset, the `site_name`, `region_3`, and `event` columns all contain
categorical data, so they should be factors.

You can convert these columns to factors with R's `factor` function:

```{r}
terns$site_name = factor(terns$site_name)
terns$region_3 = factor(terns$region_3)
terns$event = factor(terns$event)
```

:::{.callout-note title="See Also"}
For more about factors, see the [Factors][r-basics-factors] section of
DataLab's R Basics workshop reader.

[r-basics-factors]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/data-types-structures/data-types-structures.html#sec-factors
:::

:::{.callout-tip}
There's another way we could've done this that uses only two lines of code, no
matter how many columns there are:

```{r, eval = FALSE}
cols = c("site_name", "region_3", "event")
terns[cols] = lapply(terns[cols], factor)
```

You can learn more about the `lapply` function in the [The purrr
Package][r-basics-purrr] section of DataLab's R Basics workshop reader.

[r-basics-purrr]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/organizing-code/control-flow.html#sec-the-purrr-package

You can use whichever approach is more convenient and makes more sense to you.
If there were other columns to convert, we'd go through the same steps with the
appropriate conversion function.
:::

R provides `as.` functions to convert to the most common data types. For
instance, `as.character` converts an object to a string:

```{r}
x = 3.1
class(x)

y = as.character(x)
y
class(y)
```

The `read.csv` function does a good job at identifying columns of numbers, so
it's rarely necessary to convert columns of numbers manually. However, you may
have to do this for data you got some other way (rather than loading a file).
For instance, it's common to make these conversions when scraping data from the
web.

It's also a good idea to convert categorical columns into factors with the
`factor` function and to convert columns of dates into dates. For more about
the latter, see DataLab's [Intermediate R: Cleaning & Reshaping Data][int-r]
workshop reader.

[int-r]: https://ucdavisdatalab.github.io/dwr_workshop_intermediate_r/chapters/cleaning-reshaping/01_dates-times.html


<!--
## References {.unnumbered}

:::{#refs}
:::
-->
