```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Missing Values {#sec-missing-values}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Locate and count missing values in a data set
:::

This chapter covers


## Missing Values

If your data contains missing values, it's important to think about why the
values are missing. Statisticians use two different terms to describe why data
is missing:

* missing at random (MAR)
* missing not at random (MNAR) - causes bias!

When values are **missing at random**, the cause for missingness is one or more
features in the data set. For example, if a soil moisture sensor overheats and
doesn't work on hot days, but air temperature is recorded in the data set,
values are missing at random.

When values are **missing not at random**, the cause for missingness is one or
more features *not* in the data set. Think of this as a form of censorship. For
example, if people in a food survey refuse to report how much sugar they ate on
days where they ate junk food, values are missing not at random. Values MNAR
can bias an analysis.

:::{.callout-note}
Technically, there's a third kind of missing data: **missing completely at
random** (MCAR). When values are missing completely at random, the cause for
missingness is completely unrelated to the research question and features of
interest. This kind of missing data is rare in practice and impossible to
identify from the data alone. Values MCAR can be ignored without causing any
bias.
:::

The default way to handle missing values in R is to ignore them. This is just a
default, not necessarily the best or even an appropriate way to deal with them.
You can remove missing values from a data set by indexing:

```{r}
nonpred_eggs_no_na = terns[!is.na(terns$nonpred_eggs), ]

head(nonpred_eggs_no_na)
```

The `na.omit` function is less precise than indexing, because it removes rows
that have a missing value in _any_ column. This means lots of information gets
lost.

Another way to handle missing values is to **impute**, or fill in, the values
with estimates based on other data in the data set. We won't get into the
details of how to impute missing values here, since it is a fairly deep
subject. Generally it is safe to impute MAR values, but not MNAR values.



