```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Missing Values {#sec-missing-values}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Locate and count missing values in a dataset
:::



When a dataset contains missing values, investigate how many there are and in
which observations and features they occur. Then think carefully about
potential causes and whether there are any patterns to the missingness.

Missing values introduce bias into an analysis when their cause is related to
the research question(s). As an example, consider a study of incidence of
disease in coyotes over a five year period. The coyotes in the study are
tagged, so that they can be located and examined by a veterinarian annually.
Occasionally, the researchers are unable to locate or catch a coyote, so they
record missing values. If healthy coyotes are more difficult to locate and
catch, ignoring the missing values will cause the researchers to overestimate
incidence of disease.

This chapter explains how to use the methods of prior chapters, particularly
@sec-categorical-features and @sec-relationships-between-features, to summarize
and look for patterns in missing values. It also briefly describes how
statisticians think about and address missing values.


## Searching for Patterns

In order to summarize the missing values in a feature, transform the feature
into categories that indicate the presence or absence of a missing value. Then
we can apply any of the methods from @sec-categorical-features to summarize the
missing values, and we can also use methods from
@sec-relationships-between-features to search for patterns among the missing
values.

In R, we can use the `is.na` function to transform a feature into a categorical
feature with `TRUE` at each missing value and `FALSE` at each complete value.
The `is.na` function can also transform an entire dataset this way, going
feature-by-feature.

As an example, let's summarize the missing values in the `bp_min` feature of
the California least terns dataset. First, we'll construct a new categorical
feature that indicates missing values:

```{r}
bp_min_is_na = is.na(terns$bp_min)
```

@sec-categorical-features recommends using frequencies to summarize categorical
features. Since there are only two categories, we'll use a frequency table
rather than a bar plot:

```{r}
table(bp_min_is_na)
```

So there are only 8 missing values in the `bp_min` feature. The number of
complete values is two orders of magnitude larger. So it's unlikely that the
missing values in this feature will pose any major problem for analysis.

From here, we could proceed to investigate whether the missing values in
`bp_min` are related to any of the other features (or their missing values) in
the dataset. To do so, we could use the methods described in
@sec-relationships-between-features.

:::{.callout-tip}
R's `summary` function computes missing value frequencies for every feature in
a dataset, so it's a convenient way to get started and identify features whose
missing values warrant a deeper investigation.
:::


## Kinds of Missingness

Statisticians use two different terms to describe why data is missing:

* When values are **missing at random** (MAR), the cause for missingness is one
  or more features in the dataset. For example, if a soil moisture sensor
  overheats and doesn't work on hot days, but air temperature is recorded in
  the dataset, values are missing at random.

* When values are **missing not at random** (MNAR), the cause for missingness
  is one or more features *not* in the dataset. Think of this as a form of
  censorship. For example, if people in a food survey refuse to report how much
  sugar they ate on days when they ate junk food, values are missing not at
  random.

In both cases, the missing data may bias analyses. The key difference is that
for values MAR, we can sometimes correct the missingness
(@sec-handling-missing-values). Unfortunately, it's not possible to fully
determine whether values are MAR or MNAR from a dataset alone, because this
depends on the data generation and collection processes.

:::{.callout-note}
Technically, there's a third kind of missing data: **missing completely at
random** (MCAR). When values are missing completely at random, the cause for
missingness is completely unrelated to the research question and features of
interest. This kind of missing data is rare in practice and impossible to
identify from the data alone. Values MCAR can be ignored without causing any
bias.
:::


## Handling Missing Values {#sec-handling-missing-values}

The default way to handle missing values in R is to ignore them. This is just a
default, not necessarily the best or even an appropriate way to deal with them.
We can remove missing values from a dataset by indexing:

```{r}
nonpred_eggs_no_na = terns[!is.na(terns$nonpred_eggs), ]

head(nonpred_eggs_no_na)
```

The `na.omit` function is less precise than indexing, because it removes rows
that have a missing value in _any_ column. This means lots of information gets
lost.

Another way to handle missing values is to **impute**, or fill in, the values
with estimates based on other data in the dataset. We won't get into the
details of how to impute missing values here, since it's a fairly deep subject.
Generally it is safe to impute MAR values, but not MNAR values.

:::{.callout-note title="See Also"}
The [mice][] package for R provides a variety of functions for imputing values,
as well as references to statistical literature about imputation.

[mice]: https://amices.org/mice/
:::

