```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Outliers {#sec-outliers}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what it means for a value to be an "outlier"
* Locate and count outliers in a dataset
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

An **outlier** is an observation that's substantively different from most of
the others. Outliers are unusual, and unusual observations deserve extra
scrutiny because they might be:

* Unusually informative;
* An error caused by how data were measured, recorded, or cleaned.

The definition of an outlier is broad, open to interpretation, and should be
tailored to specific datasets and analysis goals, as is the case with many
statistical ideas. This chapter explains a few ways to find outliers (among
many), how to think through whether outliers are natural or erroneous, and what
to do with erroneous outliers.


## Searching Low-Density Regions

Picture the density curve of a distribution, like the one shown in
@fig-total_nests_density. Most observations occur where the density is high;
that's *why* it's high. So outliers, observations that are unlike most of the
others, must lie in regions of relatively low density.

As such, visualizing the distribution of a dataset (or feature) is an effective
way to identify outliers. We can use familiar visualizations like histograms
(@sec-binning-histograms) and density plots (@sec-density-plots). To help us
see where individual observations are, we can mark them on the axis. The marks
are called a **rug**.

Let's make a density plot of `total_nests` from the California least terns
dataset one more time, now with a rug:

```{r}
library("ggplot2")

ggplot(terns) +
  aes(x = total_nests) +
  geom_density() +
  geom_rug()
```

Most of the observations are less than about 250 nests, and we can distinguish
individual observations beyond about 750 nests. There's no sharp dividing line
where observations clearly become outliers, so how do we decide? One way is to
use external information about the nesting habits of California least terns,
such as research literature or personal expertise.

But remember why we're looking for outliers: they simply get extra scrutiny. As
long as they seem to be valid (not errors), we'll treat them like any other
observations and use them in our analyses. How we decide which observations are
outliers shouldn't affect the results much as long as we examine enough to
catch errors.

:::{.callout-note title="Note: Jitter" collapse="true"}
Rugs are difficult to read where there are lots of observations, because many
marks get plotted on top of each another.

An alternative approach is to plot the observations as points. We can make the
points slightly transparent to make the overlapping points easier to
distinguish. We can also add small, random offset, called **jitter**, along the
y-axis. Then the plot becomes:

```{r}
ggplot(terns) +
  aes(x = total_nests) +
  geom_density() +
  geom_jitter(aes(y = -0.0005), height = 0.0005, alpha = 0.5) +
  labs(y = "density")
```

This takes more effort than a rug, since we have to poisition the points along
the y-axis, but it's also easier to read. See the [documentation for
`geom_jitter`][jitter] to learn more about how to use jitter.

[jitter]: https://ggplot2.tidyverse.org/reference/geom_jitter.html
:::


## Box Plots {#sec-box-plots}

Observations are unlikely to be very far away from the center of their
distribution, so we can use distance from the center to identify outliers. A
**box plot** makes this approach visual. A box plot is yet another kind of
distribution plot, like a histogram or density plot.

In a box plot, the central 50% of the data are represented by a box. The left
edge of the box is at the 25th percentile, and the right edge of the box is at
the 75th percentile. The box is split by a heavy line at the median.

The box also has **whiskers**, lines that represent the tails of the data. The
whiskers extend from each end of the box to the farthest observation within
$1.5 \times \textrm{IQR}$. Observations farther away are plotted as points and
often treated as outliers.

:::{.callout-note title="Note: Whiskers"}
There are actually many different conventions for how to draw the whiskers in a
box plot. The $1.5 \times \textrm{IQR}$ convention is widely-used and is the
default in R.

To learn more about the other conventions for box plots, see the [ggplot2
documentation][ggplot-box] or the [Wikipedia article][wiki-box].

[ggplot-box]: https://ggplot2.tidyverse.org/reference/geom_boxplot.html
[wiki-box]: https://en.wikipedia.org/wiki/Box_plot
:::

To demonstrate box plots, suppose we want to know whether there are any
unusually dangerous sites where California least terns nest, where the danger
is caused by something other than natural predators. There are several
`nonpred_` features in the dataset we can use to investigate. Let's focus on
`nonpred_eggs`, which the documentation describes as non-predator-related
mortalities of eggs. Here's a box plot of the feature:

```{r}
#| label: fig-nonpred_eggs_box
ggplot(terns) +
  aes(x = nonpred_eggs) +
  geom_boxplot()
```

Most observations are between 0 and about 75 mortalities. The left whisker is
barely visible, because `nonpred_eggs` is never negative. The right whisker
extends to about 175 mortalities. We might want to treat the observations
beyond the whisker as outliers, and the plot shows there are a lot of them,
with the largest around 750 mortalities.

Some sites have more eggs than others, so a high count of non-predator egg
mortalities might not mean a site is unusually dangerous. We can find dangerous
sites by looking at non-predator egg mortalities per nest:

```{r}
#| label: fig-nonpred_eggs_total_nests_box
ggplot(terns) +
  aes(x = nonpred_eggs / total_nests) +
  geom_boxplot()
```

There are now just a few observations that stand out as outliers compared to
the others. We could continue this analaysis by examining each of these
observations individually, to check whick sites they're from and whether they
have any other unusual features.

:::{.callout-tip}
For most datasets, we recommend density plots over box plots, because density
plots show more detail about the shape of the entire distribution. That said,
when the number of observations is small, we prefer box plots. Less detail
becomes a safeguard against overinterpretation when the evidence for a detail
may be just 1 or 2 values.
:::

:::{.callout-warning title="Warning: Outliers at the Center" collapse="true"}
Observations far from the center of a distribution are often outliers, but
outliers can also occur close to the center! Remember that an outlier is just
an unusual or unlikely observation.

For example, consider this multimodal distribution with mean 0:

```{r}
#| label: fig-multimodal-density
#| echo: false
f = function(x) 0.5 * dnorm(x, -10) + 0.5 * dnorm(x, 10)
ggplot() + xlim(-15, 15) + geom_function(fun = f, n = 501) + ylab("density")
```

In this case, we might say an observation at 0 is an outlier, since the density
very low there. Also notice how a sample from the distribution looks as a box
plot:

```{r}
#| label: fig-multimodal-box
#| echo: false
bimodal = data.frame(x = c(rnorm(30, -10), rnorm(30, 10)))
ggplot(bimodal) + aes(x = x) + geom_boxplot()
```

The short whiskers are the only indication that almost all of the observations
are close to -10 and 10 rather than 0. This shows how box plots can mislead
more easily than density plots.
:::


## Standardization {#sec-standardization}

The mean (@sec-mean) estimates the center of a distribution, so we can look for
outliers among observations that are far away from the mean. A convenient way
to do this is to **standardize** each observation by subtracting the mean and
dividing by the standard deviation. The resulting statistic is called the
**z-score**.

:::{.callout-note title="Definition: z-score"}
Mathematically, for an observation $x$ from a distribution with mean $\mu$ and
standard deviation $\sigma$, the z-score $z$ is defined as

$$
z
= \frac{x - \mu}{\sigma}
\approx \frac{x - \bar{x}}{s}
$$

When the mean and standard deviation are unknown, but we have a sample, we can
estimate with the sample mean $\bar{x}$ and sample standard deviation $s$.
:::

The z-score is convenient because it measures distance in multiples of the
standard deviation rather than the original units. This means we can compare
z-scores across different features with different units. Moreover, for almost
any distribution, at least 88% of all observations must be within 3 standard
deviations of the mean. So to search for outliers, we can simply search for
observations with a z-score greater than or equal to 3.

:::{.callout-note title="Note: Chebyshev's Inequality"}
The at least 88% within 3 standard deviations rule is a specific case of a
statistical theorem called [Chebyshev's Inequality][chebyshev].

[chebyshev]: https://en.wikipedia.org/wiki/Chebyshev%27s_inequality

For some distributions, we can make [even stronger guarantees][68-95].

[68-95]: https://en.wikipedia.org/wiki/68%E2%80%9395%E2%80%9399.7_rule
:::

In R, the `scale` function computes z-scores. Let's compute the z-scores for
the `nonpred_eggs` feature, then find the row numbers and values of the
observations where `nonpred_eggs` is more than 3 standard deviations from the
mean:

```{r}
z = scale(terns$nonpred_eggs)
head(z)

rows = which(z <= -3 | 3 <= z)
rows

terns$nonpred_eggs[rows]
```

All of these observations are more than 3 standard deviations greater than the
mean. For the most part, this is consistent with what we found by using a box
plot in @sec-box-plots. So we might want to treat these observations as
outliers and do a more thorough investigation.


## Investigating Outliers {#sec-investigating-outliers}

After locating outliers in a dataset, the next step is to investigate each one
indvidually. The investigation has two goals:

First, we want to determine whether the outlier is valid or erroneous. Start by
checking all of the observation's features. Some questions to ask are:

* How many features are outliers?
* Are any of them clearly erroneous, nonsensical, or impossible?
* Are any of the features derived from others? If so, are they consistent with
  the feature(s) from which they're derived?
* Does the documentation say anything about this particular kind of outlier?

External sources, such as related research literature, domain expertise, and
direct communication with the people who created the dataset, can also provide
strong evidence as to whether an observation is valid or not.

Second, we want to determine whether the outlier provides any special insight
relevant to the purpose of the overarching analysis. How to check this depends
on what that purpose is, but the exploration techniques described in this
reader are often suitable.

When an outlier is valid, keep it.

When an outlier is erroneous, sometimes it's possible to correct it. Check
whether any of the observation's other features suggest the correct value.

If the error can't be corrected with the other features, sometimes it's
possible to approximate the correct value by using the mean or median (or some
other suitable statistic) of a sample of similar observations. This is called
**imputing** the value. Whether it's statistically appropriate to impute a
value depends on what the observation and feature represent.

If neither correction nor imputation are possible, replace each of the
observation's erroneous features with a missing value `NA`. This way the
observation's valid features are preserved. If the entire observation is
erroneous, remove it from the dataset.

:::{.callout-important}
Above all else, make sure to document the process you use to locate and handle
the outliers in a dataset. The idea is to make the analysis as transparent as
possible, so that no one is left wondering whether your conclusions are valid.
:::

:::{.callout-tip}
If an outlier is valid but interferes with a plot you want to make, adjust the
x and y limits of the plot rather than removing the outlier from the dataset.
If the outlier is no longer visible, make sure to mention it in the plot's
caption or description.
:::

