```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Outliers

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what it means for a value to be an "outlier"
* Locate and count outliers in a data set
:::

This chapter covers


## Outliers

An **outlier** is an anomalous or extreme value in a data set. We can picture
this as a value that's far away from most of the other values. Sometimes
outliers are a natural part of the data set. In other situations, outliers can
indicate errors in how the data were measured, recorded, or cleaned.

There's no specific definition for "extreme" or "far away". A good starting
point for detecting outliers is to make a plot that shows how the values are
distributed. Box plots and density plots work especially well for this (you'll
learn about how to make plots in a later lesson):

```{r}
library("ggplot2")

ggplot(terns, aes(x = nonpred_eggs)) + geom_boxplot()

# Some sites might have more eggs than others, so a high number of non-predator
# egg mortalities does not necessarily mean a site is unusually dangerous. We
# can find dangerous sites by looking at non-predator egg mortalities per nest:

ggplot(terns, aes(x = nonpred_eggs / total_nests)) + geom_boxplot()
```

Statisticians tend to use the rule of thumb that any value more than 3 standard
deviations away from the mean is an outlier. You can use the `scale` function
to compute how many standard deviations the elements in a column are from their
mean:

```{r}
z = scale(terns$nonpred_eggs)
head(z)

which(z <= -3 | 3 <= z)
```

Be careful to think about what your specific data set measures, as this
definition isn't appropriate in every situation.

How can you handle outliers? First, try inspecting other features from the row
to determine whether the outlier is a valid measurement or an error. When an
outlier is valid, keep it.

If the outlier interferes with a plot you want to make, you can adjust the x
and y limits on plots as needed to "ignore" the outlier. Make sure to mention
this in the plot's title or caption.

When an outlier is not valid, first try to correct it. For example:

* Correct with a different covariate from the _same observation_.
* Estimate with a mean or median of similar observations. This is another
  example of imputing values.

<!--
For example, in the Craigslist data, we can use the `text` column to try to
correct outliers:

```r
message(cl$text[1261])
```

Based on the text, this apartment is 819 square feet, not 8190 square feet. So
we can reassign the value:
```r
cl$sqft[1261] = 819
```
-->

If other features don't help with correction, try getting information from
external sources. If you can't correct the outlier but know it's invalid,
replace it with a missing value `NA`.
