```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Relationships Between Features {#sec-relationships-between-features}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* TODO
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

Until now, we've focused almost exclusively on exploring one feature at a time.
In order to understand relationships between features, we need to explore two
or more together. This chapter introduces a few strategies for visualizing and
measuring relationships between features.


## Visualizing Relationships

Visualizations are an excellent starting point for exploring multiple features
at once, because they show more details than any individual statistic.
Moreover, in R (especially with ggplot2), changing a plot of one feature to
show more is relatively straightforward.

When thinking about visualizing multiple features, the first thing to do is
choose an appropriate plot. The types of the features generally hint at
specific kinds of plots:

| Feature 1   | Feature 2   | Plot
| :---------- |:----------- |:----
| categorical |             | bar, dot
| categorical | categorical | bar, dot, mosaic
| numerical   |             | box, density, histogram
| numerical   | categorical | box, density, ridge
| numerical   | numerical   | line, scatter, smooth scatter

: Ways to plot 1-2 features. {#tbl-choose-plot .striped .hover tbl-cap-location="bottom"
}


If you want to add a:

* 3rd numerical feature, use it to change point/line sizes.
* 3rd categorical feature, use it to change point/line styles.
* 4th categorical feature, use side-by-side plots. Often these are called
  **faceted** plots.

Keep in mind these are all suggestions rather than rules. Which plot is best
also depends on the questions you want to answer, the distribution of the data,
and the audience for the plot.

:::{.callout-note title="See Also"}
Visualization design is a deep topic, and whole books have been written about
it. One resource where you can learn more is DataLab's [Principles of Data
Visualization][dataviz] workshop reader.

[dataviz]: https://ucdavisdatalab.github.io/workshop_data_viz_principles/

R, the ggplot2 package, and the [ggplot2 extensions][ggplot2-ext] can create a
wide variety of visualizations. To learn more about what's possible, see
DataLab's [Intermediate R: Next-Level Data Visualization][int-r-dataviz] and
the [ggplot2 documentation][ggplot2].

[ggplot2-ext]: https://exts.ggplot2.tidyverse.org/gallery/
[int-r-dataviz]: https://ucdavisdatalab.github.io/dwr_workshop_intermediate_r/chapters/visualizing-analyzing/01_visualization.html
[ggplot2]: ttps://ggplot2.tidyverse.org/
:::


### Categorical vs. Categorical

As a demonstration, let's plot some features from the California least terns
dataset. In @sec-categorical-features, we found that `region_3` includes three
unexpected extra regions. We also made a frequency bar plot for `year`
(@fig-year_bar) and found that there's a sharp decline in observations in 2018.
Now let's try to answer a two new questions:

1. Are the regions evenly sampled across years?
2. Is the 2018 decline associated with specific regions?

The `region_3` feature is categorical. The `year` feature is discrete, but we
can treat it as categorical. So @tbl-choose-plot suggests a bar, dot, or mosaic
plot. Let's make a bar plot. We can use fill color to show the regions:

```{r}
#| label: fig-year_region_3_bar
library("ggplot2")

ggplot(terns) +
  aes(x = year, fill = region_3) +
  geom_bar()
```

@fig-year_region_3_bar shows that the three expected regions S.F. Bay, Central,
and Southern are present across all of the years with observations. There's
some variation in the numbers of observations across the years. The decrease in
observations in 2018 is mostly caused by a decrease in observations for the
Southern and Central regions, although there's also a slight decrease for S.F.
Bay.

Among the three unexpected regions, Arizona is only present in 2009. Kings is
present in 2000 and 2004-2017. Sacramento is present in 2008-2017. For all
three of these regions, the numbers of observations appear to be constant and
are negligible compared to the expected regions.


### Numerical vs. Categorical

Next, let's explore whether the regions have similar numbers of nests. We'll
use `region_3` again, and we'll also use the numerical feature `total_nests`.

Before we make a plot, let's filter the dataset so that it only includes the
observations for the expected regions. The other regions have very low numbers
of observations, so it's unlikely they'll support strong conclusions. We'll
call the filtered dataset `terns_ok`:

```{r}
library("dplyr")
ok_regions = c("S.F._BAY", "CENTRAL", "SOUTHERN")
terns_ok = filter(terns, region_3 %in% ok_regions)
```

According to @tbl-choose-plot, a box, density, or ridge plot may be appropriate
for this pair of features. Let's use a density plot. We can put the numerical
feature `total_nests` on the x-axis and color-code the density curves by
`region_3`:

```{r}
#| label: fig-total_nests_region_3_density
ggplot(terns_ok) +
  aes(x = total_nests, color = region_3) +
  geom_density()
```

The distribution of `total_nests` appears to be very different from region to
region in @fig-total_nests_region_3_density. In the Central region, the number
of nests is almost always relatively low, with most observations below about
125 nests. In the S.F. Bay region, there are still many observations with low
numbers of nests, but there's also a big group of observations with 250-500
nests. Curiously, there are few observations with 125-250 nests. In the
Southern region, the distribution is much flatter than the others, meaning a
greater proportion of the observations have relatively large numbers of nests.

Now consider another feature, `event`, which indicates the climate condition
(El Niño, La Niña, or neutral) in the year of each observation. Suppose we want
to know whether numbers of nests vary with the climate condition, and whether
this differs by region. Then we need to make a plot with `region_3`,
`total_nests`, and `event`. We could try to include `event` in the density
plot (@fig-total_nests_region_3_density), but putting more curves on the plot
might make it difficult to read.

Instead, let's use a box plot. We can put `total_nests` on the x-axis,
`region_3` on the y-axis, and color-code the boxes by `event`:

```{r}
#| label: fig-total_nests_event_box
ggplot(terns_ok) +
  aes(x = total_nests, y = region_3, fill = event) +
  geom_boxplot()
```

With @fig-total_nests_event_box, we can compare the distribution of
`total_nests` under different climate conditions for each region. Within each
region, the differences between climate conditions are small. For Southern, the
La Niña observations appear to have slightly fewer nests. This pattern is
reversed for S.F. Bay and Central, where the La Niña observations have slightly
higher median numbers of nests.


### Numerical vs. Numerical {#sec-numerical-numerical}

Finally, let's examine the relationships between some of the numerical
features. We'll start by asking whether more nests means more mortalities from
non-predator causes. So we'll compare `total_nests` with `nonpred_eggs`. Based
on @tbl-choose-plot, we can use a line, scatter, or smooth scatter plot for
this. Let's use a scatter plot:

```{r}
#| label: fig-total_nests_nonpred_eggs_scatter
ggplot(terns) +
  aes(x = total_nests, y = nonpred_eggs) +
  geom_point()
```

@fig-total_nests_nonpred_eggs_scatter suggests that `total_nests` and
`nonpred_eggs` do generally increase together, but there's a lot of individual
variation, and exact nature of the relationship is unclear.

Thinking about `nonpred_eggs`, we might wonder whether non-predator and
predator egg mortality trend together. In other words, does `nonpred_eggs`
increase with `pred_eggs`? Again we can use a scatter plot:

```{r}
#| label: fig-pred_eggs_nonpred_eggs_scatter
ggplot(terns) +
  aes(x = pred_eggs, y = nonpred_eggs) +
  geom_point()
```

In @fig-pred_eggs_nonpred_eggs_scatter, it's apparent that the relationship
between `pred_eggs` and `nonpred_eggs` is weak. This suggests that, perhaps
unsurprisingly, the causes of non-predator mortality are probably unrelated to
the causes of predation.

It seems plausible that that observations with more breeding pairs will have
more nests, so let's investigate whether this is true as our final exploration.
We can use the `total_nests` and `bp_min` features for this. Once again, we'll
make a scatter plot:

```{r}
#| label: fig-total_nests_bp_min_scatter
ggplot(terns) +
  aes(x = total_nests, y = bp_min) +
  geom_point()
```

@fig-total_nests_bp_min_scatter shows a strong, nearly linear relationship
between the two features. Breeding pairs and nests do tend to increase
together.


## Correlation

**Correlation** is a statistic that measures the direction and strength of the
linear relationship between two numerical features. In other words, correlation
measures whether the two features trace out a line when plotted, and whether
the slope of the line is positive or negative. Think of correlation as
answering the question "How much does one feature tell us about the other
(and vice-versa)?"

:::{.callout-note title="Definition: Sample Correlation"}
Mathematically, for two features $x_1, x_2, \dots, x_n$ and $y_1, y_2, \dots,
y_n$, the sample correlation $r$ is defined as

$$
r_{x,\,y}
= \frac{1}{n-1} \sum_{i=1}^n \frac{(x_i - \bar{x})(y_i - \bar{y})} {s_x s_y}
$$

Notice that this is similar to the definition of sample variance and sample
standard deviation (@sec-standard-deviation). In fact, the sample standard
deviation of a feature is the sample correlation of that feature with itself.

Statisticians conventionally write the unknown population correlation, which is
estimated by the sample correlation, as $\rho$. It's often written with a
subscript to indicate the two associated features.
:::

Correlation is always between -1 and 1. Negative correlation means that as one
feature increases, the other decreases. Positive correlation means they
increase together. Correlation close to 0 means the features are unrelated or
have a non-linear relationship. Correlation close to -1 or 1 means the features
have a strong linear relationship.

In R, the `cor` function computes correlation. As examples, let's compute
correlations for the pairs of features we plotted in @sec-numerical-numerical.
Some of the features contain missing values, so we'll set the argument `use =
"complete.obs"` in the `cor` function; this makes it skip the missing values
(similar to `na.rm = TRUE` in other functions).

First up are `total_nests` and `nonpred_eggs`. When we plotted these in
@fig-total_nests_nonpred_eggs_scatter, there was a clearly a relationship, but
it wasn't an exact line. The correlation is:

```{r}
cor(terns$total_nests, terns$nonpred_eggs, use = "complete.obs")
```

The correlation is positive and fairly high, in spite of the individual
variation in observations. The two features tend to increase together linearly.

Now let's compute the correlation of `pred_eggs` and `nonpred_eggs`. In
@fig-pred_eggs_nonpred_eggs_scatter, these two features didn't appear to be
related. The correlation is:

```{r}
cor(terns$pred_eggs, terns$nonpred_eggs, use = "complete.obs")
```

So there's a very weak tendency to increase together linearly.

Finally, we'll compute the correlation of `total_nests` and `bp_min`. These
features exhibited a strong linear relationship in
@fig-total_nests_bp_min_scatter. The correlation is:

```{r}
cor(terns$total_nests, terns$bp_min, use = "complete.obs")
```

This is consistent with the strong relationship we saw in the plot.

:::{.callout-warning}
Keep in mind that correlation measures the *linear* relationship between
features. When features have a non-linear relationship, correlation will
usually underestimate its strength.

The understimation is often mild as long as the relationship is **monotonic**,
meaning the features always trend in the same direction or always trend in
opposite directions. For instance, consider two features where one is the cube
of the other ($y = x^3$). This is a monotonic relationship, because as one
increases, so does the other:

```{r}
#| echo: false
set.seed(101)
x = runif(100, -3, 3)
y = x^3
df = data.frame(x = x, y = y)
ggplot(df) + aes(x = x, y = y) + geom_point(alpha = 0.5)
```

This relationship is as strong as possible---given one feature, we can compute
the other. Correlation underestimates the strength of the relationship at
`r round(cor(x, y), 3)`.

When the relationship is not monotonic, correlation can be close to 0 even if
the relationship is very strong. For example, consider two features which lie
on the perimeter of a circle:

```{r}
#| echo: false
x = runif(100, -3, 3)
y = sqrt(9 - x^2)
df = data.frame(x = x, y = y)
ggplot(df) + aes(x = x, y = y) + geom_point(alpha = 0.5) + coord_fixed()
```

Once again, this relationship is as strong as possible. Nevertheless, because
the relationship is not monotonic, the correlation is only
`r round(cor(x, y), 3)`.

If you're unsure of how to interpret a correlation, making a visualization is a
great way to get more insight into the relationship between features.
:::
