```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Statistical Summaries {#sec-data-forensics}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the location statistics mean, median, and mode
* Explain the scale statistics range and standard deviation
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

This chapter presents a variety of methods to summarize categorical and
numerical data, including statistical estimates and visualizations. For each of
these methods, we discuss how to interpret the summary and explain the method's
advantages and disadvantages relative to others. The unifying theme of the
chapter is summarizing, making sense of, and communicating how data are
distributed within populations and samples.


## Categorical Data

After investigating the data's structure, it's a good idea to check some basic
statistical properties. This step is important because it can help you identify
limitations of and patterns in the data.

Let's continue to explore the California least terns dataset
(@sec-ca-least-terns). As a reminder, the structure of the dataset is:

```{r}
str(terns)
```

We'll start by examining some of the categorical features, since there are only
a few ways to summarize categorical data. As a motivating question, we'll try
to understand which parts of California are included in the dataset and whether
they're represented equally. To do this at a coarse scale, we can look at the
`region_3` column, which, according to the documentation, has categories for
three regions. The first few values are:

```{r}
head(terns$region_3)
```

The documentation claims the three regions are S.F. Bay, Central, and Southern,
but we can already see that this isn't accurate. There's another region, Kings.
Presumably this corresponds to sites in [Kings County][kings]. Let's check
whether there are any other unexpected regions:

[kings]: https://en.wikipedia.org/wiki/Kings_County,_California

```{r}
unique(terns$region_3)
```

The output reveals there are three undocumented regions: Kings, Sacramento, and
Arizona. Later, in @sec-fixme, we'll try to figure out what the Arizona region
represents, since there aren't any counties or regions of California with this
name.

We can count the number of times each region appears in the `region_3` column
to get a sense of how many observations come from each region. The `table`
function counts things:

```{r}
table(terns$region_3)
```

This reveals a few things about the `region_3` column and the dataset. First,
the Kings, Sacramento, and Arizona regions are uncommon. Each has 15
observations or less and is an order of magnitude smaller than the S.F. Bay,
Central, and Southern regions. Second, at 576 observations, the Southern region
occurs far more than any of the others, with S.F. Bay next, at 112
observations.

The disparity in counts between the three documented regions could be due to
nesting preferences of the birds: perhaps there are more nesting sites in
southern California. On the other hand, it could also be due to how the dataset
is collected: perhaps the regions have different coverage over time, or CDFW
uses geographically smaller sites in southern California. Thinking through the
different ways a dataset might have been generated like this can help you
identify subtle details that change or temper interpretations and lead to
further lines of inquiry. We'll revisit the questions about `region_3` in
@sec-fixme.

Computing **frequencies** or **counts**---the number of times each category
appears---is by far the most common, useful, and complete way to summarize
categorical data.

The `table` function is an essential tool for computing frequencies, but it
does have one shortcoming: even small tables can be easy to misread, especially
if the names of the categories are long or include numbers. Large tables make
the problem worse. Sorting by frequency partially addresses the problem:

```{r}
sort(table(terns$region_3))
```

This makes it easy to see which categories are relatively small or large, but
the categories are no longer in alphabetical order, so it's harder to find
specific categories. For ordinal data, where there's a natural order to the
categories, sorting by frequency can break that order and make tables
confusing.

We can use a bar plot to display frequencies visually instead of in a table.
Here's a bar plot for the `region_3` column:

```{r}
library("ggplot2")

ggplot(terns) +
  aes(x = region_3) +
  geom_bar()
```

The bar plot is convenient for comparing different categories, even when there
are a lot of categories. As with a table, we can sort the categories by
frequency in a bar plot to make the extremes more pronounced, but these are
usually evident even without sorting.

Bar plots have two disadvantages compared to tables. First, the exact count for
a category can be difficult to read. A simple way to address this is to display
the count on the plot:

```{r}
ggplot(terns) +
  aes(x = region_3) +
  geom_bar() +
  geom_text(
    # Use counts for the labels (geom_bar also uses stat_count).
    stat = "count", aes(label = after_stat(count)),
    # Set vertical justification of the labels, so they sit atop the bars.
    vjust = -0.5
  )
```

Second, bar plots are often less legible than tables to people with low vision.
One way to get the best of both bar plots and tables is to simply include both.

The **mode** of a feature is the category that occurs most frequently. In a bar
plot, the mode corresponds to the category with the tallest bar. A feature can
have multiple modes, where several values are tied for most frequent. For the
`region_3` column, the mode is the Southern category.

:::{.callout-warning}
R's `mode` function is unrelated to the statistical mode and returns something
different.

There's no built-in function to calculate modes in R. Use the `table` function
or a bar plot to identify modes instead.
:::

We can think of the mode of a feature as the category that's most likely if we
were to randomly select an observation from the dataset. If the dataset is a
random sample from a larger population, we can also think of the **sample
mode** (computed with the dataset) as an estimate of the population's mode.


### Discrete Data as Categorical

Discrete data are unique, because they can be treated as both categories and
numbers. Let's think about how we can apply summaries for categorical data to
the `year` column in the California least terns dataset. As motivation, we'll
try to understand the timespan of the dataset.

We can compute frequencies of the distinct values for a discrete feature in the
same way as for a categorical feature:

```{r}
table(terns$year)
```

Likewise, we can make a bar plot, with the same advantages and disadvantages as
before:

```{r}
ggplot(terns) +
  aes(x = year) +
  geom_bar()
```

The bar plot makes it apparent that the dataset, which spans years 2000-2023,
doesn't include any observations for the years 2001-2003. It also shows that
there are roughly 30-40 observations for every included year. That might mean
most sites were consistently measured over the entire time period, or it might
be a coincidence due to mostly equal numbers of sites being added and removed
over time. There's a distinct drop in observations in 2018, which could
indicate a major change in which sites were being monitored. The mode is 2017.

When a feature has a large number of possible values, tables and bar plots
can become difficult to interpret. If the feature is ordered (whether ordinal
or numerical), we're sometimes more interested in general trends in frequency
as we move through the values in order, rather than frequencies at specific
values.

We can get a sense of general trends by **binning** the values, which means
treating values close together as a single value. Imagine setting out a small
number of bins and dropping a few consecutive values into each. We then compute
frequencies (or other statistics) for the bins rather than the original values.
Binning tends to smooth out extremes, because it combines multiple values (some
of which will be less extreme).

A **histogram** is a bar plot where the values on the x-axis are binned. We can
use R to bin data and make a histogram automatically, although we recommend
(and the ggplot2 documentation recommends) setting the bin width manually for
numerical data. For instance, here's how to make a histogram for the `year`
column with 3-year bins:

```{r}
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 3, color = "black")
```

In the histogram, notice that the trends from the bar plot are smoothed out.
Key details still remain, like the low frequency for 2000-2003 and the drop in
frequency in 2018. For intuition, think of the original bar plot as having a
bin width of 1; wider bins produce more smoothing.

We'll revisit histograms for continuous data in @sec-numerical-data.

:::{.callout-tip}
When you make histograms, try several different bin widths. Sometimes one bin
width reveals details that are hidden in another. Choose a bin width that
preserves the most relevant details to your questions about the data.

If different bin widths lead to different interpretations of the data, include
both in your reporting and address why they do.
:::

:::{.callout-note}
R's functions for making histograms assume numerical data, but we could make a
histogram for ordinal data in two steps, by first binning and then making a bar
plot.
:::


## Numerical Data {#sec-numerical-data}

What about numerical data?

Two important questions to ask about data are:

1. Where is it? This is the **location** of the data.

2. How spread out is it? This is the **scale** of the data.

Let's use the data

```{r}
x = c(-2, -1, -1, -1, 0, 2, 6)
```

as an example.

Location is generally summarized with a number near the middle or center of
the data. A few options are:

1.  **Mode** - the value that appears most frequently. The mode can be
    calculated for any kind of data, but doesn't work well for continuous data.

    For our example, the mode of `x` is -1. You can compute the mode with
    `table`:

    ```{r}
    table(x)
    ```

2.  **Median** - sort the data, then find the value in the middle. The median
    can be calculated for ordinal or numerical data.

    For our example, the median is -1. Compute this with `median`:

    ```{r}
    median(x)
    ```

3.  **Mean** - the balancing point of the data, if a waiter was trying to
    balance the data on a tray. The mean can only be calculated for numerical
    data.

    For our example the mean is 0.4285. Compute this with `mean`:

    ```{r}
    mean(x)
    ```

Adding large values to the data affects the mean more than the median:

```{r}
y = c(x, 100)
mean(y)
median(y)
```

Because of this, we say that the median is **robust**.

The mean is good for getting a general idea of where the center of the data
is, while comparing it with the median reveals whether there are any
unusually large or small values.

Scale is generally summarized by a number that says how far the data is from
the center (mean, median, etc...). Two options are:

1.  **Standard Deviation** - square root of the average squared distance to the
    mean (the distance from a point to a mean is called a **deviation**). You
    can think of this as approximately the average distance from a data point
    to the mean. As a rule of thumb, most of the data will be within 3 standard
    deviations of the mean.

    You can compute the standard deviation with `sd`:

    ```{r}
    sd(x)
    ```

2.  **Interquartile Range** (IQR) - difference between the 75th and 25th
    percentile. The median is the 50th **percentile** of the data; it's at the
    middle of the sorted data. We can also consider other percentiles. For
    instance, the 25th percentile is the value one-quarter of the way through
    the sorted data.

    **Quantile** is another word for percentile. **Quartile** specifically
    refers to the 25th, 50th, and 75th percentiles because they separate the
    data into four parts (hence "quart-").

    You can compute quantiles with `quantile`, or compute the IQR directly with
    `IQR`:

    ```{r}
    quantile(x)

    # IQR
    IQR(x)
    ```

The IQR is more robust than the standard deviation.

Many of the functions for computing statistical summaries have a parameter
`na.rm` to ignore missing values. Setting `na.rm = TRUE` is often useful when
you're just trying to do an initial investigation of the data. However, in a
more complete analysis, you should think carefully about what the missing
values mean, whether they follow any patterns, and whether there are enough
non-missing values for statistical summaries to be good representatives of the
data.

Finally, the `summary` function computes a detailed statistical summary of an R
object. For data frames, the function computes a summary of each column,
guessing an appropriate statistic based on the column's data type.
