```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Statistical Summaries {#sec-data-forensics}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the location statistics mean, median, and mode
* Explain the scale statistics range and standard deviation
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

After investigating (and possibly transforming) a dataset's structure, the next
step is to examine the statistical properties of its features. This chapter
focuses on individual features, and presents a variety of statistical
visualizations and estimators to summarize them, whether they're categorical or
numerical. These methods will help us identify and communicate trends,
patterns, anomalies, and limitations. For each method, we discuss how to
interpret the summary and explain the method's strengths and weaknesses
relative to others. The unifying theme of the chapter is summarizing, making
sense of, and communicating how data are distributed within populations and
samples.


## Categorical Features {#sec-categorical-features}

To introduce ideas gently, we'll start by focusing on summaries for categorical
features. Most of the time, categories are not numbers, so we won't use much
math, and when we do, it will mostly be counting.

As motivation and a source of examples, we'll work to understand which regions
of California and which time periods are recorded in the California least terns
dataset (@sec-ca-least-terns). We'll also examine whether regions and time
periods are equally represented in the dataset.

To refresh our memory, the structure of the California least terns dataset is:

```{r}
str(terns)
```

We can study regional coverage at a coarse scale by looking at the `region_3`
or `region_4` columns. Let's focus on `region_3`. According to the
documentation, it's categorical, with categories for three regions. The first
few values are:

```{r}
head(terns$region_3)
```

The documentation claims the three regions are S.F. Bay, Central, and Southern,
but we can already see that this isn't accurate. There's another region: Kings.

The Kings region likely corresponds to sites in [Kings County][kings]. We can
try to confirm this by examining the site names in the region (with R's
[dplyr][] package):

[kings]: https://en.wikipedia.org/wiki/Kings_County,_California
[dplyr]: https://dplyr.tidyverse.org/

```{r}
#| warning: false
library("dplyr")

kings = filter(terns, region_3 == "KINGS")
unique(kings$site_name)
```

[Kettleman City][kc] is indeed a place in Kings County, so it seems like our
guess was correct.

[kc]: https://en.wikipedia.org/wiki/Kettleman_City,_California

:::{.callout-tip}
As you explore data, make a habit of following up on patterns and details that
seem confusing, surprising, or unusual. In some cases, as with the `region_3`
column, it's possible to use other parts of a dataset to confirm a guess or
make sense of what's happening. When it's not, turn to first-hand or expert
knowledge, such as the people who collected the dataset or related literature
and datasets.

For unusual details that seem tangential to the primary analysis goals, make a
note and revisit them later---they can turn out to be more relevant than you
expect.
:::


### Categories Are Constructs

When working with categorical features, keep in mind that categories are
created by people. We often use categories to simplify and organize
measurements that would otherwise be nuanced and messy. This process is
double-sided: it can be clarifying, making patterns easier to see, but it can
also be reductive, obscuring meaningful details. Poorly chosen categories can
lead us into overlooking or misattributing effects.

At a minimum, for each categorical feature, ask:

* What does the feature measure?
* Why is the feature categorical?
* Who defined the categories?
* Are the categories clearly defined?
* Are there any cases that are not covered by the defined categories? In other
  words, can you think of a case that would be difficult or impossible to
  categorize?
* Are there any cases that might be of interest on their own but are placed in
  a broader category?

The key is to approach categorical features skeptically. Think carefully about
how the categories represent what they claim to measure, and try to identify
which details they gloss over.


### Frequencies and Bar Plots

To summarize a categorical feature, we can **count** or compute **frequencies**
of the categories. Taking this a step further, we can visualize the counts with
a bar plot so that categories with low or high counts stand out.

Let's make a bar plot of the `region_3` column. In R, the [ggplot2][] package's
`geom_bar` function, which makes a bar plot, counts categories automatically:

[ggplot2]: https://ggplot2.tidyverse.org/

```{r}
#| label: fig-region_3_bar
library("ggplot2")

ggplot(terns) +
  aes(x = region_3) +
  geom_bar()
```

The plot reveals two especially interesting details about `region_3`.

First, there are actually three unexpected regions: Kings, Sacramento, and
Arizona. Each has 15 observations or less and is an order of magnitude smaller
than the S.F. Bay, Central, and Southern regions.

The Sacramento region likely corresponds to [Sacramento county][sac] and
surrounding areas. Taking the same approach as we did for the Kings region, see
if you can confirm this.

[sac]: https://en.wikipedia.org/wiki/Sacramento_County,_California

<!--
```{r}
sac = filter(terns, region_3 == "SACRAMENTO")
unique(sac$site_name)
```
-->

The Arizona region is a bit of a mystery---there's no Arizona county in
California, nor are there any cities or places with Arizona in the name. Let's
see if the site names give any clues:

```{r}
arizona = filter(terns, region_3 == "ARIZONA")
unique(arizona$site_name)
```

Glendale is [a city in California][glendale-ca], but also [a city in
Arizona][glendale-az]. In this case, it seems likely that the Arizona region
corresponds to the state, and that perhaps these observations were not supposed
to be included in the dataset.

[glendale-ca]: https://en.wikipedia.org/wiki/Glendale,_California
[glendale-az]: https://en.wikipedia.org/wiki/Glendale,_Arizona

Second, the Southern region occurs 4 or 5 times more than any of the others.

The disparity in counts between the three expected regions could be due to
nesting preferences of the birds: perhaps there are more nesting sites in
southern California. On the other hand, it could also be due to how the dataset
is collected: perhaps the regions have different coverage over time, or CDFW
uses geographically smaller sites in southern California.

:::{.callout-tip}
Thinking through the different ways a dataset might have been generated like
this can help you identify subtle details that change or temper interpretations
and lead to further lines of inquiry.
:::


### Distributions {#sec-distributions}

Frequencies show us how the categories of a feature are distributed across the
observations. Because of this, we say they show us the **distribution** of the
feature. For example, @fig-region_3_bar shows us the distribution of
`region_3`.

Visualizing, estimating, summarizing, and interpreting distributions is a major
part of any data explorer or data scientist's job. The distribution in a sample
can give us insight into the distribution in a population, especially if the
sample was taken randomly.

We'll revisit the concept of distributions in @sec-numerical-features.


### The Mode {#sec-mode}

For a categorical feature, we call the category that occurs most frequently the
**mode**. In the case of a tie between two or more categories, we say the
feature has multiple modes. In a bar plot, the mode corresponds to the category
with the tallest bar. For the `region_3` column, the mode is the Southern
category.

Think of the mode of a feature as the category that's most likely if we were to
randomly select an observation from the dataset. If the dataset is a random
sample from a population, we can also think of the sample mode (computed with
the dataset) as an estimator of the population's mode.

:::{.callout-warning}
There's no built-in function to calculate the statistical mode in R. The `mode`
function is unrelated and returns something different.

Use a bar plot (or as we'll soon see, the `table` function) to identify modes.
:::


### Frequency Tables

A table is another way to display category frequencies. In R, we can use the
`table` function to count categories and return the counts in a table. For
example, the counts for `region_3` are:

```{r}
table(terns$region_3)
```

Even small tables can be easy to misread, especially if the names of the
categories are long or include numbers. Large tables make the problem worse.
Sorting by frequency partially addresses the problem:

```{r}
sort(table(terns$region_3))
```

This makes it easy to see which categories are relatively small or large, but
the categories are no longer in alphabetical order, so it's harder to find
specific categories. For ordinal data, where there's a natural order to the
categories, sorting by frequency can break that order and make tables
confusing.

We recommend bar plots rather than tables in most situations, because they work
well for comparing category frequencies even when there are a lot of
categories. As with a table, we can sort the categories by frequency in a bar
plot to make the extremes more pronounced, but these are usually evident even
without sorting.

Bar plots have two disadvantages compared to tables. First, the exact count for
a category can be difficult to read. That's certainly the case in
@fig-region_3_bar. A simple way to address this is to display the count on the
plot:

```{r}
ggplot(terns) +
  aes(x = region_3) +
  geom_bar() +
  geom_text(
    # Use counts for the labels (geom_bar also uses stat_count).
    stat = "count", aes(label = after_stat(count)),
    # Set vertical justification of the labels, so they sit atop the bars.
    vjust = -0.5
  )
```

Second, bar plots are often less legible than tables to people with low vision.
One way to get the best of both bar plots and tables is to simply include both.


### Numbers as Categories

When a discrete feature only takes a few values and those values are repeated
across many observations, we can treat it like a categorical feature. The
categories just happen to be numbers.

In the California least terns dataset, the `year` column is a good example of
this. According to the documentation, the dataset only covers about 20 years.
There are multiple observations each year, because there are many different
sites. Let's make a bar plot of frequencies for `year`:

```{r}
#| label: fig-year_bar
ggplot(terns) +
  aes(x = year) +
  geom_bar()
```

The bar plot shows that the dataset, which spans years 2000-2023, doesn't
include any observations for the years 2001-2003. It also shows that there are
roughly 30-40 observations for every included year. That might mean most sites
were consistently measured over the entire time period, or it might be a
coincidence due to mostly equal numbers of sites being added and removed over
time. The mode is 2017, and there's a distinct drop in observations in 2018,
which could indicate a major change in which sites were being monitored.

The counterpoint is a discrete feature with many values that are rarely
repeated. A good example of this is is the `total_nests` column, which is an
estimate of the total number of nests for each site-year pair. Here's a bar
plot of frequencies for `total_nests`:

```{r}
#| label: fig-total_nests_bar
ggplot(terns) +
  aes(x = total_nests) +
  geom_bar()
```

We can see that a few values close to 0 are repeated frequently, but most of
the other values are rare. Because of this and because there are so many
different values, reading the bar plot is difficult. The situation is even
worse with a table:

```{r}
table(terns$total_nests)
```

To make sense of a feature like `total_nests`, we'll need an approach
well-suited to numerical features. We'll learn several in
@sec-numerical-features.


## Numerical Features {#sec-numerical-features}

Whether the features we're exploring are categorical or numerical, what we
ultimately want to understand is the distribution (@sec-distributions): how
values are spread out across the range of possible values, which values are
common, and which are rare.

The values of a numerical feature are numbers, so in addition to calculating
how frequently they occur, we can also calculate how far apart they are. In
other words, we can use math to summarize numerical features in ways that
aren't possible for categorical features.


### Binning and Histograms {#sec-binning-histograms}

If we compute frequencies for a numerical feature, we run into a problem: most
of the frequencies are 1, because most of the values are distinct. As a
consequence, it's hard to tell whether there are patterns in the distribution
from the frequencies alone. @fig-total_nests_bar (and the associated frequency
table) demonstrates this problem.

One way to address the problem is by grouping values that are nearby each
other, a process called **binning**, and then computing frequencies for the
groups. Think of binning as slicing up the range of possible values into a few
pieces, called **bins**. Binning turns a numerical feature into a categorical
feature, so that we can use the methods from @sec-categorical-features.

When we bin a feature, we have to make choices about how many bins to use and
where their endpoints are. Different choices can lead to different conclusions,
so in an analysis that uses binning, it's important to investigate several
choices and discuss how the conclusions change. To make comparisons between
bins fair, we'll often chose bins that have equal widths, although whether this
really leads to fair comparisons depends on the nature of the data---think it
through carefully!

A **histogram** is a bar plot where the values on the x-axis are binned. Since
there are no gaps between the bins, in a histogram there are no gaps between
the bars.

We can use R to bin data and make a histogram automatically. We strongly
recommend setting the bin width manually (and this is also what the ggplot2
documentation recommends). For example, here's how to make a histogram for the
`total_nests` column with 10-nest bins:

```{r}
#| label: fig-total_nests_hist10
ggplot(terns) +
  aes(x = total_nests) +
  # Set color = "black" to outline the bars, for better legibility.
  geom_histogram(binwidth = 10, color = "black")
```

This is a bit clearer than the bar plot in @fig-total_nests_bar, because now we
can read the count for the 0 bin, whereas in the bar plot, the count for 0 was
obscured. That said, there are still a lot of bars, so let's try a wider bin
width:

```{r}
#| label: fig-total_nests_hist100
ggplot(terns) +
  aes(x = total_nests) +
  geom_histogram(binwidth = 100, color = "black")
```

Notice that this histogram has fewer sudden jumps and drops than the previous
one (@fig-total_nests_hist10). Increasing bin width has a smooting effect,
making large-scale trends easier to see but fine details harder to see.

:::{.callout-important}
When you bin a feature (whether for a histogram or some other purpose), always
check several different binning strategies, because they might lead to
different conclusions. Point out and explain the discrepancies when you share
the results. This is a form of statistical due diligence.
:::

A histogram is usually more appropriate than a bar plot for visualizing
features that represent a continuous quantity, even if they measure it in a
discrete way, because there are no gaps between the bars. For instance, the
`year` feature is discrete, but it measures time, which we usually think of as
continuous. We can make a histogram without binning by setting the bin width to
1:

```{r}
#| label: fig-year_hist1
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 1, color = "black")
```

Compared to the bar plot @fig-year_bar, the histogram makes it clear that each
bar spans an entire year.

There's no need to bin the `year` feature, and in this case, the smoothing
effect of binning actually makes it harder to interpret:

```{r}
#| label: fig-year_hist2
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 2, color = "black")
```

With 2-year bins, the drop in 2018 is still visible, but it's hard to tell
exactly which year has the drop (is it 2018 or 2019 or both?). Other details
are also obscured.

:::{.callout-note}
R's functions for making histograms assume numerical data, but we can make a
histogram for ordinal data in two steps, by first binning and then making a bar
plot.

R's `cut` function bins data (without making a histogram).
:::


### Density Plots {#sec-density-plots}

Binning and histograms help us understand the shape of the distribution of a
numerical feature, but there's still a problem. For continuous features,
histograms can overstate differences in frequency from one bin to the next,
because the edge of the bin is a sharp dividing line. We can remedy this by
blurring or smoothing the divisions between bins.

A **density** is a function that represents the shape of a distribution. Think
of a density as a smooth curve that traces the tops of the bars in a histogram.
A **density plot** is a plot that shows a density. Compared to a histogram, the
divisions between the bins are no longer visible, but the pattern of the tops
of the bars still is.

We can use R and ggplot2 to draw a density plot. Here's a density plot for the
`total_nests` feature:

```{r}
#| label: fig-total_nests_density
ggplot(terns) +
  aes(x = total_nests) +
  geom_density()
```

Notice that the trends from @fig-total_nests_hist10 are still present. The
density is high near 0, drops off quickly, and then has a long tail with a few
small peaks.

Although the density plot has the same shape as the histogram, the quantity on
the y-axis is not counts or frequencies. Instead, we scale the density curve
the that total area underneath it is 1 or 100%. This is because the area under
any part of the density is an estimate of the probability of randomly drawing
the corresponding values on the x-axis.

Mathematically, we estimate a density by putting a small, smooth "bump"
function, called a **kernel**, at each observed value of the feature and then
adding all of these up. How much the bump spreads out, called the
**bandwidth**, is a bit like a bin width, with larger bandwidths leading to
smoother densities.

When we make a density plot, R uses a statistical formula to compute a default
bandwidth. The default bandwidth often strikes a good balance between showing
coarse trends and fine details. We can change the bandwidth through the
`adjust` parameter, which is multiplied by the default (so `adjust = 0.5` means
half of the default). This way the bandwidth is still estimated from and on a
similar scale as the data. Here's the density plot for `total_nests` with a
smaller bandwidth:

```{r}
#| label: fig-total_nests_density0.5
ggplot(terns) +
  aes(x = total_nests) +
  geom_density(adjust = 0.5)
```

And here it is with a larger bandwidth:

```{r}
#| label: fig-total_nests_density2
ggplot(terns) +
  aes(x = total_nests) +
  geom_density(adjust = 2)
```

Decreasing the bandwidth makes the density rougher, while increasing it makes
it smoother. As with bin width for histograms, it's important to try different
bandwidths for density plots, since different amounts of smoothing can lead to
different conclusions.

The lack of bin edges in density plots can make them easier to read than
histograms, but an important limitation to keep in mind is that density plots
are a kind of estimate, and as a consequence, they're sensitive to the number
of observations in the dataset. For a dataset with a relatively small number of
observations (for example, less than 30), a histogram or bar plot is a more
faithful representation.

Density plots are especially helpful for comparing how a feature is distributed
across multiple groups, because we can display all of the densities in a single
plot. Let's try this for the `total_nests` feature by using the regions from
`region_3` as groups. We'll first filter out the unexpected regions Arizona,
Kings, and Sacramento, because they have low numbers of observations. Here's
the plot:

```{r}
#| label: fig-total_nests_region_3_density
EXPECTED_REGION_3 = c("S.F._BAY", "CENTRAL", "SOUTHERN")
terns_region = filter(terns, region_3 %in% EXPECTED_REGION_3)

ggplot(terns_region) +
  aes(x = total_nests, color = region_3) +
  geom_density()
```

This reveals that the nest counts for Central tend to be low in comparison to
S.F. Bay and Southern. Curiously, a substantial proportion of nest counts for
S.F. Bay are between 250 and 500. This could mean that there are sites in the
S.F. Bay region that are sufficiently well-protected that large flocks of terns
use them every year, whereas other sites see varying numbers of birds each
year, depending on their condition. There's certainly more we could hypothesize
based on the plot and investigate further using other features. This is an
example of exploring two features simultaneously, which we'll learn more about
in @sec-relationships-between-features.


### Location Statistics

So far, we've only used visualizations to summarize numerical features. In some
cases, we might also want to use quantitative summaries. We'll focus on
quantities that attempt to answer two questions about a feature's distribution:

1. What is the **location** of the data? For 1-dimensional data, where are most
   of the observations on the number line? For higher-dimensional data, where
   are most of the observations in the space?
2. What is the **scale** of the data? In other words, how much do the
   observations spread out?

We can also ask other questions about the distribution, such as how many peaks
it has and whether it's symmetrical around some central point, but these are
often easier to answer with visualizations. Asking about location and scale is
a sensible starting point.

The questions leave how to quantify location and scale open to interpretation,
but there are a few widely-used and well-studied choices. The remainder of this
section introduces location statistics, and @sec-scale-statistics introduces
scale statistics.

To help develop intuition, we'll use this small, artificial "toy" dataset in
several examples:

```{r}
toy = c(-2, -1, -1, -1, 0, 2, 6)
```

We'll also use the `total_nests` feature from the California least terns
dataset. Recall the density plot for the feature (@fig-total_nests_density).
Where would you say the distribution is located, if you had to pick just one
point?


#### Mode

First, let's think of location as the value that occurs most frequently. This
is the mode, which we introduced for categorical features in @sec-mode. For
continuous features, the mode is at the tallest peak of the density curve.
Values at or near the mode occur more frequently than any others.

For the toy dataset, the mode is -1, because it occurs more than any of the
other values. If we didn't want to count ourselves, we could compute this with
the `table` function.

For `total_nests`, the mode is at or close to 0, depending on whether we treat
the feature as discrete and estimate with the frequencies or treat it as
continuous and estimate with the density curve.

An advantage of the mode is that we can compute it for any type of feature,
categorical or numerical.

The mode also has some disadvantages. The mode might occur only slightly more
often than other values, and it might be far away from many other values that
can occur. The latter is certainly the case for `total_nests`, where the mode
is 0 but some of the values exceed 1,500.

#### Median {#sec-median}

Next, let's think of location as the midpoint of the values. In other words,
we'll sort the values and find the one in the middle. This statistic is called
the **median**.

In R, the `median` function computes the median of a feature. For the toy
dataset, the median is:

```{r}
median(toy)
```

This means half of the observed values are less than or equal to -1 and half
are greater than or equal. By coincidence, the median is the same as the mode.
So -1 also occurs more frequently than any other value.

For `total_nests`, the median is:

```{r}
median(terns$total_nests, na.rm = TRUE)
```

It's meaningful that the median is larger than the mode (0). The median tells
us half of the observed values of `total_nests` are less than or equal to 42
and half are greater than or equal. Even though the mode is more common than
any other value, if we only paid attention to it, we'd be ignoring more than
half of the observed values!

:::{.callout-note title="Note: Removing Missing Values"}
Most statistical summary functions in R provide an `na.rm` parameter to control
whether missing values are removed before computing the statistic. By default,
it's `FALSE` and missing values are kept. Set `na.rm = TRUE` to remove the
missing values. This is useful for quick, initial investigations of a dataset.

In a complete analysis, you should think carefully about what the missing
values mean, whether they exhibit any patterns, and whether there are enough
non-missing values for statistical summaries to be good representatives of the
data. We'll learn more about this in @sec-missing-values.
:::

Another name for the median is the **50th percentile**, because the median is
greater than roughly 50% of the values. We can also estimate other percentiles.
For instance, the 25th percentile is the value one-quarter of the way through
the sorted data. **Quantile** is another word for percentile.

In R, we can compute percentiles with the `quantile` function:

```{r}
quantile(toy)
quantile(terns$total_nests, na.rm = TRUE)
```

Statisticians call the median a **robust** statistic, because it's not affected
much by extreme (very small or very large) values. For example, if we change
the largest value 6 in the toy dataset to 1,000,000, the median stays the same:

```{r}
toy2 = toy
toy2[match(6, toy2)] = 1000000
median(toy2)
```


Likewise, if we put a new value, 100, in the dataset, the median only increases
slightly:

```{r}
toy2 = c(toy2, 100)
median(toy2)
```

Robustness is a useful property because extremes rarely represent a majority of
the data---that's why they're extremes. Sometimes extreme values can even be
erroneous, but the median is relatively unaffected.

The cost of robustness is that the median uses information less efficiently
than the next statistic we'll learn about (@sec-mean). To see this, think about
how we compute the median: we only use the values in the middle, but the other
values might also contain important information.

#### Mean {#sec-mean}

Finally, let's think of location as a balancing point for the values. Imagine
trying to balance the distribution (picture a density plot, such as
@fig-total_nests_density) on a triangular wedge. Where would you have to place
the wedge? This statistic is called the **mean** or **average**.

:::{.callout-note title="Definition: Sample Mean"}
Mathematically, for data $x_1, x_2, \dots, x_n$, the sample mean $\bar{x}$ is
defined as
$$
\begin{align*}
\bar{x}
&= \frac{1}{n} (x_1 + x_2 + \dots + x_n) \\
&= \frac{1}{n} \sum_{i=1}^n x_i \\
\end{align*}
$$

In words, we add up all of the values and then divide by the number of values.

Statisticians conventionally write an overbar to indicate a sample mean, as in
$\bar{x}$. They write the unknown population mean, which is estimated by the
sample mean, as the [Greek letter][greek] $\mu$.

[greek]: https://en.wikipedia.org/wiki/Greek_letters_used_in_mathematics,_science,_and_engineering
:::

In R, the `mean` function computes the mean of a feature. For the toy dataset,
the mean is:

```{r}
mean(toy)
```

Notice that the mean is quite a bit larger than the median (-1). This is
because of the value 6, which is relatively far from the other values and thus
exerts a lot of influence (in the balancing metaphor, it has a lot of
leverage).

For `total_nests`, the mean is:

```{r}
mean(terns$total_nests, na.rm = TRUE)
```

Once again, the mean is larger than than the median (42). As we saw in the
density plot (@fig-total_nests_density), the `total_nests` feature contains a
few very large values. These don't affect the median much, but they do affect
on the mean.

The mean is good for getting a general idea of where the center of the data is.
It's more efficient than the median for this purpose, because it takes all of
the data into account.

On the other hand, the mean is sensitive to extreme values, especially in
comparison to the median. If the mean seems unexpectedly small or large, check
for extreme values.

:::{.callout-tip}
We recommend using the mean and median together to get the best of both.
Comparing the mean and median can provide insight into how asymmetric the
distribution is. For instance, if the mean is much larger than the median, then
the distribution likely has a long right tail. Of course, a visualization is
still generally a better summary than any number of statistics.
:::

:::{.callout-note title="Note: Other Means" collapse="true"}
There are other ways to define the mean, with different advantages and
disadvantages.

The mean defined in this section is formally called the **arithmetic mean**.
The arithmetic mean is well-suited to values that can be combined naturally by
adding, such as physical measurements (length, weight, etc.) and counts.

To learn more about other ways to define the mean, see [How to Choose the Right
Mean for Your Data][choose-mean] from @mean-calculator.

[choose-mean]: https://meancalculator.com/choosing-right-mean.html
:::


### Scale Statistics {#sec-scale-statistics}

The scale of a distribution refers to how spread out its values are. As with
location, there are many different ways to define scale and thus many different
statistics. We'll focus on two popular scale statistics.


#### Interquartile Range {#sec-interquartile-range}

One way to think about the scale of a distribution is as the distance between
its tails. If we take a representative value from each tail, we can compute the
distance between them by subtracting. The question is: which values should we
choose?

The minimum and maximum might seem tempting, since they're the farthest along
each tail. But that becomes a problem if most of the data lie close to the
center, with the minimum and maximum among just a few extreme values. By using
them, we'd overestimate how much the data spread out. The minimum and maximum
can also vary substantially from one sample to the next, which means our scale
estimates could as well---they'd be unreliable for estimating the scale of the
population.

A better choice is something near the tails but less sensitive to extremes.
@sec-median explained that the median is robust, meaning it's mostly unaffected
by extreme values. It turns out most other percentiles are also relatively
robust. We can use the distance between the 75th and 25th percentile as a
robust estimator for scale. Conveniently, we can also interpret this as the
width of the middle 50% of the data.

The 25th, 50th, and 75th percentiles are called **quartiles** (with an "r")
because they separate the data into four parts (hence "quart-"). So the
distance between the 75th and and 25th percentile is called the **interquartile
range** (IQR).

In R, the `IQR` function computes the IQR of a feature. For the toy dataset,
the IQR is:

```{r}
IQR(toy)
```

As a reminder, the dataset looks like this:

```{r}
toy
```

And the quartiles are:

```{r}
quantile(toy)
```

The IQR is 2 because the 25th percentile is -1 and the 75th percentile is 1.
The middle 50% of the data is 2 units wide.

As we mentioned, IQR is relatively unaffected by extreme values. If we change
the largest value 6 in the toy dataset to 1,000,000, the IQR stays the same.
The IQR has the same trade-offs as the median: it's a robust estimator of
scale, but it's less efficient than estimators that use all of the data.

For `total_nests`, the IQR is:

```{r}
IQR(terns$total_nests, na.rm = TRUE)
```

This means the middle 50% of the data is 159 nests wide. We can gain a better
understanding of the scale by examining the quartiles, too:

```{r}
quantile(terns$total_nests, na.rm = TRUE)
```

Notice that the 25th percentile is closer to the median than the 75th
percentile. The large IQR shows that the data are very spread out; the
quartiles show that most of that spread is in the right tail. In other words,
the distribution is asymmetrical, as we saw in @fig-total_nests_density.

:::{.callout-tip}
Visualizations are almost always the best way to study distributions. If you
can't make a visualization, examine multiple statistics rather than just one. A
lone statistic is unable to capture nuance and can even be misleading.
:::


#### Standard Deviation {#sec-standard-deviation}

Another way to think about scale is as the typical distance from any value to
the center of the distribution. This interpretation raises three questions:

1. What do we mean by "typical" distance?
2. How do we measure distance?
3. What do we mean by the "center" of the distribution?

As usual, different answers to these questions lead to different statistics.
For our purposes, we'll say:

1. Typical distance is mean distance.
2. We measure distance as the squared difference. Squaring ensures that the
   distances are non-negative.
3. The center of the distribution is the sample mean.

The difference between a point and a mean is called a **deviation**. If we
square the deviations and compute their mean, the result is called the
**variance** and is measured in squared units. To get back to the original
units, we can take a square root of the variance. This is called the **standard
deviation**.

:::{.callout-note
  title="Definition: Sample Variance and Sample Standard Deviation"
}
Mathematically, for data $x_1, x_2, \dots, x_n$ with sample mean $\bar{x}$
(@sec-mean), the sample variance $s^2$ is defined as

$$
\begin{align*}
s^2
&= \frac{1}{n-1} \left[
  (x_1 - \bar{x})^2 + (x_2 - \bar{x})^2 + \dots + (x_n - \bar{x})^2
\right] \\
&= \frac{1}{n-1} \sum_{i=0}^n (x_i - \bar{x})^2 \\
\end{align*}
$$

In words, we compute the mean of the squared distances to the sample mean. We
divide by $n - 1$ rather than $n$ as a statistical correction for the fact that
$\bar{x}$ is also an estimator.

The sample standard deviation $s$ is defined as the square root of the variance
$s^2$.

Statisticians conventionally write the unknown population variance, which is
estimated by the sample variance, as $\sigma^2$. Thus they write the unknown
population standard deviation as $\sigma$.
:::

Think of the standard deviation as the average distance from a value to the
mean. Larger standard deviations mean the data are more spread out. As a rule
of thumb, most of the data will be within 3 standard deviations of the mean.

In R, the `sd` function computes the standard deviation of a feature. For the
toy dataset, the standard deviation is:

```{r}
sd(toy)
```

This means that on average, values are about 2.76 units from the mean 0.43.
That's roughly consistent with the range of values in the toy dataset.

For `total_nests`, the standard deviation is:

```{r}
sd(terns$total_nests, na.rm = TRUE)
```

This means that on average, the number of nests is 162 (the mean) plus or minus
293. We know the number of nests can't be negative, so the lower bound is
actually 0. The large standard deviation reflects both the numerous 0 values on
the left and the very large values on the right.

The standard deviation and IQR are analogous to the mean and median,
respectively. The standard deviation is not robust, but uses the available data
more efficiently. We recommend examining both.


## A Final Note

The `summary` function computes a detailed statistical summary of an R object.
For instance, here's the summary for the `total_nests` column:

```{r}
summary(terns$total_nests)
```

For data frames, the `summary` function computes a summary of each column,
guessing an appropriate statistic based on the column's data type:

```{r}
summary(terns)
```

These summaries are useful as a quick overview of an entire dataset. We
strongly recommend following up with visualizations to develop a more nuanced
understanding of individual features and their distributions.

:::{.callout-note title="See Also"}
[Anscombe's Quartet][anscombe] [@anscombe1973] and the [Datasaurus
Dozen][datasaurus] [@datasaurus] are two datasets that exemplify the importance
of making visualizations. For these datasets, statistics alone lead to
completely incorrect conclusions.

[anscombe]: https://en.wikipedia.org/wiki/Anscombe%27s_quartet
[datasaurus]: https://www.research.autodesk.com/publications/same-stats-different-graphs/
:::

