```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Statistical Summaries {#sec-data-forensics}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the location statistics mean, median, and mode
* Explain the scale statistics range and standard deviation
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

This chapter covers a variety of ways to investigate and summarize tabular data
in order to understand the data better and identify potential problems. The
lesson also describes how to fix some of the most common data problems.

## What Is a Distribution?


## Categorical Data


## Numerical Data

After investigating the data's structure, it's a good idea to check some basic
statistical properties. This step is important because it can help you identify
limitations of and patterns in the data.

Which statistics are appropriate for a given feature often depends on the type
of the feature. Recall from @sec-types-of-data that the types data scientists
typically think about are:

* **Categorical**
    + **Nominal** - data separated into specific categories, with no order. For
      example, hair color (red, brown, blonde, ...) is categorical.
    + **Ordinal** - data separated into specific categories, with an order. For
      example, school level (elementary, middle, high, college) is ordinal.
* **Numerical**
    + **Discrete** - integers, or a finite set of decimal numbers with no
      values in between. Sometimes discrete values can also be treated as
      ordinal. For example, month as a number (1, 2, ..., 12) is discrete.
    + **Continuous** - decimal numbers. There are no specific categories, but
      there is an order. For example, height in inches is numerical.

The `table` function is great for summarizing categorical (and sometimes
discrete) data. For example:

```{r}
table(terns$region_3)
```

What about numerical data?

Two important questions to ask about data are:

1. Where is it? This is the **location** of the data.

2. How spread out is it? This is the **scale** of the data.

Let's use the data

```{r}
x = c(-2, -1, -1, -1, 0, 2, 6)
```

as an example.

Location is generally summarized with a number near the middle or center of
the data. A few options are:

1.  **Mode** - the value that appears most frequently. The mode can be
    calculated for any kind of data, but doesn't work well for continuous data.

    For our example, the mode of `x` is -1. You can compute the mode with
    `table`:

    ```{r}
    table(x)
    ```

2.  **Median** - sort the data, then find the value in the middle. The median
    can be calculated for ordinal or numerical data.

    For our example, the median is -1. Compute this with `median`:

    ```{r}
    median(x)
    ```

3.  **Mean** - the balancing point of the data, if a waiter was trying to
    balance the data on a tray. The mean can only be calculated for numerical
    data.

    For our example the mean is 0.4285. Compute this with `mean`:

    ```{r}
    mean(x)
    ```

Adding large values to the data affects the mean more than the median:

```{r}
y = c(x, 100)
mean(y)
median(y)
```

Because of this, we say that the median is **robust**.

The mean is good for getting a general idea of where the center of the data
is, while comparing it with the median reveals whether there are any
unusually large or small values.

Scale is generally summarized by a number that says how far the data is from
the center (mean, median, etc...). Two options are:

1.  **Standard Deviation** - square root of the average squared distance to the
    mean (the distance from a point to a mean is called a **deviation**). You
    can think of this as approximately the average distance from a data point
    to the mean. As a rule of thumb, most of the data will be within 3 standard
    deviations of the mean.

    You can compute the standard deviation with `sd`:

    ```{r}
    sd(x)
    ```

2.  **Interquartile Range** (IQR) - difference between the 75th and 25th
    percentile. The median is the 50th **percentile** of the data; it's at the
    middle of the sorted data. We can also consider other percentiles. For
    instance, the 25th percentile is the value one-quarter of the way through
    the sorted data.

    **Quantile** is another word for percentile. **Quartile** specifically
    refers to the 25th, 50th, and 75th percentiles because they separate the
    data into four parts (hence "quart-").

    You can compute quantiles with `quantile`, or compute the IQR directly with
    `IQR`:

    ```{r}
    quantile(x)

    # IQR
    IQR(x)
    ```

The IQR is more robust than the standard deviation.

Many of the functions for computing statistical summaries have a parameter
`na.rm` to ignore missing values. Setting `na.rm = TRUE` is often useful when
you're just trying to do an initial investigation of the data. However, in a
more complete analysis, you should think carefully about what the missing
values mean, whether they follow any patterns, and whether there are enough
non-missing values for statistical summaries to be good representatives of the
data.

Finally, the `summary` function computes a detailed statistical summary of an R
object. For data frames, the function computes a summary of each column,
guessing an appropriate statistic based on the column's data type.
