```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Statistical Summaries {#sec-data-forensics}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the location statistics mean, median, and mode
* Explain the scale statistics range and standard deviation
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

After investigating (and possibly transforming) a dataset's structure, the next
step is to examine the statistical properties of its features. This chapter
focuses on individual features, and presents a variety of statistical
visualizations and estimates to summarize them, whether they're categorical or
numerical. These methods will help us identify and communicate trends,
patterns, anomalies, and limitations. For each method, we discuss how to
interpret the summary and explain the method's strengths and weaknesses
relative to others. The unifying theme of the chapter is summarizing, making
sense of, and communicating how data are distributed within populations and
samples.


## Categorical Features {#sec-categorical-features}

To introduce ideas gently, we'll start by focusing on summaries for categorical
features. Most of the time, categories are not numbers, so we won't use much
math, and when we do, it will mostly be counting.

As motivation and a source of examples, we'll work to understand which regions
of California and which time periods are recorded in the California least terns
dataset (@sec-ca-least-terns). We'll also examine whether regions and time
periods are equally represented in the dataset.

To refresh our memory, the structure of the California least terns dataset is:

```{r}
str(terns)
```

We can study regional coverage at a coarse scale by looking at the `region_3`
or `region_4` columns. Let's focus on `region_3`. According to the
documentation, it's categorical, with categories for three regions. The first
few values are:

```{r}
head(terns$region_3)
```

The documentation claims the three regions are S.F. Bay, Central, and Southern,
but we can already see that this isn't accurate. There's another region: Kings.

The Kings region likely corresponds to sites in [Kings County][kings]. We can
try to confirm this by examining the site names in the region (with R's
[dplyr][] package):

[kings]: https://en.wikipedia.org/wiki/Kings_County,_California
[dplyr]: https://dplyr.tidyverse.org/

```{r}
#| warning: false
library("dplyr")

kings = filter(terns, region_3 == "KINGS")
unique(kings$site_name)
```

[Kettleman City][kc] is indeed a place in Kings County, so it seems like our
guess was correct.

[kc]: https://en.wikipedia.org/wiki/Kettleman_City,_California

:::{.callout-tip}
As you explore data, make a habit of following up on patterns and details that
seem confusing, surprising, or unusual. In some cases, as with the `region_3`
column, it's possible to use other parts of a dataset to confirm a guess or
make sense of what's happening. When it's not, turn to first-hand or expert
knowledge, such as the people who collected the dataset or related literature
and datasets.

For unusual details that seem tangential to the primary analysis goals, make a
note and revisit them later---they could turn out to be more relevant than you
expect.
:::


### Categories Are Constructs

When working with categorical features, keep in mind that categories are
created by people. We often use categories to simplify and organize
measurements that would otherwise be nuanced and messy. This process is
double-sided: it can be clarifying, making patterns easier to see, but it can
also be reductive, obscuring meaningful details. Poorly chosen categories can
lead us into overlooking or misattributing effects.

At a minimum, for each categorical feature, ask:

* What does the feature measure?
* Why is the feature categorical?
* Who defined the categories?
* Are the categories clearly defined?
* Are there any cases that are not covered by the defined categories? In other
  words, can you think of a case that would be difficult or impossible to
  categorize?
* Are there any cases that might be of interest on their own but are placed in
  a broader category?

The key is to approach categorical features skeptically. Think carefully about
how the categories represent what they claim to measure, and try to identify
which details they gloss over.


### Frequencies

To summarize a categorical feature, we can **count** or compute **frequencies**
of the categories. Taking this a step further, we can visualize the counts with
a bar plot so that categories with low or high counts stand out.

Let's make a bar plot of the `region_3` column. In R, the [ggplot2][] package's
`geom_bar` function, which makes a bar plot, counts categories automatically:

[ggplot2]: https://ggplot2.tidyverse.org/

```{r}
#| label: fig-region_3_bar
library("ggplot2")

ggplot(terns) +
  aes(x = region_3) +
  geom_bar()
```

The plot reveals two especially interesting details about `region_3`.

First, there are actually three unexpected regions: Kings, Sacramento, and
Arizona. Each has 15 observations or less and is an order of magnitude smaller
than the S.F. Bay, Central, and Southern regions.

The Sacramento region likely corresponds to [Sacramento county][sac] and
surrounding areas. Taking the same approach as we did for the Kings region, see
if you can confirm this.

[sac]: https://en.wikipedia.org/wiki/Sacramento_County,_California

<!--
```{r}
sac = filter(terns, region_3 == "SACRAMENTO")
unique(sac$site_name)
```
-->

The Arizona region is a bit of a mystery---there's no Arizona county in
California, nor are there any cities or places with Arizona in the name. Let's
see if the site names give any clues:

```{r}
arizona = filter(terns, region_3 == "ARIZONA")
unique(arizona$site_name)
```

Glendale is [a city in California][glendale-ca], but also [a city in
Arizona][glendale-az]. In this case, it seems likely that the Arizona region
corresponds to the state, and that perhaps these observations were not supposed
to be included in the dataset.

[glendale-ca]: https://en.wikipedia.org/wiki/Glendale,_California
[glendale-az]: https://en.wikipedia.org/wiki/Glendale,_Arizona

Second, the Southern region occurs 4 or 5 times more than any of the others.

The disparity in counts between the three expected regions could be due to
nesting preferences of the birds: perhaps there are more nesting sites in
southern California. On the other hand, it could also be due to how the dataset
is collected: perhaps the regions have different coverage over time, or CDFW
uses geographically smaller sites in southern California.

:::{.callout-tip}
Thinking through the different ways a dataset might have been generated like
this can help you identify subtle details that change or temper interpretations
and lead to further lines of inquiry.
:::


### Distributions {#sec-distributions}

Frequencies show us how the categories of a feature are distributed across the
observations. Because of this, we say they show us the **distribution** of the
feature. For example, @fig-region_3_bar shows us the distribution of
`region_3`.

Visualizing, estimating, summarizing, and interpreting distributions is a major
part of any data explorer or data scientist's job. The distribution in a sample
can give us insight into the distribution in a population, especially if the
sample was taken randomly.

We'll revisit the concept of distributions in @sec-numerical-features.


### The Mode

For a categorical feature, we call the category that occurs most frequently the
**mode**. In the case of a tie between two or more categories, we say the
feature has multiple modes. In a bar plot, the mode corresponds to the category
with the tallest bar. For the `region_3` column, the mode is the Southern
category.

Think of the mode of a feature as the category that's most likely if we were to
randomly select an observation from the dataset. If the dataset is a random
sample from a population, we can also think of the sample mode (computed with
the dataset) as an estimate of the population's mode.

:::{.callout-warning}
There's no built-in function to calculate the statistical mode in R. The `mode`
function is unrelated and returns something different.

Use a bar plot (or as we'll soon see, the `table` function) to identify modes.
:::


### Tables Versus Bar Plots

A table is another way to display category frequencies. In R, we can use the
`table` function to count categories and return the counts in a table. For
example, the counts for `region_3` are:

```{r}
table(terns$region_3)
```

Even small tables can be easy to misread, especially if the names of the
categories are long or include numbers. Large tables make the problem worse.
Sorting by frequency partially addresses the problem:

```{r}
sort(table(terns$region_3))
```

This makes it easy to see which categories are relatively small or large, but
the categories are no longer in alphabetical order, so it's harder to find
specific categories. For ordinal data, where there's a natural order to the
categories, sorting by frequency can break that order and make tables
confusing.

We recommend bar plots rather than tables in most situations, because they work
well for comparing category frequencies even when there are a lot of
categories. As with a table, we can sort the categories by frequency in a bar
plot to make the extremes more pronounced, but these are usually evident even
without sorting.

Bar plots have two disadvantages compared to tables. First, the exact count for
a category can be difficult to read. That's certainly the case in
@fig-region_3_bar. A simple way to address this is to display the count on the
plot:

```{r}
ggplot(terns) +
  aes(x = region_3) +
  geom_bar() +
  geom_text(
    # Use counts for the labels (geom_bar also uses stat_count).
    stat = "count", aes(label = after_stat(count)),
    # Set vertical justification of the labels, so they sit atop the bars.
    vjust = -0.5
  )
```

Second, bar plots are often less legible than tables to people with low vision.
One way to get the best of both bar plots and tables is to simply include both.


### Numbers as Categories

When a discrete feature only takes a few values and those values are repeated
across many observations, we can treat it like a categorical feature. The
categories just happen to be numbers.

In the California least terns dataset, the `year` column is a good example of
this. According to the documentation, the dataset only covers about 20 years.
There are multiple observations each year, because there are many different
sites. Let's make a bar plot of frequencies for `year`:

```{r}
#| label: fig-year_bar
ggplot(terns) +
  aes(x = year) +
  geom_bar()
```

The bar plot shows that the dataset, which spans years 2000-2023, doesn't
include any observations for the years 2001-2003. It also shows that there are
roughly 30-40 observations for every included year. That might mean most sites
were consistently measured over the entire time period, or it might be a
coincidence due to mostly equal numbers of sites being added and removed over
time. The mode is 2017, and There's a distinct drop in observations in 2018,
which could indicate a major change in which sites were being monitored.

The counterpoint is a discrete feature with many values that are rarely
repeated. A good example of this is is the `total_nests` column, which is an
estimate of the total number of nests for each site-year pair. Here's a bar
plot of frequencies for `total_nests`:

```{r}
#| label: fig-total_nests_bar
ggplot(terns) +
  aes(x = total_nests) +
  geom_bar()
```

We can see that a few values close to 0 are repeated frequently, but most of
the other values are rare. Because of this and because there are so many
different values, reading the bar plot is difficult. The situation is even
worse with a table:

```{r}
table(terns$total_nests)
```

To make sense of a feature like `total_nests`, we'll need an approach
well-suited to numerical features. We'll learn several in
@sec-numerical-features.


## Numerical Features {#sec-numerical-features}

Whether the features we're exploring are categorical or numerical, what we
ultimately want to understand is the distribution (@sec-distributions): how
values are spread out across the range of possible values, which values are
common, and which are rare.

The values of a numerical feature are numbers, so in addition to calculating
how frequently they occur, we can also calculate how far apart they are. In
other words, we can use math to summarize numerical features in ways that
aren't possible for categorical features.


### Binning

If we compute frequencies for a numerical feature, we run into a problem: most
of the frequencies are 1, because most of the values are distinct. As a
consequence, it's hard to tell whether there are patterns in the distribution
from the frequencies alone. @fig-total_nests_bar (and the associated frequency
table) demonstrates this problem.

One way to address the problem is by grouping values that are nearby each
other, a process called **binning**, and then computing frequencies for the
groups. Think of binning as slicing up the range of possible values into a few
pieces, called **bins**. Binning turns a numerical feature into a categorical
feature, so that we can use the methods from @sec-categorical-features.

When we bin a feature, we have to make choices about how many bins to use and
where their endpoints are. Different choices can lead to different conclusions,
so in an analysis that uses binning, it's important to investigate several
choices and discuss how the conclusions change. To make comparisons between
bins fair, we'll often chose bins that have equal widths, although whether this
really leads to fair comparisons depends on the nature of the data---think it
through carefully!

A **histogram** is a bar plot where the values on the x-axis are binned. Since
there are no gaps between the bins, in a histogram there are no gaps between
the bars.

We can use R to bin data and make a histogram automatically. We strongly
recommend setting the bin width manually (and this is also what the ggplot2
documentation recommends). For example, here's how to make a histogram for the
`total_nests` column with 10-nest bins:

```{r}
#| label: fig-total_nests_hist10
ggplot(terns) +
  aes(x = total_nests) +
  # Set color = "black" to outline the bars, for better legibility.
  geom_histogram(binwidth = 10, color = "black")
```

This is a bit clearer than the bar plot in @fig-total_nests_bar, because now we
can read the count for the 0 bin, whereas in the bar plot, the count for 0 was
obscured. That said, there are still a lot of bars, so let's try a wider bin
width:

```{r}
#| label: fig-total_nests_hist100
ggplot(terns) +
  aes(x = total_nests) +
  geom_histogram(binwidth = 100, color = "black")
```

Notice that this histogram has fewer sudden jumps and drops than the previous
one (@fig-total_nests_hist10). Increasing bin width has a smooting effect,
making large-scale trends easier to see but fine details harder to see.

:::{.callout-important}
When you bin a feature (whether for a histogram or some other purpose), always
check several different binning strategies, because they might lead to
different conclusions. Point out and explain the discrepancies when you share
the results. This is a form of statistical due diligence.
:::

A histogram is usually more appropriate than a bar plot for visualizing
features that represent a continuous quantity, even if they measure it in a
discrete way, because there are no gaps between the bars. For instance, the
`year` feature is discrete, but it measures time, which we usually think of as
continuous. We can make a histogram without binning by setting the bin width to
1:

```{r}
#| label: fig-year_hist1
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 1, color = "black")
```

Compared to the bar plot @fig-year_bar, the histogram makes it clear that each
bar spans an entire year.

There's no need to bin the `year` feature, and in this case, the smoothing
effect of binning actually makes it harder to interpret:

```{r}
#| label: fig-year_hist2
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 2, color = "black")
```

With 2-year bins, the drop in 2018 is still visible, but it's hard to tell
exactly which year has the drop (is it 2018 or 2019 or both?). Other details
are also obscured.

:::{.callout-note}
R's functions for making histograms assume numerical data, but we could make a
histogram for ordinal data in two steps, by first binning and then making a bar
plot.

R's `cut` function bins data (without making a histogram).
:::


### Density Plots

Binning and histograms help us understand the shape of the distribution of a
numerical feature, but there's still a problem. For continuous features,
histograms can overstate differences in frequency from one bin to the next,
because the edge of the bin is a sharp dividing line. We can remedy this by
blurring or smoothing the divisions between bins.

A **density** is a function that represents the shape of a distribution. Think
of a density as a smooth curve that traces the tops of the bars in a histogram.
A **density plot** is a plot that shows a density. Compared to a histogram, the
divisions between the bins are no longer visible, but the pattern of the tops
of the bars still is.

We can use R and ggplot2 to draw a density plot. Here's a density plot for the
`total_nests` feature:

```{r}
#| label: fig-total_nests_density
ggplot(terns) +
  aes(x = total_nests) +
  geom_density()
```

Notice that the trends from @fig-total_nests_hist10 are still present. The
density is high near 0, drops off quickly, and then has a long tail with a few
small peaks.

Although the density plot has the same shape as the histogram, the quantity on
the y-axis is not counts or frequencies. Instead, we scale the density curve
the that total area underneath it is 1 or 100%. This is because the area under
any part of the density is an estimate of the probability of randomly drawing
the corresponding values on the x-axis.

Mathematically, we estimate a density by putting a small, smooth "bump"
function, called a **kernel**, at each observed value of the feature and then
adding all of these up. How much the bump spreads out, called the
**bandwidth**, is a bit like a bin width, with larger bandwidths leading to
smoother densities.

When we make a density plot, R uses a statistical formula to compute a default
bandwidth. The default bandwidth often strikes a good balance between showing
coarse trends and fine details. We can change the bandwidth through the
`adjust` parameter, which is multiplied by the default (so `adjust = 0.5` means
half of the default). This way the bandwidth is still estimated from and on a
similar scale as the data. Here's the density plot for `total_nests` with a
smaller bandwidth:

```{r}
#| label: fig-total_nests_density0.5
ggplot(terns) +
  aes(x = total_nests) +
  geom_density(adjust = 0.5)
```

And here it is with a larger bandwidth:

```{r}
#| label: fig-total_nests_density2
ggplot(terns) +
  aes(x = total_nests) +
  geom_density(adjust = 2)
```

Decreasing the bandwidth makes the density rougher, while increasing it makes
it smoother. As with bin width for histograms, it's important to try different
bandwidths for density plots, since different amounts of smoothing can lead to
different conclusions.

The lack of bin edges in density plots can make them easier to read than
histograms, but an important limitation to keep in mind is that density plots
are a kind of estimate, and as a consequence, they're sensitive to the number
of observations in the dataset. For a dataset with a relatively small number of
observations (for example, less than 30), a histogram or bar plot is a more
faithful representation.

Density plots are especially helpful for comparing how a feature is distributed
across multiple groups, because we can display all of the densities in a single
plot. Let's try this for the `total_nests` feature by using the regions from
`region_3` as groups. We'll first filter out the unexpected regions Arizona,
Kings, and Sacramento, because they have low numbers of observations. Here's
the plot:

```{r}
#| label: fig-total_nests_region_3_density
EXPECTED_REGION_3 = c("S.F._BAY", "CENTRAL", "SOUTHERN")
terns_region = filter(terns, region_3 %in% EXPECTED_REGION_3)

ggplot(terns_region) +
  aes(x = total_nests, color = region_3) +
  geom_density()
```

This reveals that the nest counts for Central tend to be low in comparison to
S.F. Bay and Southern. Curiously, a substantial proportion of nest counts for
S.F. Bay are between 250 and 500. This could mean that there are sites in the
S.F. Bay region that are sufficiently well-protected that large flocks of terns
use them every year, whereas other sites see varying numbers of birds each
year, depending on their condition. There's certainly more we could hypothesize
based on the plot and investigate further using other features. This is an
example of exploring two features simultaneously, which we'll learn more about
in @sec-relationships-between-features.


### Location Statistics

Two questions that make good starting points when exploring numerical data are:

1. What is the **location** of the data? For 1-dimensional data, where are most
   observations on the number line? For 2-, 3-, or higher-dimensional data,
   where are most of the observations in the space?
2. What is the **scale** of the data? In other words, how much does it spread
   out?

To illustrate a few ways to approach these questions, let's use this toy data:

```{r}
x = c(-2, -1, -1, -1, 0, 2, 6)
```

One way to summarize location is to try to find the middle or center of the
data. Two different statistics for this are:

1.  **Mode** - the value that appears most frequently. The mode can be
    calculated for any kind of data, but doesn't work well for continuous data.

    For our example, the mode of `x` is -1. You can compute the mode with
    `table`:

    ```{r}
    table(x)
    ```

2.  **Median** - sort the data, then find the value in the middle. The median
    can be calculated for ordinal or numerical data.

    For our example, the median is -1. Compute this with `median`:

    ```{r}
    median(x)
    ```

3.  **Mean** - the balancing point of the data, if a waiter was trying to
    balance the data on a tray. The mean can only be calculated for numerical
    data.

    For our example the mean is 0.4285. Compute this with `mean`:

    ```{r}
    mean(x)
    ```

Adding large values to the data affects the mean more than the median:

```{r}
y = c(x, 100)
mean(y)
median(y)
```

Because of this, we say that the median is **robust**.

The mean is good for getting a general idea of where the center of the data
is, while comparing it with the median reveals whether there are any
unusually large or small values.


### Scale Statistics

Scale is generally summarized by a number that says how far the data is from
the center (mean, median, etc...). Two options are:

1.  **Standard Deviation** - square root of the average squared distance to the
    mean (the distance from a point to a mean is called a **deviation**). You
    can think of this as approximately the average distance from a data point
    to the mean. As a rule of thumb, most of the data will be within 3 standard
    deviations of the mean.

    You can compute the standard deviation with `sd`:

    ```{r}
    sd(x)
    ```

2.  **Interquartile Range** (IQR) - difference between the 75th and 25th
    percentile. The median is the 50th **percentile** of the data; it's at the
    middle of the sorted data. We can also consider other percentiles. For
    instance, the 25th percentile is the value one-quarter of the way through
    the sorted data.

    **Quantile** is another word for percentile. **Quartile** specifically
    refers to the 25th, 50th, and 75th percentiles because they separate the
    data into four parts (hence "quart-").

    You can compute quantiles with `quantile`, or compute the IQR directly with
    `IQR`:

    ```{r}
    quantile(x)

    # IQR
    IQR(x)
    ```

The IQR is more robust than the standard deviation.


### Other Stuff

Many of the functions for computing statistical summaries have a parameter
`na.rm` to ignore missing values. Setting `na.rm = TRUE` is often useful when
you're just trying to do an initial investigation of the data. However, in a
more complete analysis, you should think carefully about what the missing
values mean, whether they follow any patterns, and whether there are enough
non-missing values for statistical summaries to be good representatives of the
data.

Finally, the `summary` function computes a detailed statistical summary of an R
object. For data frames, the function computes a summary of each column,
guessing an appropriate statistic based on the column's data type.
