```{r, echo = FALSE}
terns = read.csv("data/2000-2023_ca_least_tern.csv")
```

# Statistical Summaries {#sec-data-forensics}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the location statistics mean, median, and mode
* Explain the scale statistics range and standard deviation
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

After investigating (and possibly transforming) a dataset's structure, the next
step is to examine the statistical properties of its features. This chapter
focuses on individual features, and presents a variety of statistical
visualizations and estimates to summarize them, whether they're categorical or
numerical. These methods will help us identify and communicate trends,
patterns, anomalies, and limitations. For each method, we discuss how to
interpret the summary and explain the method's strengths and weaknesses
relative to others. The unifying theme of the chapter is summarizing, making
sense of, and communicating how data are distributed within populations and
samples.


## Categorical Features

To introduce ideas gently, we'll start by focusing on summaries for categorical
features. Most of the time, categories are not numbers, so we won't use much
math, and when we do, it will mostly be counting.

As motivation and a source of examples, we'll work to understand which regions
of California and which time periods are recorded in the California least terns
dataset (@sec-ca-least-terns). We'll also examine whether regions and time
periods are equally represented in the dataset.

To refresh our memory, the structure of the California least terns dataset is:

```{r}
str(terns)
```

We can study regional coverage at a coarse scale by looking at the `region_3`
or `region_4` columns. Let's focus on `region_3`. According to the
documentation, it's categorical, with categories for three regions. The first
few values are:

```{r}
head(terns$region_3)
```

The documentation claims the three regions are S.F. Bay, Central, and Southern,
but we can already see that this isn't accurate. There's another region: Kings.

The Kings region likely corresponds to sites in [Kings County][kings]. We can
try to confirm this by examining the site names in the region (with R's
[dplyr][] package):

[kings]: https://en.wikipedia.org/wiki/Kings_County,_California
[dplyr]: https://dplyr.tidyverse.org/

```{r}
#| warning: false
library("dplyr")

kings = filter(terns, region_3 == "KINGS")
unique(kings$site_name)
```

[Kettleman City][kc] is indeed a place in Kings County, so it seems like our
guess was correct.

[kc]: https://en.wikipedia.org/wiki/Kettleman_City,_California

:::{.callout-tip}
As you explore data, make a habit of following up on patterns and details that
seem confusing, surprising, or unusual. In some cases, as with the `region_3`
column, it's possible to use other parts of a dataset to confirm a guess or
make sense of what's happening. When it's not, turn to first-hand or expert
knowledge, such as the people who collected the dataset or related literature
and datasets.

For unusual details that seem tangential to the primary analysis goals, make a
note and revisit them later---they could turn out to be more relevant than you
expect.
:::


### Categories Are Constructs

When working with categorical features, keep in mind that categories are
created by people. We often use categories to simplify and organize
measurements that would otherwise be nuanced and messy. This process is
double-sided: it can be clarifying, making patterns easier to see, but it can
also be reductive, obscuring meaningful details. Poorly chosen categories can
lead us into overlooking or misattributing effects.

At a minimum, for each categorical feature, ask:

* What does the feature measure?
* Why is the feature categorical?
* Who defined the categories?
* Are the categories clearly defined?
* Are there any cases that are not covered by the defined categories? In other
  words, can you think of a case that would be difficult or impossible to
  categorize?
* Are there any cases that might be of interest on their own but are placed in
  a broader category?

The key is to approach categorical features skeptically. Think carefully about
how the categories represent what they claim to measure, and try to identify
which details they gloss over.


### Frequencies

To summarize a categorical feature, we can **count** or compute **frequencies**
of the categories. Taking this a step further, we can visualize the counts with
a bar plot so that categories with low or high counts stand out.

Let's make a bar plot of the `region_3` column. In R, the [ggplot2][] package's
`geom_bar` function, which makes a bar plot, counts categories automatically:

[ggplot2]: https://ggplot2.tidyverse.org/

```{r}
#| label: fig-region_3_bar
library("ggplot2")

ggplot(terns) +
  aes(x = region_3) +
  geom_bar()
```

The plot reveals two especially interesting details about `region_3`.

First, there are actually three unexpected regions: Kings, Sacramento, and
Arizona. Each has 15 observations or less and is an order of magnitude smaller
than the S.F. Bay, Central, and Southern regions.

The Sacramento region likely corresponds to [Sacramento county][sac] and
surrounding areas. Taking the same approach as we did for the Kings region, see
if you can confirm this.

[sac]: https://en.wikipedia.org/wiki/Sacramento_County,_California

<!--
```{r}
sac = filter(terns, region_3 == "SACRAMENTO")
unique(sac$site_name)
```
-->

The Arizona region is a bit of a mystery---there's no Arizona county in
California, nor are there any cities or places with Arizona in the name. Let's
see if the site names give any clues:

```{r}
arizona = filter(terns, region_3 == "ARIZONA")
unique(arizona$site_name)
```

Glendale is [a city in California][glendale-ca], but also [a city in
Arizona][glendale-az]. In this case, it seems likely that the Arizona region
corresponds to the state, and that perhaps these observations were not supposed
to be included in the dataset.

[glendale-ca]: https://en.wikipedia.org/wiki/Glendale,_California
[glendale-az]: https://en.wikipedia.org/wiki/Glendale,_Arizona

Second, the Southern region occurs 4 or 5 times more than any of the others.

The disparity in counts between the three expected regions could be due to
nesting preferences of the birds: perhaps there are more nesting sites in
southern California. On the other hand, it could also be due to how the dataset
is collected: perhaps the regions have different coverage over time, or CDFW
uses geographically smaller sites in southern California.

:::{.callout-tip}
Thinking through the different ways a dataset might have been generated like
this can help you identify subtle details that change or temper interpretations
and lead to further lines of inquiry.
:::


### Distributions {#sec-distributions}

Frequencies show us how the categories of a feature are distributed across the
observations. Because of this, we say they show us the **distribution** of the
feature. For example, @fig-region_3_bar shows us the distribution of
`region_3`.

Visualizing, estimating, summarizing, and interpreting distributions is a major
part of any data explorer or data scientist's job. The distribution in a sample
can give us insight into the distribution in a population, especially if the
sample was taken randomly.

We'll revisit the concept of distributions in @sec-numerical-features.


### The Mode

For a categorical feature, we call the category that occurs most frequently the
**mode**. In the case of a tie between two or more categories, we say the
feature has multiple modes. In a bar plot, the mode corresponds to the category
with the tallest bar. For the `region_3` column, the mode is the Southern
category.

Think of the mode of a feature as the category that's most likely if we were to
randomly select an observation from the dataset. If the dataset is a random
sample from a population, we can also think of the sample mode (computed with
the dataset) as an estimate of the population's mode.

:::{.callout-warning}
There's no built-in function to calculate the statistical mode in R. The `mode`
function is unrelated and returns something different.

Use a bar plot (or as we'll soon see, the `table` function) to identify modes.
:::


### Tables Versus Bar Plots

A table is another way to display category frequencies. In R, we can use the
`table` function to count categories and return the counts in a table. For
example, the counts for `region_3` are:

```{r}
table(terns$region_3)
```

Even small tables can be easy to misread, especially if the names of the
categories are long or include numbers. Large tables make the problem worse.
Sorting by frequency partially addresses the problem:

```{r}
sort(table(terns$region_3))
```

This makes it easy to see which categories are relatively small or large, but
the categories are no longer in alphabetical order, so it's harder to find
specific categories. For ordinal data, where there's a natural order to the
categories, sorting by frequency can break that order and make tables
confusing.

We recommend bar plots rather than tables in most situations, because they work
well for comparing category frequencies even when there are a lot of
categories. As with a table, we can sort the categories by frequency in a bar
plot to make the extremes more pronounced, but these are usually evident even
without sorting.

Bar plots have two disadvantages compared to tables. First, the exact count for
a category can be difficult to read. That's certainly the case in
@fig-region_3_bar. A simple way to address this is to display the count on the
plot:

```{r}
ggplot(terns) +
  aes(x = region_3) +
  geom_bar() +
  geom_text(
    # Use counts for the labels (geom_bar also uses stat_count).
    stat = "count", aes(label = after_stat(count)),
    # Set vertical justification of the labels, so they sit atop the bars.
    vjust = -0.5
  )
```

Second, bar plots are often less legible than tables to people with low vision.
One way to get the best of both bar plots and tables is to simply include both.


### Numbers as Categories

When a discrete feature only takes a few values and those values are repeated
across many observations, we can treat it like a categorical feature. The
categories just happen to be numbers.

In the California least terns dataset, the `year` column is a good example of
this. According to the documentation, the dataset only covers about 20 years.
There are multiple observations each year, because there are many different
sites. Let's make a bar plot of frequencies for `year`:

```{r}
#| label: fig-year_bar
ggplot(terns) +
  aes(x = year) +
  geom_bar()
```

The bar plot shows that the dataset, which spans years 2000-2023, doesn't
include any observations for the years 2001-2003. It also shows that there are
roughly 30-40 observations for every included year. That might mean most sites
were consistently measured over the entire time period, or it might be a
coincidence due to mostly equal numbers of sites being added and removed over
time. The mode is 2017, and There's a distinct drop in observations in 2018,
which could indicate a major change in which sites were being monitored.

The counterpoint is a discrete feature with many values that are rarely
repeated. A good example of this is is the `total_nests` column, which is an
estimate of the total number of nests for each site-year pair. Here's a bar
plot of frequencies for `total_nests`:

```{r}
#| label: fig-total_nests_bar
ggplot(terns) +
  aes(x = total_nests) +
  geom_bar()
```

We can see that a few values close to 0 are repeated frequently, but most of
the other values are rare. Because of this and because there are so many
different values, reading the bar plot is difficult. The situation is even
worse with a table:

```{r}
table(terns$total_nests)
```

To make sense of a feature like `total_nests`, we'll need an approach
well-suited to numerical features. We'll learn several in
@sec-numerical-features.

When a feature has a large number of possible values, tables and bar plots
can become difficult to interpret. If the feature is ordered (whether ordinal
or numerical), we're sometimes more interested in general trends in frequency
as we move through the values in order, rather than frequencies at specific
values.

We can get a sense of general trends by **binning** the values, which means
treating values close together as a single value. Imagine setting out a small
number of bins and dropping a few consecutive values into each. We then compute
frequencies (or other statistics) for the bins rather than the original values.
Binning tends to smooth out extremes, because it combines multiple values (some
of which will be less extreme).

A **histogram** is a bar plot where the values on the x-axis are binned. We can
use R to bin data and make a histogram automatically, although we recommend
(and the ggplot2 documentation recommends) setting the bin width manually for
numerical data. For instance, here's how to make a histogram for the `year`
column with 3-year bins:

```{r}
ggplot(terns) +
  aes(x = year) +
  geom_histogram(binwidth = 3, color = "black")
```

In the histogram, notice that the trends from the bar plot are smoothed out.
Key details still remain, like the low frequency for 2000-2003 and the drop in
frequency in 2018. For intuition, think of the original bar plot as having a
bin width of 1; wider bins produce more smoothing.

We'll revisit histograms for continuous data in @sec-numerical-data.

:::{.callout-tip}
When you make histograms, try several different bin widths. Sometimes one bin
width reveals details that are hidden in another. Choose a bin width that
preserves the most relevant details to your questions about the data.

If different bin widths lead to different interpretations of the data, include
both in your reporting and address why they do.
:::

:::{.callout-note}
R's functions for making histograms assume numerical data, but we could make a
histogram for ordinal data in two steps, by first binning and then making a bar
plot.
:::


## Numerical Data {#sec-numerical-data}

What about numerical data?

Two important questions to ask about data are:

1. Where is it? This is the **location** of the data.

2. How spread out is it? This is the **scale** of the data.

Let's use the data

```{r}
x = c(-2, -1, -1, -1, 0, 2, 6)
```

as an example.

Location is generally summarized with a number near the middle or center of
the data. A few options are:

1.  **Mode** - the value that appears most frequently. The mode can be
    calculated for any kind of data, but doesn't work well for continuous data.

    For our example, the mode of `x` is -1. You can compute the mode with
    `table`:

    ```{r}
    table(x)
    ```

2.  **Median** - sort the data, then find the value in the middle. The median
    can be calculated for ordinal or numerical data.

    For our example, the median is -1. Compute this with `median`:

    ```{r}
    median(x)
    ```

3.  **Mean** - the balancing point of the data, if a waiter was trying to
    balance the data on a tray. The mean can only be calculated for numerical
    data.

    For our example the mean is 0.4285. Compute this with `mean`:

    ```{r}
    mean(x)
    ```

Adding large values to the data affects the mean more than the median:

```{r}
y = c(x, 100)
mean(y)
median(y)
```

Because of this, we say that the median is **robust**.

The mean is good for getting a general idea of where the center of the data
is, while comparing it with the median reveals whether there are any
unusually large or small values.

Scale is generally summarized by a number that says how far the data is from
the center (mean, median, etc...). Two options are:

1.  **Standard Deviation** - square root of the average squared distance to the
    mean (the distance from a point to a mean is called a **deviation**). You
    can think of this as approximately the average distance from a data point
    to the mean. As a rule of thumb, most of the data will be within 3 standard
    deviations of the mean.

    You can compute the standard deviation with `sd`:

    ```{r}
    sd(x)
    ```

2.  **Interquartile Range** (IQR) - difference between the 75th and 25th
    percentile. The median is the 50th **percentile** of the data; it's at the
    middle of the sorted data. We can also consider other percentiles. For
    instance, the 25th percentile is the value one-quarter of the way through
    the sorted data.

    **Quantile** is another word for percentile. **Quartile** specifically
    refers to the 25th, 50th, and 75th percentiles because they separate the
    data into four parts (hence "quart-").

    You can compute quantiles with `quantile`, or compute the IQR directly with
    `IQR`:

    ```{r}
    quantile(x)

    # IQR
    IQR(x)
    ```

The IQR is more robust than the standard deviation.

Many of the functions for computing statistical summaries have a parameter
`na.rm` to ignore missing values. Setting `na.rm = TRUE` is often useful when
you're just trying to do an initial investigation of the data. However, in a
more complete analysis, you should think carefully about what the missing
values mean, whether they follow any patterns, and whether there are enough
non-missing values for statistical summaries to be good representatives of the
data.

Finally, the `summary` function computes a detailed statistical summary of an R
object. For data frames, the function computes a summary of each column,
guessing an appropriate statistic based on the column's data type.
