# Hypothesis Tests

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* TODO
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2
* palmerpenguins
* purrr

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

Statistics help us make informed decisions about the world. Sometimes the goal
is to evaluate whether a particular belief or **hypothesis** is true. We can
weigh the evidence in a dataset for (or against) a hypothesis with a
statistical procedure called a **hypothesis test**. Hypothesis tests take many
forms and are often tailored to specific problems. Nonetheless, all hypothesis
tests are based on a general framework. This chapter introduces the hypothesis
testing framework as well as a specific kind of hypothesis test that's broadly
applicable because of its minimal assumptions.


## Forming Hypotheses

The first step in hypothesis testing is forming a suitable hypothesis. The best
time to come up with a hypothesis is before you collect the dataset you'll use
to evaluate it. If the dataset has already been collected, the best time is
*before* doing any exploration. This way the dataset serves as a fair,
independent sample for testing the hypothesis.

:::{.callout-warning}
When you form a hypothesis based on patterns you see in a dataset, no matter
how slight, using the same dataset to test the hypothesis will overstate the
strength of the evidence.

To test a hypothesis formed by exploring a dataset, you must collect an
entirely new dataset from the same population.
:::

Ideally, your hypothesis is based on a clear conceptual model rooted in
personal experience and expert knowledge about the topic. This knowledge may
include prior work with *other* datasets. A good way to check is to write an
explanation of your hypothesis and why you believe it might be true. Ask
yourself whether a peer would find your argument convincing enough to warrant
further investigation, and if not, adjust the hypothesis.

:::{.callout-important}
We form hypotheses about a population and use hypothesis tests to test them on
samples. If you can collect data for the entire population, there's no need to
conduct a hypothesis test---the sample statistics are the population
statistics!
:::

In order to test a hypothesis with data, there are two requirements. First, the
hypothesis must relate to something that can be measured. If you can't measure
it, you can't test it (although in some cases you can choose and test a proxy).
If you already have a dataset, make sure that its features allow asking the
kinds of questions you want to ask.

Second, the hypothesis must be specific and falsifiable. Data and statistics
can't help with a hypothesis that's too broad or impossible to disprove. Try to
translate your hypothesis into a statement about a population parameter.

As an example, suppose you study trout in California rivers. You suspect that
the trout population in one river increased over the past year. You're not able
to conduct a complete census of the trout population, so you collect counts at
specific locations and times. These counts are the feature you can use to
address the hypothesis. There are several ways to make the hypothesis specific
enough to test; a simple one is to ask whether the mean counts in the current
year (averaging over locations) are higher than the mean counts in the previous
year.

<!--
As an example, suppose you've developed a new medicine to lower blood pressure.
Naturally, your hypothesis is that people who take the medicine will have lower
blood pressure than people who don't, all other things being equal. When you
collect a dataset, you must measure blood pressure in order to be able to
conclude anything about the hypothesis. You can make the hypothesis more
specific by comparing the mean blood pressure of the two groups.
-->


## The Framework {#sec-framework}

Every hypothesis test consists of the following steps:

1. Describe the hypothesis in terms of two competing claims. The first, called
   the **null hypothesis**, should be a reasonable default or baseline. The
   second, called the **alternative hypothesis** should be the hypothesis you
   believe might be true and want to test. The null and alternative hypotheses
   must be mutually exclusive, meaning only one can possibly be true.

2. Identify a statistic related to the hypothesis with a known sampling
   distribution. Many hypothesis tests use the sample mean, since its
   distribution is well-understood (@sec-central-limit-theorem). Compute the
   statistic on the sample.

3. Assume the null hypothesis is true. Use the sampling distribution to compute
   the probability of getting a statistic at least as unlikely as the sample
   statistic. For most hypothesis tests, we interpret "unlikely" as extreme
   (farther out in the distribution's tails). This probability, called a
   **p-value**, is a measure of evidence against the null hypothesis.

4. If the p-value is very small, the sample statistic is unlikely under the
   null hypothesis. In this case, we say that the result is **statistically
   discernable** and reject the null hypothesis. Otherwise, the sample
   statistic is compatible with the null hypothesis, so we do not reject it.

In step 4, the convention is to choose a specific threshhold for "very small,"
called the **discernibility level**. This is usually denoted by the Greek
letter $\alpha$ (alpha). The discernibility level is the approximate
probability of rejecting the null hypothesis when it's true (a false positive).
Most statistical software defaults to $\alpha = 0.05$ (5%), but you should
think carefully about what an acceptable false positive rate is for your work.

:::{.callout-important}
This framework only allows us to test whether the null hypothesis is unlikely!

A small p-value does not guarantee that the alternative hypothesis is correct
if multiple alternatives are possible. Along the same lines, a large p-value
does not guarantee that the null hypothesis is correct, nor that the
alternative hypothesis is incorrect.
:::

:::{.callout-note title="Note: Statistical Significance"}
"Statistically significant" is another term for for statistically discernable.
We prefer the latter, because somtimes a result is statistically discernable
but not important or meaningful in any real-world sense. Calling such a result
"significant" can easily give the wrong impression.
:::

:::{.callout-note title="Note: Power"}
After learning about the discernibility level, you might wonder: what about the
probability of failing to reject the null hypothesis when it's false (a false
negative)?

Statisticians usually denote this probability by the Greek letter $\beta$
(beta). Its inverse, $1 - \beta$, is called the **power** of the hypothesis
test. The power is the approximate probability that the test (correctly)
rejects the null hypothesis when it's false. Hypothesis tests are often
designed to maximize the power $1 - \beta$ for a given discernibility level
$\alpha$.

Power depends on the sampling distribution(s) under the alternative hypothesis.
It's usually impossible to compute the exact power of a test, because the
alternative hypothesis depends on unknown population parameters rather than
specifying them. That said, for many tests, we can estimate power by choosing
reasonably likely parameter estimates (for example, based on other data).

Power also depends on sample size. If you plan to collect a dataset and carry
out a hypothesis test, you should take power estimates into account when
selecting the sample size. An experiment that uses an underpowered hypothesis
test is a waste of resources.
:::


## Dataset: Palmer Penguins

![
  Artwork by [Allison Horst][horst].
](/images/applied-statistics/lter_penguins.png){#fig-penguins}

[horst]: https://allisonhorst.com/

As a source of examples, we'll use the [Palmer Penguins dataset][penguins],
which was collected by [Dr. Kristen Gorman][gorman] at [Palmer Station,
Antarctica][palmer]. [Alison Horst][horst] packaged the data for public use.
The package is called palmerpenguins.

[penguins]: https://allisonhorst.github.io/palmerpenguins/
[gorman]: https://www.uaf.edu/cfos/people/faculty/detail/kristen-gorman.php
[palmer]: https://pallter.marine.rutgers.edu/

:::{.callout-important}
Install the palmerpenguins package now if you didn't earlier.

Load the package with the `library` function to access the dataset, which is
called `penguins`.
:::

:::{.callout-warning}
In recent versions of R, the `penguins` dataset is built-in, but has slightly
different column names. If you want to follow along with this reader, make sure
to use the dataset from the package rather than the built-in dataset.
:::

The dataset consists of observations of 344 penguins from 3 different species:
[Adélie][], [Chinstrap][], and [Gentoo][]. The measurements include flipper
length, bill width, sex, and body mass.

[Adélie]: https://en.wikipedia.org/wiki/Ad%C3%A9lie_penguin
[Chinstrap]: https://en.wikipedia.org/wiki/Chinstrap_penguin
[Gentoo]: https://en.wikipedia.org/wiki/Gentoo_penguin


## Permutation Tests

We can use the Palmer Penguins dataset to explore how hypothesis tests work in
practice. Suppose we hypothesize that Adélie and Chinstrap penguins have
different mean body masses. Let $\mu_a$ denote the population mean body mass
for Adélie penguins and $\mu_c$ denote the same for Chinstrap penguins. Then we
can formalize our hypothesis in two parts:

* A null hypothesis $H_0: \mu_a = \mu_c$ that the means are equal. Another way
  to write this is $H_0: \mu_a - \mu_c = 0$, that the difference in means is 0.

* An alternative hypothesis $H_A: \mu_a \neq \mu_c$ that the means are not
  equal.

With the dataset, we can estimate $\mu_a - \mu_c$ with the difference of sample
means $\bar{x}_a - \bar{x}_c$. In R, we'll first set up a helper function to
compute the sample mean of each group and then compute their difference:

```{r}
#| warning: false
library("dplyr")

# This function computes the difference in sample means for Adelie and
# Chinstrap penguins.
difference_of_means_ac = function(data) {
  adelie = filter(data, species == "Adelie")
  chinstrap = filter(data, species == "Chinstrap")

  mean_a = mean(adelie$body_mass_g, na.rm = TRUE)
  mean_c = mean(chinstrap$body_mass_g, na.rm = TRUE)

  mean_a - mean_c
}
```

We can use the function to estimate the difference in mean body mass for Adélie
and Chinstrip penguins:

```{r}
#| warning: false
library("palmerpenguins")

penguins_ac = filter(penguins, species %in% c("Adelie", "Chinstrap"))
diff_ac = difference_of_means_ac(penguins_ac)
diff_ac
```

The difference is not zero, but that could be due to the sample rather than a
property of the entire population. But let's assume the null hypothesis is
true, so the two species have the same mean body mass. Is it likely that a
sample would have a `r round(diff_ac, 3)` difference? To address this, we need
to know the sampling distribution of the difference.

Under the null hypothesis, the two species are indistinguishable in terms of
mean body mass. So if we resample by permuting the species, the difference in
sample means should still be close to 0. By computing many resamples, we can
approximate the sampling distribution of the difference when the null
hypothesis is true. Then we can determine whether the difference for the
original sample is likely. This procedure is a kind of hypothesis test called a
**permutation test** or **randomization test**.

:::{.callout-note title="Algorithm: Permutation Test"}
Suppose we want to test a hypothesis about the relationship between two
features, and we can quantify the hypothesis with a statistic. To test the
hypothesis with a permutation test:

1. Construct a new sample by permuting one of the two features. Permuting means
   shuffling or sampling *without* replacement.
2. Compute the statistic on the new sample.
3. Repeat steps 1-2 to compute a collection of $k$ statistics.
4. Use the collected statistics to estimate the sampling distribution.
5. Use the sampling distribution to estimate a p-value for the statistic
   computed on the original sample.

Depending on the size of the original sample, it's sometimes possible to
compute all permutations (without any random sampling). Then we say the test is
**exact**.
:::

Let's work through a permutation test for the Adélie and Chinstrap penguins. In
R, we can use the `sample` function to permute a feature. We'll also use the
purrr package's `map` function to repeat the permutation step:

```{r}
library("purrr")

set.seed(12099)

# Resample `species` by shuffling (no replacement).
perm_resamples = map(seq_len(1000), \(i) {
  samp = penguins_ac
  samp$species = sample(samp$species)

  samp
})
```

Next, we need to compute the difference of sample means on each new sample. For
instance, on the first new sample:

```{r}
difference_of_means_ac(perm_resamples[[1]])
```

We can use the purrr package again to repeat this for all of the new samples:

```{r}
perm_diffs_ac = map_dbl(perm_resamples, difference_of_means_ac)
```

This collection of statistics enables us to estimate the sampling distribution.
We can make a density plot and mark where the original sample statistic lies:

```{r}
#| label: fig-permutation-density
#| fig-cap: >
#|     Density plot of the sampling distribution of the difference of means,
#|     assuming the null hypothesis that the means are equal is true. The
#|     dotted vertical line marks the value of the original sample statistic.
library("ggplot2")

ggplot(data.frame(difference_of_means = perm_diffs_ac)) +
  aes(x = difference_of_means) +
  geom_density() +
  geom_vline(xintercept = diff_ac, linetype = "dotted")
```

Based on @fig-permutation-density, the original sample statistic is fairly
close to the center of the distribution. In other words, it's a plausible
difference even if the null hypothesis is true.

To compute a p-value, we need to estimate the probability of getting a value at
least as extreme as the original sample statistic. We can use a **cumulative
distribution function**, a function that estimates the cumulative probability
of all values up to a given value, to do this. Since we don't know the
mathematical form of the sampling distribution, we'll use an **empirical
cumulative distribution function** (ECDF), an estimate based on the collection
of statistics we computed by permuting.

R's `ecdf` function computes an ECDF based on a sample. The ECDF is itself a
function, which we can use to estimate the p-value:

```{r}
ecdf(perm_diffs_ac)(diff_ac)
```

So under the null hypothesis, there's a roughly 30% chance of getting a more
extreme (more negative) difference than `r round(diff_ac, 3)`. At level $\alpha
= 0.05$, the difference in means is not statistically discernable. We fail to
reject the null hypothesis. It's still possible that Adélie and Chinstrap
penguins do have different mean body masses, but we would need to collect
additional data to prove it.

:::{.callout-warning}
The conclusion of a hypothesis test is not guaranteed to be correct, and other
sources of knowledge about the hypothesis should also be taken into account!
:::

:::{.callout-note title="Note: Classical Hypothesis Tests"}
Like the bootstrap algorithm, permutation tests only became possible in the age
of computing. Before that, statisticians developed a wide variety of hypothesis
tests based on mathematical approximations. Many of these rely on the CLT
(@sec-central-limit-theorem).

We don't cover classical hypothesis tests here (beyond the framework in
@sec-framework) because there are so many different kinds---they're usually
tailored to specific kinds of hypotheses and statistics.

Permutation tests are superior to classical hypothesis tests in most situations
because they make fewer assumptions. However, they do have a few disadvantages:

* They're only suitable for testing datasets where the observations are
  exchangeable under the null hypothesis. In other words, datasets where
  permuting won't break some other inherent structure. For instance,
  permuatation tests are usually not appropriate for testing time series.
* They take longer to compute. This is usually only a problem if the sample
  size is very large.
:::

:::{.callout-note title="See Also"}
For a more thorough explanation of hypothesis tests and permutation tests, as
well as many more examples, see [_Introduction to Modern Statistics_][ims]
[@ims]. This book also presents some classical hypothesis tests.

[ims]: https://openintrostat.github.io/ims/
:::
