```{r}
#| echo: false
#| warning: false
library("dplyr")
squid = readRDS("data/1989-1994_ca_market_squid.rds")
squid = filter_out(squid, mass_g >= 400 | is.na(mass_g))
```

# Hypothesis Tests {#sec-statistical-tests}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain the purpose of a hypothesis test
* Describe the characteristics of a testable hypothesis
* Construct a testable hypothesis for a research question
* Explain what null and alternative hypotheses are
* Explain what a p-value is
* Explain what it means for an effect to be statistically discernable
* Use a permutation test to test a hypothesis
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2
* purrr

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

Statistics help us make informed decisions about the world. Sometimes the goal
is to evaluate whether a particular belief or **hypothesis** is true. We can
weigh the evidence in a dataset for (or against) a hypothesis with a
statistical procedure called a **hypothesis test**. Hypothesis tests take many
forms and are often tailored to specific problems. Nonetheless, all hypothesis
tests are based on a general framework. This chapter introduces the hypothesis
testing framework as well as a specific kind of hypothesis test that's broadly
applicable because of its minimal assumptions.


## Forming Hypotheses

The first step in hypothesis testing is forming a suitable hypothesis. The best
time to come up with a hypothesis is before you collect the dataset you'll use
to evaluate it. If the dataset has already been collected, the best time is
*before* doing any exploration. This way the dataset serves as a fair,
independent sample for testing the hypothesis.

:::{.callout-warning}
When you form a hypothesis based on patterns you see in a dataset, no matter
how slight, using the same dataset to test the hypothesis will overstate the
strength of the evidence.

To test a hypothesis formed by exploring a dataset, you must collect an
entirely new dataset from the same population.
:::

Ideally, your hypothesis is based on a clear conceptual model rooted in
personal experience and expert knowledge about the topic. This knowledge may
include prior work with *other* datasets. A good way to check is to write an
explanation of your hypothesis and why you believe it might be true. Ask
yourself whether a peer would find your argument convincing enough to warrant
further investigation, and if not, adjust the hypothesis.

:::{.callout-important}
We form hypotheses about a population and use hypothesis tests to test them on
samples. If you can collect data for the entire population, there's no need to
conduct a hypothesis test---the sample statistics are the population
statistics!
:::

In order to test a hypothesis with data, there are two requirements. First, the
hypothesis must relate to something that can be measured. If you can't measure
it, you can't test it (although in some cases you can choose and test a proxy).
If you already have a dataset, make sure that its features allow asking the
kinds of questions you want to ask.

Second, the hypothesis must be specific and falsifiable. Data and statistics
can't help with a hypothesis that's too broad or impossible to disprove. Try to
translate your hypothesis into a statement about a population parameter.

As an example, suppose you study trout in California rivers. You suspect that
the trout population in one river increased over the past year. You're not able
to conduct a complete census of the trout population, so you collect counts at
specific locations and times. These counts are the feature you can use to
address the hypothesis. There are several ways to make the hypothesis specific
enough to test; a simple one is to ask whether the mean counts in the current
year (averaging over locations) are higher than the mean counts in the previous
year.

<!--
As an example, suppose you've developed a new medicine to lower blood pressure.
Naturally, your hypothesis is that people who take the medicine will have lower
blood pressure than people who don't, all other things being equal. When you
collect a dataset, you must measure blood pressure in order to be able to
conclude anything about the hypothesis. You can make the hypothesis more
specific by comparing the mean blood pressure of the two groups.
-->


## The Framework {#sec-framework}

Every hypothesis test consists of the following steps:

1. Describe the hypothesis in terms of two competing claims. The first, called
   the **null hypothesis**, should be a reasonable default or baseline. The
   second, called the **alternative hypothesis** should be the hypothesis you
   believe might be true and want to test. The null and alternative hypotheses
   must be mutually exclusive, meaning only one can possibly be true.

2. Identify a **test statistic** related to the hypothesis with a known
   sampling distribution. Many hypothesis tests use the sample mean, since its
   distribution is well-understood (@sec-central-limit-theorem). Compute the
   statistic on the sample.

3. Assume the null hypothesis is true. Use the sampling distribution to compute
   the probability of getting a statistic at least as unlikely as the sample
   statistic. For most hypothesis tests, we interpret "unlikely" as extreme
   (farther out in the distribution's tails). This probability, called a
   **p-value**, is a measure of evidence against the null hypothesis.

4. If the p-value is very small, the sample statistic is unlikely under the
   null hypothesis. In this case, we say that the result is **statistically
   discernable** and reject the null hypothesis. Otherwise, the sample
   statistic is compatible with the null hypothesis, so we do not reject it.

In step 4, the convention is to choose a specific threshhold for "very small,"
called the **discernibility level**. This is usually denoted by the Greek
letter $\alpha$ (alpha). The discernibility level is the approximate
probability of rejecting the null hypothesis when it's true (a false positive).
Most statistical software defaults to $\alpha = 0.05$ (5%), but you should
think carefully about what an acceptable false positive rate is for your work.

:::{.callout-important}
This framework only allows us to test whether the null hypothesis is unlikely!

A small p-value does not guarantee that the alternative hypothesis is correct
if multiple alternatives are possible. Along the same lines, a large p-value
does not guarantee that the null hypothesis is correct, nor that the
alternative hypothesis is incorrect.
:::

:::{.callout-note title="Note: Statistical Significance"}
"Statistically significant" is another term for for statistically discernable.
We prefer the latter, because somtimes a result is statistically discernable
but not important or meaningful in any real-world sense. Calling such a result
"significant" can easily give the wrong impression.
:::

:::{.callout-note title="Note: Power"}
After learning about the discernibility level, you might wonder: what about the
probability of failing to reject the null hypothesis when it's false (a false
negative)?

Statisticians usually denote this probability by the Greek letter $\beta$
(beta). Its inverse, $1 - \beta$, is called the **power** of the hypothesis
test. The power is the approximate probability that the test (correctly)
rejects the null hypothesis when it's false. Hypothesis tests are often
designed to maximize the power $1 - \beta$ for a given discernibility level
$\alpha$.

Power depends on the sampling distribution(s) under the alternative hypothesis.
It's usually impossible to compute the exact power of a test, because the
alternative hypothesis depends on unknown population parameters rather than
specifying them. That said, for many tests, we can estimate power by choosing
reasonably likely parameter estimates (for example, based on other data).

Power also depends on sample size. If you plan to collect a dataset and carry
out a hypothesis test, you should take power estimates into account when
selecting the sample size. An experiment that uses an underpowered hypothesis
test is a waste of resources.
:::


## Permutation Tests

We can use the California market squid dataset (@sec-dataset-ca-market-squid)
to explore how hypothesis tests work in practice. Recall the structure of the
dataset:

```{r}
str(squid)
```

Suppose we want to know whether male and female market squid differ in size. We
can use dorsal mantle length as a measure of size.

To formalize our hypothesis, we'll denote the mean dorsal mantle length for the
population of female market squid by $\mu_f$ and for the population of male
market squid by $\mu_m$. Then we can write our hypothesis in two parts:

* A null hypothesis $H_0: \mu_f = \mu_m$, that the means are equal.

* An alternative hypothesis $H_A: \mu_f \neq \mu_m$, that the means are not
  equal.

Another way to write the null hypothesis is $H_0: \mu_f - \mu_m = 0$, that the
difference in means is 0. With the dataset, we can estimate $\mu_f - \mu_m$
with the difference of sample means $\bar{x}_f - \bar{x}_m$.

In order to test this in R, first let's create a helper function that computes
the difference in mean dorsal mantle length between female and male squid for a
given sample:

```{r}
#| warning: false
library("dplyr")

difference_of_fm_lengths = function(data) {
  females = filter(data, sex == "Female")
  males = filter(data, sex == "Male")

  length_f = mean(females$dorsal_mantle_length_mm, na.rm = TRUE)
  length_m = mean(males$dorsal_mantle_length_mm, na.rm = TRUE)

  length_f - length_m
}
```

For the sample we have, the difference in means between females and males is:

```{r}
fm_length_diff = difference_of_fm_lengths(squid)
fm_length_diff
```

So female squid are about `r -round(fm_length_diff, 2)` millimeters shorter.
This difference is not zero, but that could be due to the sample rather than a
property of the entire population. If the null hypothesis is true---meaning
females and males have the same mean length---is it likely that a sample would
show a `r round(fm_length_diff, 2)` difference? To address this, we need to
think about the sampling distribution of the difference.

Under the null hypothesis, females and males are indistinguishable in terms of
length. So if we resample by permuting the sex, the difference in sample means
should still be close to 0. By computing many resamples, we can approximate the
sampling distribution of the difference when the null hypothesis is true. Then
we can determine whether the difference for the original sample is likely. This
procedure is a kind of hypothesis test called a **permutation test** or
**randomization test**.

:::{.callout-tip}
[_The Permutation Test: A Visual Explanation of Statistical Testing_][perm]
[@wilber2019] is an excellent online visual introduction permutation tests.

[perm]: https://www.jwilber.me/permutationtest/
:::

:::{.callout-note title="Algorithm: Permutation Test"}
Suppose we want to test a hypothesis about the relationship between two
features, and we can quantify the hypothesis with a statistic. To test the
hypothesis with a permutation test:

1. Construct a new sample by permuting one of the two features. Permuting means
   shuffling or sampling *without* replacement.
2. Compute the statistic on the new sample.
3. Repeat steps 1-2 to compute a collection of $k$ statistics.
4. Use the collected statistics to estimate the sampling distribution.
5. Use the sampling distribution to estimate a p-value for the statistic
   computed on the original sample.

Depending on the size of the original sample, it's sometimes possible to
compute all permutations (without any random sampling). Then we say the test is
**exact**.
:::

Let's work through a permutation test for the squid. We can use R's `sample`
function to permute a feature. We'll also use the purrr package's `map`
function to repeat the permutation step:

```{r}
library("purrr")

set.seed(12099)

# Resample by shuffling `sex` (no replacement).
perm_resamples = map(seq_len(10000), \(i) {
  samp = squid
  samp$sex = sample(samp$sex)

  samp
})
```

Next, we need to compute the difference in mean dorsal mantle length between
female and male squid on each new sample. For instance, on the first new
sample:

```{r}
difference_of_fm_lengths(perm_resamples[[1]])
```

We can use the purrr package again to repeat this for all of the new samples:

```{r}
perm_fm_length_diffs = map_dbl(perm_resamples, difference_of_fm_lengths)
```

This collection of statistics enables us to estimate the sampling distribution.
We can make a density plot and mark where the original sample statistic lies:

```{r}
#| label: fig-permutation-density
#| fig-cap: >
#|     Density plot of the sampling distribution of the difference of means,
#|     assuming the null hypothesis that the means are equal is true. The
#|     dotted vertical line marks the value of the original sample statistic.
library("ggplot2")

ggplot(data.frame(difference_of_means = perm_fm_length_diffs)) +
  aes(x = difference_of_means) +
  geom_density() +
  geom_rug() +
  geom_vline(xintercept = fm_length_diff, linetype = "dotted")
```

Based on @fig-permutation-density, the original sample statistic is an extreme
value! If the null hypothesis is true, it's very unlikely that the difference
`r round(fm_length_diff, 2)`. would occur by chance.

Let's complete the permutation test by computing a p-value and comparing it to
discernibility level $\alpha = 0.05$. To compute the p-value, we need to
estimate the probability of getting a value at least as extreme as the original
sample statistic. We can use a **cumulative distribution function**, a function
that estimates the cumulative probability of all values up to a given value, to
do this. Since we don't know the mathematical form of the sampling
distribution, we'll use an **empirical cumulative distribution function**
(ECDF), an estimate based on the collection of statistics we computed by
permuting.

R's `ecdf` function computes an ECDF based on a sample. The ECDF is itself a
function, which we can use to estimate the p-value:

```{r}
ecdf(perm_fm_length_diffs)(fm_length_diff)
```

So under the null hypothesis, there's almost 0% chance of getting a more
extreme (more negative) difference than `r round(fm_length_diff, 2)`. At level
$\alpha = 0.05$, the difference in means is statistically discernable and we
reject the null hypothesis. The sample provides strong evidence that female and
male squid have different mean dorsal mantle lengths.

:::{.callout-important title="Important: One- and Two-Sided Hypothesis Tests"}
Technically, our hypothesis test of difference in mean squid lengths is a
**two-sided test**: we'd reject the null hypothesis whether the difference very
negative or very positive. This is implied by the stated null hypothesis $H_0:
\mu_f = \mu_m$.

In a two-sided test with discernibility level $\alpha$, we check whether the
p-value is less than $\alpha / 2$, since we allocate half of the probability to
extreme negative values and half to extreme positive values.

We can write the test as a **one-sided test** instead if we only want to test
for a negative (or positive) difference. Then the null hypothesis is $H_0:
\mu_f \geq \mu_m$ and the alternative hypothesis is $H_A: \mu_f < \mu_m$ (to
test for a positive difference, flip the inequalities).

In a one-sided test with discernibility level $\alpha$, we simply check whether
the p-value is less than $\alpha$.

The distinction between one- and two-sided tests is subtle, but it matters when
reporting discernibility levels. Make sure to think about the research question
and choose the most appropriate test *before* computing the p-value.
:::

:::{.callout-warning}
The conclusion of a hypothesis test is not guaranteed to be correct, and other
sources of knowledge about the hypothesis should also be taken into account!
:::


## Other Tests

Like the bootstrap algorithm, permutation tests only became possible in the age
of computing. Before that, statisticians developed a wide variety of hypothesis
tests based on mathematical approximations. Many of these rely on the CLT
(@sec-central-limit-theorem) or other statistical results about large-sample
behavior.

We don't cover classical hypothesis tests here (beyond the framework in
@sec-framework) because there are so many different kinds---they're usually
tailored to specific kinds of hypotheses and statistics---and because they're
covered in detail in most introductory statistics textbooks.

:::{.callout-tip title="Note: Hypothesis Tests as Models"}
Introductory statistics courses often teach and encourage students to memorize
a long list of hypothesis tests tailored to specific circumstances. Most of
these hypothesis tests can also be described in terms of linear models
(@sec-linear-models); different terms in the model lead to different tests.

The website ["Common statistical tests are linear models"][tests-are-models]
[@lindelov2019] provides an introduction and overview of this idea.

[tests-are-models]: https://lindeloev.github.io/tests-as-linear/
:::

Permutation tests are more versatile than classical hypothesis tests because
they make fewer assumptions. However, they do have a few disadvantages:

* They're designed to compare two or more samples, features, or categories. For
  a single feature in a single sample, other tests are more appropriate.
* They're only suitable for testing datasets where the observations are
  exchangeable under the null hypothesis. In other words, datasets where
  permuting won't break some other inherent structure. For instance,
  permuatation tests are usually not appropriate for testing time series.
* They take longer to compute. This is usually only a problem if the sample
  size is very large.

In addition to classical hypothesis tests, it's also possible to use the
bootstrap (@sec-bootstrap-algorithm} for hypothesis tests. The key idea is that
a $(1 - \alpha)\%$ confidence interval is equivalent to a hypothesis test with
discernibility level $\alpha$, regardless of how we construct the confidence
interval. If the parameter value specified by the null hypothesis is in the
confidence interval, then the test fails to reject the null hypothesis;
otherwise, the test rejects the null hypothesis. Constructing a test this way
is called **inverting** the confidence interval. Bootstrap hypothesis tests are
often useful in cases where we can't use a permutation test.

:::{.callout-note title="See Also"}
[_Introduction to Modern Statistics_][ims] [@ims] provides a more thorough
introduction to hypothesis tests (permutation, bootstrap, and classical tests)
with many examples.

[ims]: https://openintrostat.github.io/ims/

[_OpenIntro Statistics_][openintro] [@openintro] introduces a wide variety of
classical hypothesis tests (more than _Introduction to Modern Statistics_), but
doesn't cover permutation or bootstrap tests.

[openintro]: https://www.openintro.org/book/os/

In R, there are built-in functions to carry out classical hypothesis tests. The
[coin][] package provides functions to carry out permutation tests. The
[boot][] package provides functions to carry out bootstrap tests.

[coin]: https://doi.org/10.32614/CRAN.package.coin
[boot]: https://doi.org/10.32614/CRAN.package.boot
:::
