# Linear Models

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* TODO
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggfortify
* ggplot2
* palmerpenguins

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::


We often want to study features together to make sense of the underlying
relationship. 

A common application of statistics is to study relationships between features.
@sec-relationships-between-features introduced visualizations and the
correlation statistic for this purpose.
When the
relationship between two or more features roughly follows a straight line, we
can fit a **linear model** that estimates the line's intercept and slope.


## Fitting a Model

Let's examine the Palmer Penguins dataset (@sec-dataset-palmer-penguins) once
again. As a reminder, the structure is:

```{r}
#| warning: false
library("palmerpenguins")

str(penguins)
```

For now, we'll focus on the `flipper_length_mm` and `body_mass_g` features. It
makes sense that they might have some relationship---maybe that penguins with
higher body mass tend to have longer flippers. Let's use the dataset to try to
determine whether this is true. First, we'll filter out all of the rows where
one or both of the features are missing, since they can't help us understand
the relationship:

```{r}
#| warning: false
library("dplyr")

penguins_tidy = filter_out(
  penguins, is.na(flipper_length_mm), is.na(body_mass_g)
)
```

Following the advice of @sec-relationships-between-features, let's make a
scatter plot of the two features:

```{r}
#| warning: false
#| label: fig-flipper_length_mm-body_mass_g
library("ggplot2")

ggplot(penguins) +
  aes(x = flipper_length_mm, y = body_mass_g) +
  geom_point()
```

@fig-flipper_length_mm-body_mass_g shows that we were correct to think that
body mass and flipper length increase together. In fact, the points in this
sample almost seem to follow a straight line. The line can't pass through every
point because of indivdual variation among the penguins, but it can capture the
general trend or pattern in the points. Whatever line we draw is a **linear
model**: a line that approximates the relationship between features.

We can make the linear model concrete by writing the equation for the line. A
line always has an intercept, which we'll denote by $\beta_0$, and slope, which
we'll denote by $\beta_1$. Then we can write:

$$
(\textrm{body mass})
= \beta_0 + \beta_1(\textrm{flipper length})
$$

One way to interpret this equation is that the feature on the left-hand side is
fully determined by the right-hand side. In this interpretation, we call the
feature on the left-hand side the **response** and each feature on the
right-hand side a **predictor**. For our model, body mass is the response and
flipper length is a predictor.

:::{.callout-note title="Note: Alternative Terminology" collapse="true"}
We'll refer to the feature on the left-hand side as "the response" and each
feature on the right-hand side as a "predictor" or simply a "feature." However,
there are many other names for these, and the names are often
discipline-specific.

The feature on the left-hand side is sometimes called the **dependent** or
**endogenous** feature.

A feature on the right-hand side is sometimes called a **covariate**, or an
**independent** or **exogenous** feature.
:::

:::{.callout-warning}
Despite the predictor-response interpretation, a linear model alone can't tell
us whether a relationship between features is causal. To conclude that flipper
length causes body mass (or vice-versa), we'd have to make additional
assumptions.

For an introduction to causal inference, see "Part 5: Causal Inference" of
[_Regression and Other Stories_][ros] [@gelman2020].
:::

Due to indivdual variation among the penguins, the line can't pass through
every observation (and might not pass through any!). To account for this, we
can include a **residual** (or error) term in the model:

$$
(\textrm{body mass})
= \beta_0 + \beta_1(\textrm{flipper length}) + (\textrm{residual})
$$

The residual is a reminder that this is just a model.

Writing the linear model in terms of variables makes it more general (because
the variables can refer to any features from any dataset) and also more
convenient to work with mathematically. Let $(x_i, y_i)$ denote observations
(with two features) and $\epsilon_i$ denote residuals, for $i = 1, \dots, n$.
Then we can write the model as:

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
$$

This notation is the convention in statistics.

:::{.callout-note title="Note: Hat Notation"}
Another statsitics convention is to write a "hat" above variables that are
estimators or estimates (including predictions). So we can write the model's
prediction for $y_i$ as

$$
\hat{y}_i = \beta_0 + \beta_1 x_i
$$

We can also use this notation for parameter estimates. For instance,
$\hat{\mu}$ denotes an estimate for a distribution parameter $\mu$.
:::

There are many different ways we can chose the intercept $\beta_0$ and slope
$\beta_1$ for the line. The most widely-used approach, called **ordinary least
squares** is to choose them to minimize the **residual sum of squares**
$SS_\epsilon$, defined as

$$
\textrm{SS}_\epsilon
= \sum_{i=1}^n \epsilon_i^2
= \sum_{i=1}^n (y_i - \hat{y}_i)^2
= \sum_{i=1}^n \left[ y_i - (\beta_0 + \beta_1 x_i) \right]^2
$$

<!-- TODO: say more about this? MLE? -->


In R, we can use the `lm` function to fit a linear model by least squares. The
function's first argument must be a relationship between features, written with
a tilde `~` instead of an equals sign. The intercept is included implicitly,
but you can make it explicit by writing it as a `1` (imagine it as a
coefficient for a column of ones). In R's class system, this is called a
`formula`.

The second argument must be a dataset that contains the aforementioned
features.

So the code to fit the body mass linear model for the penguins dataset is:

```{r}
model = lm(body_mass_g ~ flipper_length_mm, penguins_tidy)
model
```

We can use the `summary` function to get more details about the model:

```{r}
summary(model)
```

Let's go through the model summary a few lines at a time. It begins with the R
code we used to fit the model:

```{.text}
Call:
lm(formula = body_mass_g ~ flipper_length_mm, data = penguins_tidy)
```

Next, it lists the quartiles of the residuals:

```{.text}
Residuals:
     Min       1Q   Median       3Q      Max
-1058.80  -259.27   -26.88   247.33  1288.69
```

The distribution of the residuals is an important source of diagnostic
information for linear models, but visualizations are almost always easier to
interpret than quartiles. @sec-model-diagnostics explains how to make and
interpret plots of the residuals, as well as other diagnostics.

After the residual quartiles, the summary lists the coefficients in a table:

```{.text}
Coefficients:
                   Estimate Std. Error t value Pr(>|t|)
(Intercept)       -5780.831    305.815  -18.90   <2e-16 ***
flipper_length_mm    49.686      1.518   32.72   <2e-16 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
```

Each row of the table corresponds to one term in the model. The first is
typically the intercept $\beta_0$. The remaining rows correspond to features.
The columns contain an estimate of the coefficient, an estimate of its standard
error, and two columns related to hypothesis testing. We'll examine these last
two columns in @sec-hypothesis-tests-for-models.

Finally, the summary lists several statistics about the model as a whole,
beginning with:

```{.text}
Residual standard error: 394.3 on 340 degrees of freedom
```

Residual standard error is the sample standard deviation of the residuals (with
a statistical correction for the number of terms being estimated in the model).
Recall that standard deviation is approximately the mean distance to the mean.
In a linear model with an intercept, the residuals always have mean 0
($\bar{\epsilon} = 0$), so the residual standard error is approximately the
mean distance between the actual responses and the fitted line.

:::{.callout-note title="Definition: Residual Standard Deviation" collapse="true"}
The sample variance of the residuals for a linear model with $k$ terms
(including the intercept) is defined as

$$
s_\epsilon^2
= \frac{\textrm{SS}_\epsilon}{n - k}
= \frac{1}{n - k} \sum_{i=1}^n (\epsilon_i - 0)^2
$$

Think of this as the variance in the sample that's not explained by the model.

The sample standard deviation of the residuals $s_\epsilon$ is the square root
of $s_\epsilon^2$.
:::

:::{.callout-note title="Note: The Name \"Residual Standard Error\"" collapse="true"}
The name "residual standard error" is an R invention. Standard error refers to
the standard deviation of an estimator across many samples, but there's no
estimator and no resampling at work here. The R documentation even notes that
this name is inappropriate (in `?sigma`).

Appropriate names include "standard deviation of the residuals" and "residual
standard deviation" among others.

See [this StackOverflow post][so-rse] for additional discussion.

[so-rse]: https://stats.stackexchange.com/questions/144433/why-do-we-say-residual-standard-error
:::

```{.text}
Multiple R-squared:  0.759, Adjusted R-squared:  0.7583
```

The R-squared statistics, conventionally denoted by $R^2$ and
$R_\textrm{adj}^2$, are ratios of variance in the response that's explained by
the model to total variance.

R-squared $R^2$ is always between 0 and 1. Values closer to 1 indicate that the
model explains more of the variance in the response. R-squared is
non-decreasing as more terms are included in the model, because more terms
means more flexibility to fit the observations. This is a limitation: we can't
use R-squared to decide how many terms to include in the model, because it will
always suggest more terms are better.

Adjusted R-squared $R_\textrm{adj}^2$ decreases when terms with weak
explanatory power are added to the model, so it tends to be more useful than
ordinary R-squared $R^2$. Its value can sometimes be smaller than 0.

:::{.callout-note title="Definition: R-Squared Statistics" collapse="true"}
The R-squared statistics are based on the idea that the variance of the
response $y_i$ can be partitioned into two meaningful parts: variance explained
by the model and unexplained variance from other, often unknown, processes.

Mathematically, variance is a sum of squares. The **total sum of squares** for
the response $y_i$ is defined as

$$
\textrm{SS}
= \sum_{i=1}^n (y_i - \bar{y})^2
= (n - 1) s^2
$$

Notice that the total sum of squares is part of and proportional to the sample
variance $s^2$ (@sec-standard-deviation).

We can partition the total sum of squares into two parts, the **explained sum
of squares** $\textrm{SS}_\textrm{model}$ and the residual sum of squares
$\textrm{SS}_\epsilon$:

$$
\textrm{SS}
= \textrm{SS}_\textrm{model} + \textrm{SS}_\epsilon
= \sum_{i=1}^n (\hat{y}_i - \bar{y})^2 + \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

Then R-squared $R^2$ is defined in terms of the sums of squares as

$$
R^2
= 1 - \frac{\textrm{SS}_\epsilon}{\textrm{SS}}
= \frac{\textrm{SS}_\textrm{model}}{\textrm{SS}}
$$

For a model with $k$ terms (excluding the intercept), the adjusted R-squared
$R_\textrm{adj}^2$ is defined as

$$
R_\textrm{adj}^2
= 1 - \frac{\textrm{SS}_\epsilon}{n - k - 1} \cdot \frac{n - 1}{\textrm{SS}}
= 1 - \frac{s_\epsilon^2}{s^2}
$$

The denominator $n - k - 1$ accounts for the number of terms in the model.
:::

R-squared, whether adjusted or not, mostly tells us about how well the model
fits *the sample*. A model that fits a sample well might not fit the population
well, and a model that fits the population well might not fit a sample well.
Moreover, for some populations, observations are always noisy (accurate
measurement is difficult). Then R-squared can be large even if the model fits
the population well.

:::{.callout-warning}
Treat R-squared as one statistic among many when examining a model.

To estimate how well a model will generalize, use other methods specifically
designed for that purpose, such as cross-validation.
:::


```{.text}
F-statistic:  1071 on 1 and 340 DF,  p-value: < 2.2e-16
```

The F statistic is the ratio of variance in the response that's explained by
the model to unexplained variance. The F statistic is always positive.

The F statistic is also the relative difference between the residual sum of
squares (unexplained variance) for a simplified model with only an intercept
(the same as predicting the response with the sample mean) and for the fitted
model. This perspective is helpful for interpreting the statistic: larger
values indicate that the fitted model is better at explaining the variance of
the response than the simplified model.

:::{.callout-note title="Definition: F Statistic" collapse="true"}
For a model with $k$ terms (excluding the intercept), the F statistic is
defined as

$$
F
= \frac{\textrm{SS}_\textrm{model}} {\textrm{SS}_\epsilon}
  \cdot \frac{n - k - 1} {k}
= \frac{\textrm{SS} - \textrm{SS}_\epsilon}{\textrm{SS}_\epsilon}
  \cdot \frac{n - k - 1}{k}
$$
:::

The F statistic is primarily used for a hypothesis test where the null
hypothesis is that the model coefficients (excluding the intercept) are all 0.
As @gelman2020 points out, this null hypothesis is unrealistic---if we believe
it's plausible that a feature is entirely unrelated to the response, we
shouldn't include it in the model in the first place. The F test is often used
to decide whether a model with fewer features might be more appropriate, but
there are better approaches, such as examining standard errors for the
coefficient estimates to decide whether the estimates are too imprecise to be
useful.

:::{.callout-warning}
As with R-squared, treat the F statistic as one statistic among many when
examining a model.
:::

:::{.callout-important}
Treat every model as a simplification of reality. Statistics can help us
understand how well a model describes data, but can't tell us whether a model
is "correct" (and it's unclear what it means for a simplification to be
"correct" anyway).
:::


## Model Assumptions {#sec-model-assumptions}

A linear model involves many assumptions about the features being modeled and
the dataset used to fit the model. @gelman2020 lists these assumptions (from
most to least important):

1. Validity: the response should measure what you actually want to study, the
   model should include all relevant features, and the model should generalize
   well. For instance, a model for blood pressure should not be treated as a
   model for general health unless this can be justified by expert knowledge.

2. Representativeness: the dataset you use to train the model should be
   representative of the population to which you want the model to generalize.
   For example, a model for how long people can hold their breath should not be
   based on a sample of Olympic swimmers. On the other hand, a representative
   sample might include a few Olympic swimmers, because they're people too!

3. Linearity: the relationship between the response and the features _in the
   model_ should be roughly linear. Often the relationship between the response
   and the features _in the dataset_ isn't linear, but it's possible to
   transform the features (or the response) to get a linear relationship.

   When the relationship is not linear, it's often possible
   to transform the response or the features to get a linear relationship.

4. Independence of errors: the residual for one observation should not depend
   on the residuals of others. This assumption means linear models are
   incompatible with most temporal and spatial datasets, although there are
   modifications we can make to accomodate these kinds of datasets.

5. Equal variance of errors: the residuals should all have the same standard
   deviation. When they don't, then the model is worse at predicting for some
   inputs than others. One way to address this is by using weighted least
   squares to fit the model rather than ordinary least squares.

6. Normality of errors: the residuals should be have a normal distribution.
   This is assumption really only necessary for hypothesis testing (with
   CLT-based methods); it doesn't matter for the coefficient estimates. This
   assumption is sometimes misinterpreted to mean that the response must have a
   normal distribution, which is not true.

Check these assumptions before and after fitting a linear model. The first two
assumptions are really about the research question and data collection process,
so failing to satisfy them usually can't be fixed by changing the model. On the
other hand, the last four assumptions relate to the model directly, and failing
to satisfy them can sometimes be fixed by changing the model.

@sec-model-diagnostics describes how to check some of these assumptions with
model diagnostics.

:::{.callout-tip}
Many kinds of non-linear relationships can be made linear with transformations.

For example, suppose that the true relationship between two features $x_i$ and
$y_i$ is quadratic:

$$
y_i = \beta_0 + \beta_1 x_i^2 + \epsilon_i
$$

From another perspective, the relationship between $y_i$ and $x_i^2$ is linear.
So if the dataset only includes $x_i$, square it to get $x_i^2$ and use this
new feature in the linear model.

Choosing appropriate transformations can be difficult in practice, because we
rarely know the true nature of the relationship. Fortunately, relationships
that a model fails to account for typically show up as patterns in its
residuals. We can use the residuals to decide if and how features should be
transformed (see @sec-model-diagnostics).
:::


## Model Diagnostics {#sec-model-diagnostics}

Most of the information about how well a model fits is in the residuals, since
the residuals measure what the model got wrong. We can make a variety of plots
based on the residuals to check whether the assumptions from
@sec-model-assumptions are satisfied and whether there are any other problems.

For a linear model with only one feature, a good diagnostic starting point is
to plot the feature against the residuals. This is called a **residual plot**.
Patterns in the residual plot can reveal problems with linearity, independence
of errors, and equal variance of errors. Let's make this plot for the penguins
dataset and model.

In R, the `predict` function returns predictions from a model. Absent any
additional arguments, it returns predictions for the dataset used to fit the
model. So to get the predictions for the penguins dataset:

```{r}
penguins_tidy$predicted_body_mass_g = predict(model)
```

We can use the predictions to calculate the residuals, in order to plot the
residuals against the flipper length feature:

```{r}
#| label: fig-residuals_flipper_length_mm
#| fig-cap: >
#|    Flipper length versus residuals for the penguin body mass linear model.
ggplot(penguins_tidy) +
  aes(x = flipper_length_mm, y = body_mass_g - predicted_body_mass_g) +
  geom_point()
```

:::{.callout-tip}
The `resid` function returns the residuals from a model. It's a convenient
alternative to computing the residuals from the predictions.
:::

For a dataset and model that satisfy the assumptions (@sec-model-assumptions),
the residuals should be centered on 0 (mean 0) and should not exhibit any
patterns. Two kinds of patterns are especially important:

* A trend in the residuals (that is, a line or curve that they follow). This
  usually means that the linearity assumption isn't satisfied.
* A "fan" or other shape in the range of the residuals. This usually means that
  the equal variance assumption isn't satisfied.

The residual plot will also exhibit patterns when the independence assumption
isn't satisfied, but many different patterns are possible depending on the
nature of the dependence.

For the penguins dataset and model, the residuals in
@fig-residuals_flipper_length_mm don't exhibit any strong patterns. The
residuals are centered at 0, don't have an obvious trend, and range from about
-1,000 to 1,000 across most of the feature values. So this linear model seems
to satisfy the assumptions well, although it's certainly possible there are
better models.

For models with multiple features, we can plot each feature against the
residuals. We can also plot the predictions against the residuals. These plots
are all residual plots, and the interpretation is the same as for a model with
a single feature. Let's plot the predictions against the residuals for the
penguins model:

```{r}
#| label: fig-residuals_predicted_penguin
#| fig-cap: >
#|    Predictions versus residuals for the penguin body mass linear model.
ggplot(penguins_tidy) +
  aes(x = predicted_body_mass_g, y = body_mass_g - predicted_body_mass_g) +
  geom_point()
```

Notice that @fig-residuals_flipper_length_mm and
@fig-residuals_predicted_penguin only differ in the values on the x-axis. This
is always the case for a model with a single feature. For models with multiple
features, the residual plots for individual features will not necessarily be
the same as the plot of predictions versus residuals.

Most statistical software, including R, provide three more diagnostic plots for
linear models:

* The **normal quantile-quantile plot** (normal QQ plot) displays the quantiles
  of a normal distribution against the quantiles of the residuals. QQ plots are
  helpful for checking how similar two distributions are, because similar
  distributions will have similar quantiles. For a linear model, we can use the
  normal QQ plot to check the normality assumption. The normal QQ plot is less
  important than other diagnostics, because the normality assumption is less
  important than the other assumptions.
* The **scale-location plot** displays the predictions against the square root
  of the standardized residuals. This is a transformation of the residual plot,
  so the interpretation is the same; the transformation simply makes some
  patterns easier to notice.
* The **leverage plot** displays a statistic called leverage against the
  residuals. Leverage is a measure of how far away each observation in the
  dataset is from the others. Observations with high leverage may be outliers
  and may have undue influence over the linear model (similar to how extreme
  values affect the sample mean).

In R, we can use the ggfortify package's `autoplot` function to make all four
diagnostic plots for a model:

```{r}
#| warning: false
#| label: fig-body_mass_g-diagnostics
#| fig-cap: >
#|    Diagnostic plots for the penguin body mass linear model. Clockwise from
#|    the top left, they are the residual plot, normal QQ plot, leverage plot,
#|    and scale-location plot.
library("ggfortify")

autoplot(model)
```

:::{.callout-note title="Note: R's Built-in Diagnostic Plots" collapse="true"}
Instead of using the ggfortify package, to make diagnostic plots for a model
with ggplot2, it's also possible to make diagnostic plots with R's built-in
`plot` function:

```{r}
plot(model)
```
:::

The diagnostic plots in @fig-body_mass_g-diagnostics suggest that overall, the
model fits the dataset relatively well, as we previously concluded from the
residual plot alone. This version of the residual plot includes a fitted trend
line which hints that there could be some trend, but it's mild. The normal QQ
plot is nearly perfect, with a close match between the normal quantiles and the
residual quantiles (the dashed line is the line of exact equality). The
scale-location plot leads to similar conclusions as the residual plot, with the
trend in residuals notably less pronounced. Finally, the leverage plot
indicates a few observations with middling leverage that and very large
residuals; these are annotated with row numbers on the plot. In a thorough
analysis, we should investigate each of these as potential outliers
(@sec-outliers).

:::{.callout-note title="See Also"}
For examples of patterns to look for in residual plots, see:

* The ["Checking model conditions"][ims-model-conditions] section in
  _Introduction to Modern Statistics_ [@ims]
* The "Residual plots" section in [_Regression and Other Stories_][ros]
  [@gelman2020]

[ims-model-conditions]: https://openintrostat.github.io/ims/inf-model-slr.html#sec-tech-cond-linmod
:::


## Selecting Features

A dataset might have dozens or hundreds of features, but including every
feature in a linear model is usually not a good idea. One problem is that each
additional feature in the model means an additional coefficient to estimate.
When the number of observations is low relative to the number of features in
the model, there usually isn't enough information to compute accurate
coefficient estimates.

Another problem is that when two features are highly correlated
(@sec-correlation), they effectively contain redundant information, which again
makes it difficult (or impossible) to compute accurate coefficient estimates.
This problem is called **multicollinearity**.

The point is that we 

```{r}
ggplot(penguins) +
  aes(x = flipper_length_mm, y = body_mass_g, color = species) +
  geom_point()
```

<!-- Explain categorical vs numerical features -->

```{r}
model_species = lm(body_mass_g ~ species + flipper_length_mm, penguins)
summary(model_species)
```

```{r}
autoplot(model_species)
```

```{r}
penguins_tidy$pred = predict(model_species)
penguins_tidy$resid = resid(model_species)

ggplot(penguins_tidy) +
  aes(x = pred, y = resid, color = species) +
  geom_point()
```


### Interactions

```{r}
model_interact = lm(body_mass_g ~ species * flipper_length_mm, penguins)
summary(model_interact)
```

```{r}
autoplot(model_interact)
```

:::{.callout-note title="See Also"}
[_Feature Engineering and Selection_][feature-engineering] [@kuhn2020]

[feature-engineering]: https://feat.engineering/
:::


## Interpreting Models


### Hypothesis Tests for Models {#sec-hypothesis-tests-for-models}

$$
Y_i \sim \mathcal{N}(\beta_0 + \beta_1 x_i, \sigma^2)
$$

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
\quad\textrm{where}\quad
\epsilon_i \sim \mathcal{N}(0, \sigma^2)
$$

:::{.callout-tip title="Note: Hypothesis Tests as Models"}
Introductory statistics courses often teach and encourage students to memorize
a long list of hypothesis tests tailored to specific circumstances. Most of
these hypothesis tests can also be described in terms of linear models;
different terms in the model lead to different tests.

The website ["Common statistical tests are linear models"][tests-are-models]
[@lindelov2019] provides an introduction and overview of this idea.

[tests-are-models]: https://lindeloev.github.io/tests-as-linear/
:::

:::{.callout-note title="See Also"}
This chapter is based on many sources (see [References][refs]), but two
textbooks were especially influential:

[refs]: /chapters/references.html

* [_Introduction to Modern Statistics_][ims] [@ims] provides a
  beginner-friendly introduction to data exploration and applied statistics. If
  you're new to statistics or want a gentle refresher, it's a great starting
  point for learning more.
* [_Regression and Other Stories_][ros] [@gelman2020] provides a thorough
  introduction to linear models (as well as other kinds of models) while
  keeping the focus solidly on realistic applications. It includes details
  about several intermediate to advanced topics that aren't covered here, such
  as Bayesian statistics.

[ims]: https://openintrostat.github.io/ims/
[ros]: https://avehtari.github.io/ROS-Examples/

For more specifically about interpreting models, we recommend [_Model to
Meaning_][model-to-meaning] [@arel-bundock2025], which is oriented around case
studies and serves as an introduction to the R package marginaleffects.

[model-to-meaning]: https://marginaleffects.com/
:::


