```{r}
#| echo: false
#| warning: false
library("dplyr")
squid = readRDS("data/1989-1994_ca_market_squid.rds")
squid = filter_out(squid, mass_g >= 400 | is.na(mass_g))
```

# Linear Models {#sec-linear-models}

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain what a linear model is
* Specify a linear model in terms of an outcome and predictors
* Fit a linear model on a dataset
* Describe the assumptions of a linear model
* Create diagnotic plots for a linear model
* Interpret diagnotic plots to decide whether the linear model assumptions are
  satisfied
* Explain some strategies for selecting features to use as predictors
* Interpret linear models and their estimates
* Conduct and interpret hypothesis tests for linear model coefficients
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggfortify
* ggplot2

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::


A common application of statistics is to study relationships between features.
@sec-relationships-between-features introduced visualizations and the
correlation statistic for this purpose. When the relationship between two or
more features roughly follows a line, we can fit a **linear model** (or linear
regression) that estimates the line's intercept and slope. Linear models are
suprisingly flexible: by transforming features, we can use linear models to
summarize many kinds of lines and curves. This chapter introduces the
fundamentals of linear models.

<!--
Any straight line we draw is a linear model: a line that approximates the
relationship between features.
-->


## Fitting a Model

To demonstrate how to fit, inspect, and interpret a linear model, we'll once
again turn to the California market squid dataset
(@sec-dataset-ca-market-squid). As a reminder, the structure is:

```{r}
str(squid)
```

Suppose we want to understand the relationship between the mass and dorsal
mantle length of squid. We can start investigating by making a scatter plot of
the two relevant features:

```{r}
#| label: fig-squid-scatter
library("ggplot2")

ggplot(squid) +
  aes(x = dorsal_mantle_length_mm, y = mass_g) +
  geom_point(alpha = 0.5)
```

Most of the squid are in a dense cloud of points with dorsal mantle length less
than 200 millimeters and mass less than 150 grams. Even though we removed some
outliers earlier (@sec-dataset-ca-market-squid), there are still three squid
with unusual measurements. We'll leave these in the dataset for now, but they
do warrant further investigation.

The cloud of points is triangular or slightly curved, not a perfect straight
line. We can get a better view of it by temporarily removing the outliers:

```{r}
#| warning: false
#| label: fig-squid-scatter-clean
library("dplyr")

squid_cloud = filter(squid, dorsal_mantle_length_mm <= 200, mass_g <= 150)

ggplot(squid_cloud) +
  aes(x = dorsal_mantle_length_mm, y = mass_g) +
  geom_point(alpha = 0.5)
```

In @fig-squid-scatter-clean, the relationship looks like it might be quadratic
or exponential. Nonetheless, we'll start by modeling the relationship with a
straight line and adjust later based on model diagnostics.

Setting the nature of the relationship aside, a single line or curve can't pass
through every point because of individual variation among the squid. The best
we can hope for is that it captures the general trend or pattern in the points.

We can make the linear model concrete by writing the equation for the line. A
line has an intercept, which we'll denote by $\beta_0$, and slope, which we'll
denote by $\beta_1$. In order to account for individual variation, we'll also
include a **residual** term in the equation, which we'll denote by $\epsilon$.
The residual is a reminder that this is just a model; when we plot the model or
use it to make predictions, we'll omit the residual. So the equation for the
line is:

$$
\texttt{mass\_g}
= \beta_0 + \beta_1\cdot\texttt{dorsal\_mantle\_length\_mm} + \epsilon
$$

One way to interpret this equation is that the feature on the left-hand side is
an **outcome** predicted by the features on the right-hand side, which we call
**predictors**. In our model, mass is the outcome and flipper length is the
only predictor.

:::{.callout-warning}
Despite the outcome-predictor interpretation, a linear model alone can't tell
us whether a relationship between features is causal. To conclude that dorsal
mantle length *causes* mass (or vice-versa), we'd have to make additional
assumptions.

For an introduction to causal inference, see "Part 5: Causal Inference" of
[_Regression and Other Stories_][ros] [@gelman2020].
:::

:::{.callout-note title="Note: Alternative Terminology" collapse="true"}
We use the names "outcome" and "predictor" for features in a linear model, but
these concepts go by many names, which are often discipline-specific. Also keep
in mind that features are called "variables" in most disciplines.

Other names for outcome include response, dependent feature, and endogenous
feature.

Other names for predictor include covariate, independent feature, and exogenous
feature.
:::

Writing the linear model in terms of mathematical variables rather than feature
names makes it more concise and more general (because the variables could refer
to any features from any dataset). To do so, let $(x_i, y_i)$ denote
observations (with two features) and $\epsilon_i$ denote residuals, for $i = 1,
2, \dots, n$. Then the equation becomes:

$$
y_i = \beta_0 + \beta_1 x_i + \epsilon_i
$$

This notation is standard in most disciplines.

:::{.callout-note title="Note: Hat Notation" collapse="true"}
In statistics, another convention is to write a "hat" above variables that are
estimators or estimates (including predictions). So we can write the model's
prediction for $y_i$ as

$$
\hat{y}_i = \beta_0 + \beta_1 x_i
$$

We can also use this notation for parameter estimates. For instance,
$\hat{\mu}$ denotes an estimate for a distribution parameter $\mu$.
:::


### Least Squares

Another way to think about the residuals in a linear model is as the **errors**
in the model's predictions. We can reorganize the equation for the line to
emphasize this:

$$
\epsilon_i = y_i - (\beta_0 + \beta_1 x_i)
$$

Thinking about the residuals as errors is a hint that we might want to choose
the coefficients $\beta_0$ and $\beta_1$ to minimize error. There are many
different ways to do this, but the most widely-used approach, called **ordinary
least squares**, is to minimize the **residual sum of squares** $SS_\epsilon$,
defined as

$$
\textrm{SS}_\epsilon
= \sum_{i=1}^n \epsilon_i^2
= \sum_{i=1}^n \left[ y_i - (\beta_0 + \beta_1 x_i) \right]^2
= \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

Squaring the residuals puts extra emphasis on errors that are very large
(whether positive or negative). This is desirable if large errors are more
costly than small errors. Squaring the residuals also leads to a variety of
convenient mathematical properties, such as making it straightforward to find
the minimum.

:::{.callout-note title="Note: Maximum Likelihood Estimation" collapse="true"}
A different approach, called **maximum likelihood estimation**, treats the
outcome as a random variable $Y_i$ with an assumed distribution. We can use
the distribution to choose estimates for the coefficients $\beta_0$ and
$\beta_1$ which maximize the probability of the data.

Ordinary least squares is equivalent to maximum likelihood estimation where we
assume that the $Y_i$ are normally distributed with means $\beta_0 + \beta_1
x_i$.

Maximum likelihood estimates have many useful statistical properties, including
an analogue of the Central Limit Theorem (@sec-central-limit-theorem), which we
can use to make confidence intervals and for hypothesis tests.
:::

In R, we can use the `lm` function to fit a linear model by least squares. The
function's first argument must be a description of the features in the model; R
calls this a (model) formula. The model formula is similar to the model
equation, but replaces the equals sign with a tilde `~` and omits the
coefficients and residual. The intercept is included implicitly, but you can
make it explicit by writing it as a `1` (imagine the intercept as a coefficient
for a column of ones).

The second argument must be the dataset: a data frame with columns
corresponding to the features in the model.

So the code to fit the mass model for the California market squid is:

```{r}
model = lm(mass_g ~ dorsal_mantle_length_mm, squid)
model
```

The output shows the estimates for the coefficients. We'll learn more about how
to interpret models in @sec-interpreting-models. Before that, we need to review
the assumptions of a linear model and examine whether they're satisfied for
this model.


## Model Assumptions {#sec-model-assumptions}

A linear model involves many assumptions about the features being modeled and
the dataset used to fit the model. @gelman2020 lists these assumptions (from
most to least important):

1. Validity: the outcome should measure what you actually want to study, the
   model should include all relevant features, and the model should generalize
   well. For instance, a model for blood pressure should not be treated as a
   model for general health unless this can be justified by expert knowledge.

2. Representativeness: the dataset you use to train the model should be
   representative of the population to which you want the model to generalize.
   For example, a model for how long people can hold their breath should not be
   based on a sample of Olympic swimmers. On the other hand, a representative
   sample might include a few Olympic swimmers, because they're people too!

3. Linearity: the relationship between the outcome and the predictors should be
   roughly linear. Often the relationship between the outcome and the features
   in the dataset isn't linear, but it's possible to transform the outcome or
   features to get a linear relationship (see @sec-selecting-features).

4. Independence of errors: the residual for one observation should not depend
   on the residuals of others. This assumption means linear models are
   incompatible with most temporal and spatial datasets, although there are
   modifications we can make to accomodate these kinds of datasets.

5. Equal variance of errors: the residuals should all have the same standard
   deviation. When they don't, then the model is worse at predicting for some
   inputs than others. One way to address this is by using weighted least
   squares to fit the model rather than ordinary least squares.

6. Normality of errors: the residuals should be have a normal distribution.
   This is assumption really only necessary for hypothesis testing (with
   CLT-based methods); it doesn't matter for the coefficient estimates. This
   assumption is sometimes misinterpreted to mean that the outcome must have a
   normal distribution, which is not true.

Check these assumptions before and after fitting a linear model. The first two
assumptions are really about the research question and data collection process,
so failing to satisfy them usually can't be fixed by changing the model. On the
other hand, the last four assumptions relate to the model directly, and failing
to satisfy them can sometimes be fixed by changing the model.

@sec-model-diagnostics describes how to check some of these assumptions with
model diagnostics.


## Model Diagnostics {#sec-model-diagnostics}

Most of the information about how well a model fits is in the residuals, since
the residuals measure what the model got wrong. We can make a variety of plots
based on the residuals to check whether the assumptions from
@sec-model-assumptions are satisfied and whether there are any other problems.

For a linear model with only one feature, a good diagnostic starting point is
to plot the feature against the residuals. This is called a **residual plot**.
Patterns in the residual plot can reveal problems with linearity, independence
of errors, and equal variance of errors. Let's make this plot for the residuals
from the mass model for the California market squid dataset.

In R, the `predict` function returns predictions from a model. Absent any
additional arguments, it returns predictions for the dataset used to fit the
model. So to get the predictions for the squid mass model:

```{r}
squid$predicted_mass_g = predict(model)
```

We can use the predictions to calculate the residuals, in order to plot the
residuals against the flipper length feature:

```{r}
#| label: fig-squid-residuals-predictor
#| fig-cap: >
#|    Dorsal mantle length versus residuals for the squid mass linear model.
ggplot(squid) +
  aes(x = dorsal_mantle_length_mm, y = mass_g - predicted_mass_g) +
  geom_point()
```

:::{.callout-tip}
The `resid` function returns the residuals from a model. It's a convenient
alternative to computing the residuals from the predictions.
:::

For a dataset and model that satisfy the assumptions (@sec-model-assumptions),
the residuals should be centered on 0 (mean 0) and should not exhibit any
patterns. Four kinds of patterns are especially important signs of problems:

* A trend in the residuals (that is, a line or curve that they follow): this
  usually means that the linearity assumption isn't satisfied.
* A "fan" or other shape in the range of the residuals: this usually means that
  the equal variance assumption isn't satisfied.
* Extreme residuals: like shapes, this usually means that the equal variance
  assumption isn't satisfied.
* Extreme predictor values: this means the representativeness assumption isn't
  satisfied. Even if the residuals are small, predictions might not generalize
  well because the model doesn't have many examples (observations) around the
  extreme value.

The residual plot will also exhibit patterns when the independence assumption
isn't satisfied, but many different patterns are possible depending on the
nature of the dependence.

The residuals for the squid mass model in @fig-squid-residuals-predictor
exhibit all four of the problem patterns. The residuals lie on a U-shaped
curve which tapers at each end. There are three extreme residuals, one of which
corresponds to an extreme predictor value; we should investigate these as
potential outliers and possibly exclude them when fitting the model.

For models with multiple predictors, we can plot each one separately against
the residuals. We can also plot the predictions against the residuals. These
plots are all residual plots, and the interpretation is the same as for a model
with a single feature, with the exception that we can't check for extreme
predictor values in the predictions versus residuals plot. Let's plot the
predictions against the residuals for the squid mass model:

```{r}
#| label: fig-squid-residuals-predicted
#| fig-cap: >
#|    Predictions versus residuals for the squid mass linear model.
ggplot(squid) +
  aes(x = predicted_mass_g, y = mass_g - predicted_mass_g) +
  geom_point()
```

Notice that @fig-squid-residuals-predictor and @fig-squid-residuals-predicted
only differ in the values on the x-axis. This is always the case for a model
with a single feature. For models with multiple features, the residual plots
for individual features will not necessarily be the same as the plot of
predictions versus residuals.

Most statistical software, including R, provide three more diagnostic plots for
linear models:

* The **normal quantile-quantile plot** (normal QQ plot) displays the quantiles
  of a normal distribution against the quantiles of the residuals. QQ plots are
  helpful for checking how similar two distributions are, because similar
  distributions will have similar quantiles. For a linear model, we can use the
  normal QQ plot to check the normality assumption. The normal QQ plot is less
  important than other diagnostics, because the normality assumption is less
  important than the other assumptions.
* The **scale-location plot** displays the predictions against the square root
  of the standardized residuals. This is a transformation of the residual plot,
  so the interpretation is the same (although the mean of the standardized
  residuals will not be 0); the transformation simply makes some patterns
  easier to notice.
* The **leverage plot** displays a statistic called leverage against the
  residuals. Leverage is a measure of how far away each observation in the
  dataset is from the others. Observations with high leverage may be outliers
  and may have undue influence over the linear model (similar to how extreme
  values affect the sample mean).

In R, we can use the ggfortify package's `autoplot` function to make all four
diagnostic plots for a model:

```{r}
#| warning: false
#| label: fig-squid-diagnostics
#| fig-cap: >
#|    Diagnostic plots for the squid mass linear model. Clockwise from
#|    the top left, they are the residual plot, normal QQ plot, leverage plot,
#|    and scale-location plot.
library("ggfortify")

autoplot(model)
```

The residual plot in @fig-squid-diagnostics is the same as
@fig-squid-residuals-predicted; the scale-location plot exhibits similar
problems. The normal QQ plot shows that the residuals are probably not normally
distributed, with extreme values in both tails. The leverage plot shows that
the observation with extreme dorsal mantle length has a lot of leverage; since
it also has a large residual, this means it's exerting a lot of influence on
the model without much benefit.

Let's refit the model without the extreme observations we noted in
@fig-squid-diagnostics (and also in @fig-squid-scatter). We can use the
`squid_cloud` subset from earlier for this:

```{r}
#| label: fig-squid-diagnostics-clean
#| fig-cap: >
#|    Diagnostic plots for the squid mass linear model, after removing extreme
#|    observations.
model = lm(mass_g ~ dorsal_mantle_length_mm, squid_cloud)
autoplot(model)
```

The new diagnostic plots in @fig-squid-diagnostics-clean show that removing the
extreme observations improved two problems: the most extreme residuals are
closer to the others, and the high-leverage residuals are smaller.
Nevertheless, there's still a distinct curve and a taper to the residuals in
the residual plot that suggest non-linearity and unequal variance,
respectively; we'll address these problems in @sec-selecting-features by adding
additional terms to the model.

:::{.callout-note collapse="true"}
```{r}
resid_df = data.frame(residual = resid(model))
resid_sd = sigma(model)
ggplot(resid_df) +
  aes(x = residual) +
  geom_density() +
  geom_function(fun = dnorm, args = list(sd = resid_sd), n = 1000, linetype =
    "dotted")
```
:::

:::{.callout-note title="Note: R's Built-in Diagnostic Plots" collapse="true"}
Instead of using the ggfortify package, to make diagnostic plots for a model
with ggplot2, it's also possible to make diagnostic plots with R's built-in
`plot` function:

```{r}
plot(model)
```
:::


:::{.callout-note title="See Also"}
For examples of patterns to look for in residual plots, see:

* The ["Checking model conditions"][ims-model-conditions] section in
  _Introduction to Modern Statistics_ [@ims]
* The "Residual plots" section in [_Regression and Other Stories_][ros]
  [@gelman2020]

[ims-model-conditions]: https://openintrostat.github.io/ims/inf-model-slr.html#sec-tech-cond-linmod
:::


## Selecting Features {#sec-selecting-features}

A dataset might have dozens or hundreds of features, but including every
feature in a linear model is usually not a good idea. One problem is that each
additional predictor in the model means an additional coefficient to estimate.
When the number of observations is low relative to the number of predictors,
there usually isn't enough information to compute accurate coefficient
estimates.

Another problem, **multicollinearity**, is that when two predictors are highly
correlated (@sec-correlation), they effectively contain redundant information.
As we add more predictors to the model, it becomes more likely that two (or
more) will be correlated. Multicollinearity makes it difficult (or impossible)
to compute accurate coefficient estimates for the affected predictors.

:::{.callout-tip}
If two predictors are highly correlated, consider removing one of them from the
model.
:::

The point is that we should be cautious about which predictors we add to a
model. The best approach is to base the model on expert knowledge or a
hypothesis about the relationship between the outcome and other features in the
dataset. In exploratory work, we can instead or additionally use the model
diagonistics (@sec-model-diagnostics) to decide whether it might be missing
important predictors (and whether adding a predictor improves the fit).

The diagnostic plots (@fig-squid-diagnostics-clean) for the squid mass model
exhibit a U-shaped curve in the residuals, and the scatter plot of dorsal
mantle length versus mass (@fig-squid-scatter-clean) also exhibits a curve. The
curve in both looks like it could be quadratic; this suggests we're missing a
squared predictor. For most physical objects, mass is proportional to volume,
which is roughly length cubed; considering this, it makes sense to include
squared or even cubed dorsal mantle length in the mass model.

Let's try adding squared dorsal mantle length to the mass model. We'll also
keep the original dorsal mantle length predictor; when including a power of a
feature in a model, it's often a good idea to include lower powers as well.

We can square the `dorsal_mantle_length_mm` feature in the model formula by
writing it inside of the `I` function (which makes R treat arithmetic in a
formula as arithmetic):

```{r}
quad_model = lm(
  mass_g ~ dorsal_mantle_length_mm + I(dorsal_mantle_length_mm^2),
  squid_cloud
)
quad_model
```

To see how this changes things, we'll recreate the diagnostic plots:

```{r}
#| label: fig-squid-diagnostics-quad
#| fig-cap: >
#|    Diagnostic plots for the squid mass linear model, with a squared dorsal
#|    mantle length term.
autoplot(quad_model)
```

The diagnostics in @fig-squid-diagnostics-quad are much improved. In the
residual plot, there's no longer any signs of a curve. There still is in the
scale-location plot, but it's minor. The residuals still taper at the tails of
the prediction range, but much less than they did in
@fig-squid-diagnostics-clean. The leverage plot also shows improvement, with
high-leverage residuals nearly halved. The QQ plot is about the same. On the
whole, the diagnostics are now good enough to keep the model, with some caution
when interpreting hypothesis tests that assume normally distributed residuals.

:::{.callout-note title="Note: Adding a Cubic Term" collapse="true"}
Based on our reasoning earlier, we might want to try adding cubed dorsal mantle
length to the model too:

```{r}
#| label: fig-squid-diagnostics-cubic
#| fig-cap: >
#|    Diagnostic plots for the squid mass linear model, with a cubed dorsal
#|    mantle length term.
cubic_model = lm(
  mass_g ~
    dorsal_mantle_length_mm +
    I(dorsal_mantle_length_mm^2) +
    I(dorsal_mantle_length_mm^3),
  squid_cloud
)

autoplot(cubic_model)
```

Overall, @fig-squid-diagnostics-cubic is similar to
@fig-squid-diagnostics-quad. There's a little improvement in the high-leverage
residuals, but any other improvements are subtle. At this point, we can leave
the cubic term out to make the model simpler, or include it, because it matches
our intuition about the relationship between mass and volume. We can also check
other diagnostics, such as adjusted R-squared (more about this in
@sec-interpreting-models) or an estimate of the model's accuracy on new data
points (for example, by cross-validating the model).
:::


### Categorical Predictors

Up to this point, we've only considered linear models with numerical
predictors. Linear models can also have categorical predictors. For a
categorical predictor, we add a constant offset to the prediction for each
category. We still call the offsets coefficients, because each category can be
represented by a feature that's 1 for observations in the category and 0
otherwise.

:::{.callout-note title="Note: Indicators"}
The 0 and 1 representation for a category is called an **indicator** or
**one-hot encoding**.
:::

In R and most other statistical software, the first category is treated as a
baseline or reference category with coefficient 0, so a categorical predictor
with $k$ categories only requires $k - 1$ coefficient estimates.

Let's try fitting a model with a categorical predictor to the California market
squid dataset. Suppose we want to examine whether the age of a squid affects it
size. We'll use mass as a substitute for size use maturity as substitute for
age. Then the linear model is:

```{r}
maturity_model = lm(mass_g ~ maturity, squid_cloud)
maturity_model
```

There are three maturity categories: immature, mature, and spent. The model
treats immature as the baseline category, so the model output only lists
coefficients for the mature and spent categories. Think of these coefficients
as the additional mass for the given category compared to the reference
category.

As usual, it's a good idea to check the model diagnostics:

```{r}
#| label: fig-squid-diagnostics-maturity
#| fig-cap: >
#|    Diagnostic plots for the squid mass linear model based on maturity alone.
autoplot(maturity_model)
```

In @fig-squid-diagnostics-maturity, notice that the there are only three
different predicted values; this reflects that there are only three maturity
categories, and maturity is the only predictor in the model. The residual plot
and scale-location plot don't show any trends. On the other hand, the residuals
spread out more for more mature squid, which means there's unequal variance.
The QQ plot shows that the residuals are probably not normally distributed. The
leverage plot looks okay. The takeaway for this model is that the coefficient
and standard error estimates are likely accurate, but because there's evidence
that the residuals do not have equal variance and are not normally distributed,
we should be cautious about using standard hypothesis tests for the model.
There are several test statistics designed for groups with unequal variance; we
should use one of these if we want to conduct a test.


### Interactions

Sometimes an outcome depends on how two or more predictors **interact**. That
is, the relationship between the outcome and one of the predictors changes when
the other predictor changes. We can model a multipicative interaction in a
linear model by including a term for the product of two or more features
(whether they are numerical or categorical). Generally, if we include an
interaction term for two features, we should also include terms for each
feature separately.


### Transformations

Mathematically transforming the outcome or predictors in a linear model is
often helpful when the assumptions of linearity, equal variance of errors, and
normality of errors aren't satisfied. We saw an example of this at the
beginning of this section, where adding a quadratic term to the squid mass
model corrected problems with non-linearity and unequal variance.

Choosing appropriate transformations can be difficult in practice, because we
rarely know the true nature of the relationship. Fortunately, relationships
that a model fails to account for typically show up as patterns in its
residuals (@sec-model-diagnostics). We can use the residuals to decide if and
how features should be transformed.

:::{.callout-tip}
Many kinds of non-linear relationships can be made linear with transformations.

For example, suppose that the true relationship between two features $x_i$ and
$y_i$ is exponential:

$$
y_i = \textrm{e}^{\beta_0 + \beta_1 x_i}
$$

We can make the relationship linear by taking logarithms:

$$
\ln{y_i} = \beta_0 + \beta_1 x_i
$$

In this case, we should fit a linear model with $\ln{y_i}$ as the outcome and
$x_i$ as a predictor.
:::

:::{.callout-note title="See Also"}
We recommend the textbook [_Feature Engineering and
Selection_][feature-engineering] [@kuhn2020] as a thorough introduction to
transforming and selecting features for linear models, as well as many other
kinds of models.

[feature-engineering]: https://feat.engineering/
:::


## Interpreting Models {#sec-interpreting-models}

Once we've fitted a model and checked that the diagnostics look okay, we can
move on to interpreting the coefficients and other model statistics. This
section describes a few different tools for interpretation.


### Model Summary

We can use R's `summary` function to get a detailed description of a model:

```{r}
summary(quad_model)
```

Let's go through the model summary a few lines at a time. It begins with the R
code we used to fit the model:

```{.text}
Call:
lm(formula = mass_g ~ dorsal_mantle_length_mm + I(dorsal_mantle_length_mm^2),
    data = squid_cloud)
```

Next, it lists the quartiles of the residuals:

```{.text}
Residuals:
    Min      1Q  Median      3Q     Max
-35.237  -3.267   0.521   3.775  52.794
```

The distribution of the residuals is an important source of diagnostic
information. We recommend visualizing the residuals, as in
@sec-model-diagnostics, but their quartiles are occasionally useful for quickly
inspecting a model.

After the residual quartiles, the summary lists the coefficients in a table:

```{.text}
Coefficients:
                              Estimate Std. Error t value Pr(>|t|)
(Intercept)                  77.386910   3.670391   21.08   <2e-16 ***
dorsal_mantle_length_mm      -1.538497   0.057980  -26.54   <2e-16 ***
I(dorsal_mantle_length_mm^2)  0.009674   0.000228   42.44   <2e-16 ***
---
Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
```

Each row of the table corresponds to one term in the model. The first is
typically the intercept $\beta_0$. The remaining rows correspond to features.
The columns contain an estimate of the coefficient, an estimate of its standard
error, and two columns related to hypothesis testing. We'll examine these last
two columns in @sec-hypothesis-tests-for-models.


#### Residual Standard Deviation {.unnumbered}

Finally, the summary lists several statistics about the model as a whole,
beginning with:

```{.text}
Residual standard error: 5.9 on 6191 degrees of freedom
```

Residual standard error is the sample standard deviation of the residuals (with
a statistical correction for the number of terms being estimated in the model).
Recall that standard deviation is approximately the mean distance to the mean.
In a linear model with an intercept, the residuals always have mean 0
($\bar{\epsilon} = 0$), so the residual standard error is approximately the
mean distance between the actual outcomes and the fitted line.

:::{.callout-note title="Definition: Residual Standard Deviation" collapse="true"}
The sample variance of the residuals for a linear model with $k$ terms
(including the intercept) is defined as

$$
s_\epsilon^2
= \frac{\textrm{SS}_\epsilon}{n - k}
= \frac{1}{n - k} \sum_{i=1}^n (\epsilon_i - 0)^2
$$

Think of this as the variance in the sample that's not explained by the model.

The sample standard deviation of the residuals $s_\epsilon$ is the square root
of $s_\epsilon^2$.
:::

:::{.callout-note title="Note: The Name \"Residual Standard Error\"" collapse="true"}
The name "residual standard error" is an R invention. Standard error refers to
the standard deviation of an estimator across many samples, but there's no
estimator and no resampling at work here. The R documentation even notes that
this name is inappropriate (in `?sigma`).

Appropriate names include "standard deviation of the residuals" and "residual
standard deviation" among others.

See [this StackOverflow post][so-rse] for additional discussion.

[so-rse]: https://stats.stackexchange.com/questions/144433/why-do-we-say-residual-standard-error
:::


#### R-Squared Statistics {.unnumbered}

```{.text}
Multiple R-squared:  0.8404,    Adjusted R-squared:  0.8403
```

The R-squared statistics, conventionally denoted by $R^2$ and
$R_\textrm{adj}^2$, are ratios of variance in the outcome that's explained by
the model to total variance.

R-squared $R^2$ is always between 0 and 1. Values closer to 1 indicate that the
model explains more of the variance in the outcome. R-squared is non-decreasing
as more terms are included in the model, because more terms means more
flexibility to fit the observations. This is a limitation: we can't use
R-squared to decide how many terms to include in the model, because it will
always suggest more terms are better.

Adjusted R-squared $R_\textrm{adj}^2$ decreases when terms with weak
explanatory power are added to the model, so it tends to be more useful than
ordinary R-squared $R^2$. Its value can sometimes be smaller than 0.

:::{.callout-note title="Definition: R-Squared Statistics" collapse="true"}
The R-squared statistics are based on the idea that the variance of the outcome
$y_i$ can be partitioned into two meaningful parts: variance explained by the
model and unexplained variance from other, often unknown, processes.

Mathematically, variance is a sum of squares. The **total sum of squares** for
the outcome $y_i$ is defined as

$$
\textrm{SS}
= \sum_{i=1}^n (y_i - \bar{y})^2
= (n - 1) s^2
$$

Notice that the total sum of squares is part of and proportional to the sample
variance $s^2$ (@sec-standard-deviation).

We can partition the total sum of squares into two parts, the **explained sum
of squares** $\textrm{SS}_\textrm{model}$ and the residual sum of squares
$\textrm{SS}_\epsilon$:

$$
\textrm{SS}
= \textrm{SS}_\textrm{model} + \textrm{SS}_\epsilon
= \sum_{i=1}^n (\hat{y}_i - \bar{y})^2 + \sum_{i=1}^n (y_i - \hat{y}_i)^2
$$

Then R-squared $R^2$ is defined in terms of the sums of squares as

$$
R^2
= 1 - \frac{\textrm{SS}_\epsilon}{\textrm{SS}}
= \frac{\textrm{SS}_\textrm{model}}{\textrm{SS}}
$$

For a model with $k$ terms (excluding the intercept), the adjusted R-squared
$R_\textrm{adj}^2$ is defined as

$$
R_\textrm{adj}^2
= 1 - \frac{\textrm{SS}_\epsilon}{n - k - 1} \cdot \frac{n - 1}{\textrm{SS}}
= 1 - \frac{s_\epsilon^2}{s^2}
$$

The denominator $n - k - 1$ accounts for the number of terms in the model.
:::

R-squared, whether adjusted or not, mostly tells us about how well the model
fits *the sample*. A model that fits a sample well might not fit the population
well, and a model that fits the population well might not fit a sample well.
Moreover, for some populations, observations are always noisy (accurate
measurement is difficult). Then R-squared can be large even if the model fits
the population well.

:::{.callout-warning}
Treat R-squared as one statistic among many when examining a model.

To estimate how well a model will generalize, use other methods specifically
designed for that purpose, such as cross-validation.
:::


#### F Statistic {.unnumbered}

```{.text}
F-statistic: 1.63e+04 on 2 and 6191 DF,  p-value: < 2.2e-16
```

The F statistic is the ratio of variance in the outcome that's explained by the
model to unexplained variance. The F statistic is always positive.

The F statistic is also the relative difference between the residual sum of
squares (unexplained variance) for a simplified model with only an intercept
(the same as predicting the outcome with the sample mean) and for the fitted
model. This perspective is helpful for interpreting the statistic: larger
values indicate that the fitted model is better at explaining the variance of
the outcome than the simplified model.

:::{.callout-note title="Definition: F Statistic" collapse="true"}
For a model with $k$ terms (excluding the intercept), the F statistic is
defined as

$$
F
= \frac{\textrm{SS}_\textrm{model}} {\textrm{SS}_\epsilon}
  \cdot \frac{n - k - 1} {k}
= \frac{\textrm{SS} - \textrm{SS}_\epsilon}{\textrm{SS}_\epsilon}
  \cdot \frac{n - k - 1}{k}
$$
:::

The F statistic is primarily used for a hypothesis test where the null
hypothesis is that the model coefficients (excluding the intercept) are all 0.
As @gelman2020 point out, this null hypothesis is unrealistic---if we believe
it's plausible that a predictor is entirely unrelated to the outcome, we
shouldn't include it in the model in the first place. The F test is often used
to decide whether a model with fewer features might be more appropriate, but
there are better approaches, such as examining standard errors for the
coefficient estimates to decide whether the estimates are too imprecise to be
useful.

:::{.callout-warning}
As with R-squared, treat the F statistic as one statistic among many when
examining a model.
:::

:::{.callout-important}
Treat every model as a simplification of reality. Statistics can help us
understand how well a model describes data, but can't tell us whether a model
is "correct" (and it's unclear what it means for a simplification to be
"correct" anyway).
:::


### Predictions and Marginal Means

When a model only has one or two predictors, and they're not transformed, we
often interpret the model by examining the coefficients directly.
Interpretation is more difficult for models with a many predictors, especially
if some are transformed or have interactions. We might also want to ask
questions about the model that aren't easy to answer based on the coefficients
alone, such as how the outcome varies across different subpopulations of
interest.

One solution to these problems is to examine **marginal means**, means of (or
"marginalized over") the predictions on a grid of predictor values. The choice
of grid depends on the question we want to answer. In some disciplines,
marginal means are also called **marginal effects**.

For example, suppose we want to investigate whether a new treatment lowers
blood pressure. We conduct a study where half of the subjects receive the
treatment and half don't. In addition to blood pressure, we also collect other
information about the people in the study, such as their age and medical
history. Although we can compare the blood pressure of those who received the
treatment to those who didn't directly, what we really want is to compare
**counterfactuals**: what would the blood pressure of people who received the
treatment be if they hadn't received it, and what would the blood pressure of
people who didn't receive the treatment be if they had? We can use marginal
means to address these questions as well as many others.

In R, the [marginaleffects][] package and the [emmeans][] package each provide
a variety of functions to compute marginal means. The also have many other
features, such as functions to create grids, carry out hypothesis tests, and
compute other aggregates of model predictions.

[marginaleffects]: https://doi.org/10.32614/CRAN.package.marginaleffects
[emmeans]: https://rvlenth.github.io/emmeans/

:::{.callout-note title="See Also"}
To learn more about how to interpret models with marginal means, we recommend
[_Model to Meaning_][model-to-meaning] [@arel-bundock2025]. The book is
oriented around case studies and serves as an introduction to the R package
marginaleffects.

[model-to-meaning]: https://marginaleffects.com/
:::


## Testing Models {#sec-hypothesis-tests-for-models}

Linear models are often useful in data analyses where we want to test some
hypothesis (@sec-statistical-tests). Three different approaches (among many)
that we can take to test coefficients are:

* Permutation tests: we can test whether the outcome and all predictors are
  correlated (that is, null hypothesis $\beta_1 = \dots = \beta_k = 0$ for a
  model with $k$ predictors). However, we can't test individual coefficients
  with a permutation test, and we can't test for non-zero coefficient values.

* Bootstrap confidence intervals: we can resample the residuals with
  replacement, adding them back to the model predictions, to generate bootstrap
  estimates and confidence intervals for the coefficients and other statistics.
  By inverting the confidence intervals, we can test individual coefficients
  and other statistics.

* Classical tests: by assuming that the errors are independent, normally
  distributed, and have equal variance, we can use classical tests (typically
  based on Student's t distribution) for the individual coefficients and other
  statistics. Specifically, in the single predictor case, we assume for $i = 1,
  \dots, n$ that

  $$
  y_i = \beta_0 + \beta_1 x_i + \epsilon_i
  \quad\textrm{where}\quad
  \epsilon_i \sim \mathcal{N}(0, \sigma^2)
  $$

  In practice, these tests tend to have nearly-correct error rates as long as
  the sample size is large ($n \geq 30$) and the distribution of errors is
  symmetrical, even if it isn't normal.

For all of these, the linear model assumptions of validity, representativeness,
linearity, independence of errors, and equal variance of errors are critical.
In general, failing to satisfy these assumptions increases testing error rates,
which makes the discernibility level (and power estimates) inaccurate.

R's built-in functions focus exclusively on classical tests, and the classical
tests work well in most cases, so that's what we'll cover here. The `summary`
function displays the standard score for each coefficient. Under the linear
model assumptions, for a model trained on $n$ observations and $k$ predictors,
each standard score comes from a t distribution with normality $n - k - 1$.

The `summary` function also displays a p-value for each coefficent for a test
with null hypothesis $H_0: \beta_i = 0$ and alternative hypothesis $H_A:
\beta_i \neq 0$. This p-value is computed from the t distribution.

As an example, let's look at the summary for the squid mass model with the
squared dorsal mantle length term:

```{r}
summary(quad_model)
```

In this case, at level $\alpha = 0.05$, the coefficients for both dorsal mantle
length and squared dorsal mantle length are statistically discernable. There's
strong evidence in the data that they should not be 0. For this model, the QQ
plot (@fig-squid-diagnostics-quad) indicated that the residuals are not
normally distributed, and that they're skewed rather than symmetrical, so the
error rate of the test might be higher than chosen level 5%. Nevertheless,
because this is a very large dataset and the relationship matches our intuition
about the physical world, we have no reason to doubt that the coefficients
should not be 0.

The built-in R function `confint` returns classical (t distribution) confidence
intervals for the coefficients in a linear model:

```{r}
confint(quad_model)
```

These provide more insight into the range of likely values for each coefficient
than the p-values alone.

The [marginaleffects][] package and [emmeans][] package provide functions to
test more sophisticated hypotheses. They support a variety of testing methods,
including the bootstrap and classical tests.

:::{.callout-note title="See Also"}
This chapter is based on many sources (see [References][refs]), but two
textbooks were especially influential:

[refs]: /chapters/references.html

* [_Introduction to Modern Statistics_][ims] [@ims] provides a
  beginner-friendly introduction to data exploration and applied statistics. If
  you're new to statistics or want a gentle refresher, it's a great starting
  point for learning more.
* [_Regression and Other Stories_][ros] [@gelman2020] provides a thorough
  introduction to linear models (as well as other kinds of models) while
  keeping the focus solidly on realistic applications. It includes details
  about several intermediate to advanced topics that aren't covered here, such
  as Bayesian statistics.

[ims]: https://openintrostat.github.io/ims/
[ros]: https://avehtari.github.io/ROS-Examples/

DataLab also has two older workshops about linear models:

* [A Practical Introduction to Regression Modeling in R][dl-reg-model]
* [Model Validation for Applied Data Science][dl-model-valid]

[dl-reg-model]: https://ucdavisdatalab.github.io/workshop_regression/
[dl-model-valid]: https://ucdavisdatalab.github.io/workshop-model-validation/
:::
