# Quantifying Uncertainty

::: {.callout-note title="Learning Goals" collapse="false"}
After this lesson, you should be able to:

* Explain why we use standard deviation and other measures of scale to measure
  uncertainty
* Explain what a sampling distribution is
* Explain the difference between a population distribution, an empirical
  distribution, and a sampling distribution
* Use the bootstrap to estimate the sampling distribution of an estimator
* Explain what a confidence interval is
* Use the bootstrap to compute a confidence interval for an estimate
* Explain what the Central Limit Theorem is and why we can use it to construct
  confidence intervals
:::

:::{.callout-important title="Required Packages"}
This chapter uses the following R packages:

* dplyr
* ggplot2
* purrr

For a refresher about how to install and load packages, see the
[Packages][r-basics-packages] section of DataLab's R Basics workshop reader.

[r-basics-packages]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/getting-started/getting-started.html#sec-packages
:::

Estimating things is one of the primary motivators of statistics. There are
many different ways to estimate things, so we need to be able to assess how
well an estimator works. In the same spirit, we need a way to decide how much
confidence we should put in an estimate. This chapter introduces a versatile
strategy for addressing both needs. It also briefly explains a second strategy
that was developed before computers became widely available.


## The Sampling Distribution {#sec-sampling-distribution}

When we do data analysis, it's usually because we want to understand the
characteristics of some population. In many cases, we can't collect data for
the entire population because doing so would be unethical (for instance, when
testing a new, unproven medicine) or because the cost (in time, money, etc.) is
prohibitive. Instead, we collect a sample from the population, in hope that
results about the sample generalize to the population. Once we have a sample,
we can use it to compute estimates.

:::{.callout-note title="See Also"}
@sec-collection-strategy from DataLab's Exploring Data workshop reader also
discusses sampling.
:::

As an example, suppose we want to understand height differences (in inches)
between adult mothers and daughters. We probably can't collect height
differences for every mother and daughter in the world, or country, or even
local region. But we can choose some feasible number of height differences, say
50, to collect as a sample. After collecting the sample, suppose we compute:

```{r}
#| echo: false
#| output: false
set.seed(1215)
sample1 = rnorm(50)
sample2 = rnorm(50)
sample3 = rnorm(50)
```

```{r}
summary(sample1)
```

So the average difference in the sample is about `r round(mean(sample1), 2)`,
and we hope that this is also a good estimate of the average difference in the
population.

To get a better sense of how the data are distributed, we can also make a
density plot:

```{r}
#| label: fig-sample1-density
#| fig-cap: >
#|     Density plot of the empirical distribution of the first sample (50
#|     observations).
library("ggplot2")

ggplot(data.frame(x = sample1)) +
  aes(x = x) +
  geom_density()
```

The density plot in @fig-sample1-density shows an **empirical distribution**,
the distribution of the observations *in a dataset*. Here the dataset is the
sample of height differences. Like the sample mean estimates the population
mean, the emprical distribution estimates the population distribution. The
population distribution might have longer tails, or be more symmetrical, or
have some other characteristic that's underrepresented in the dataset. We turn
to the empirical distribution because it's the best we can do with the sample
we have.

Suppose our colleagues decide to collect two more samples of 50 height
differences each, to compare the results. For sample 2, the statistics are:

```{r}
summary(sample2)
```

And for sample 3, the statistics are:

```{r}
summary(sample3)
```

The sample mean varies from one sample to the next, because each sample
consists of different members of the population. If we continued to take
samples, we'd end up with many different sample means, all estimates of the
same population mean.

Now let's change our perspective to thinking about the sample means themselves
as a dataset. This dataset tells us about the characteristics of the sample
mean as an estimator. We can use a density plot to visualize the distribution
of the sample means. For 1,000 sample means (from 1,000 samples), it looks like
this:

```{r}
#| echo: false
#| label: fig-sample_mean-density
#| fig-cap: >
#|     Density plot of the sampling distribution of the mean, based on 1,000
#|     estimates.

n = 50  # sample size
k = 1000  # number of samples
x = rnorm(k * n)
m = rowMeans(matrix(x, k, n))

ggplot(data.frame(sample_mean = m)) +
  aes(x = sample_mean) +
  geom_density()
```

The density plot in @fig-sample_mean-density shows a **sampling distribution**,
the distribution of an estimator across many samples. In this case, it's the
sampling distribution for the sample mean in the height differences example. We
can see that the sample means are centered near 0 and vary around it
symmetrically. All are within about 0.4 inches.

:::{.callout-important}
The sampling distribution of an estimator, like the one in
@fig-sample_mean-density, will generally be different from the population
distribution. It will generally also be different from empirical distributions
(of samples), like the one in @fig-sample1-density, since they estimate the
population distribution.

Comparing @fig-sample_mean-density and @fig-sample1-density, the empirical
distribution is about 10 times more spread out.

Even if we combine all 1,000 samples and make a density plot, the distribution
is notably different from the sampling distribution of the mean in
@fig-sample_mean-density:

```{r}
#| echo: false
#| label: fig-1000samples-density
#| fig-cap: >
#|     Density plot of the empirical distribution of all 1,000 samples
#|     combined (50,000 observations).
ggplot(data.frame(x = x)) +
  aes(x = x) +
  geom_density()
```
:::

The scale (@sec-scale-statistics) of a distribution measures uncertainty about
what the value of an individual observation will be. When the scale is large,
many values are possible, so uncertainty is high. When the scale is small, only
a few values are possible, so uncertainty is low. In the extreme case, when the
scale is as small as it can be, the distribution is no longer random: it's a
single, constant value that we know every observation will take.

By the same reasoning, the scale of a sampling distribution measures
uncertainty about the estimates. When the scale is large, estimates will vary a
lot from sample to sample, so we might not trust any single estimate. When the
scale is small, the estimates tend to agree regardless of the sample, so we can
be more certain. In other words, we can judge how confident to be about an
estimator based on the scale of its sampling distribution.

For the sample means that lead to @fig-sample_mean-density, the sample standard
deviation is about `r round(sd(m), 2)`. The standard deviation of an estimator
is called its **standard error**, because it helps us judge how uncertain to be
about the estimates. In other words, it informs us about how much error there
might be in the estimates *due to random variation between samples*. The larger
the standard error is, the more uncertain we are about the estimate.

:::{.callout-important title="Important: Bias"}
Variation between samples is one of two major sources of estimator error. The
other is statistical **bias**, where an estimator estimates the wrong quantity.
Notice that this definition of bias is much narrower and more specific than the
definition of bias in ordinary speech and writing.

We say an estimator is **unbiased** if it estimates the correct quantity on
average. This is the same as the sampling distribution being centered on the
correct quantity.

Minimizing standard error is often at odds with minimizing bias. For instance,
we can always come up with a biased estimator with lower standard error than
even the best unbiased estimator. We call this tension the **bias-variance
tradeoff** (variance is the squared standard error). Whether or not it's
appropriate to accept some bias to reduce standard error depends on the
specific estimation problem.
:::

:::{.callout-note title="Note: Consistency"}
Consistency is another important statistical property. We say an estimator is
**consistent** if the estimates get closer to being correct as the sample size
increases.

An unbiased estimator is consistent if its standard error decreases as the
sample size increases. Biased estimators can also be consistent.
:::


## The Bootstrap Algorithm {#sec-bootstrap-algorithm}

Most of the time, we only have one sample. And if we had more than one sample,
we'd combine them into a single sample, since more data generally leads to
better estimates. So if we want to evaluate an estimator, we need a way to
estimate its sampling distribution with only one sample.

The **bootstrap**, or bootstrap algorithm, is one such method. The main idea is
to draw new samples from the dataset (which is itself a sample) rather than the
population. Assuming the dataset is representative of the population, these new
**bootstrap resamples** should have similar characteristics to a sample from
the population. The bootstrap resamples must be drawn with replacement. Another
way to think about this is that we're using the known empirical distribution as
a substitute for the unknown population distribution.

Once we have several bootstrap resamples, we can compute an estimate on each
one. This gives us a collection of **bootstrap estimates**. By treating the
bootstrap estimates as a dataset in their own right, we can analyze the
**bootstrap sampling distribution** of the estimator and use it to estimate
statistics such as standard error.

:::{.callout-note title="Algorithm: The Bootstrap"}
Suppose we have a dataset $x_1, x_2, \dots, x_n$ and an estimator $G$.

1. Construct a new bootstrap sample $x_1^*, x_2^*, \dots, x_3^*$ by sampling
   from $x_1, x_2, \dots, x_n$ with replacement.
2. Compute an estimate $G^* = G(x_1^*, x_2^*, \dots, x_n^*)$ on the bootstrap
   sample. This is a bootstrap estimate.
3. Repeat steps 1-2 until we have $b$ bootstrap estimates. Larger values of $b$
   are better but increase computing time; choose at least $b \geq 1,000$.
4. Use the bootstrap estimates to estimate standard error or other properties
   of the sampling distribution of $G$.
:::

A bootstrap sampling distribution is a useful but imperfect estimate of the
sampling distribution. Unlike the sampling distribution, it's centered on the
estimate for the original dataset. On the other hand, the scale is about the
same as in the sampling distribution, so we can use it to estimate standard
error.

Let's apply the bootstrap to the height differences example from
@sec-sampling-distribution. We can generate a bootstrap sample from the
original sample with this code:

```{r}
boot = sample(sample1, replace = TRUE)
```

The density plot for the bootstrap resample is:

```{r}
#| label: fig-boot1-density
#| fig-cap: >
#|     Density plot of the first bootstrap resample (50 observations).
ggplot(data.frame(x = boot)) +
  aes(x = x) +
  geom_density()
```

@fig-sample1-density shows the distribution the original sample, while
@fig-boot1-density shows the distribution of the bootstrap resample. The two
distributions are similar, but in the bootstrap resample, the mode at -1 is
much more pronounced. The sample mean is also notably lower on the bootstrap
resample (the original was `r round(mean(sample1), 2)`):

```{r}
mean(boot)
```

We can use R's apply functions or the purrr package's map functions to repeat
the resampling step many times. We'll compute $b = 1,000$ resamples to keep the
computing time manageable, but in practice $b = 10,000$ or higher is more
appropriate:

```{r}
library("purrr")

b = 1000
boot_resamples = map(seq_len(b), \(i) {
  sample(sample1, replace = TRUE)
})
```

:::{.callout-note title="See Also"}
For a review of R's apply functions and the purrr package, see the
[Loops][r-basics-loops] section of DataLab's R Basics workshop reader.

The [documentation for the purrr package][purrr] is also a useful reference.

[r-basics-loops]: https://ucdavisdatalab.github.io/dwr_workshop_r_basics/chapters/organizing-code/control-flow.html#loops
[purrr]: https://purrr.tidyverse.org/
:::

Each of the bootstrap resamples is slightly different. To demonstrate this,
here are density plots for several, as well as the density plot for the
original sample:

```{r}
#| label: fig-bootstrap_resamples-density
#| fig-cap: >
#|     Density plot of five bootstrap resamples, as well as the original
#|     sample.
slice = 1:5
df = data.frame(
  x = c(sample1, unlist(boot_resamples[slice])),
  # Make a categorical feature to identify each resample.
  sample = rep(
    c("sample1", paste0("resample", slice)),
    each = length(sample1)
  )
)

ggplot(df) +
  aes(x = x) +
  geom_density() +
  facet_wrap(vars(sample))
```

The next step is to compute an estimate, in this case the sample mean, on each
of the bootstrap resamples. Again we can use the purrr package for this:

```{r}
boot_means = map_dbl(boot_resamples, mean)
```

With the bootstrap estimates, we can make a density plot:

```{r}
#| label: fig-bootstrap_mean-density
#| fig-cap: >
#|     Density plot of 1,000 bootstrap sample means. This distribution
#|     approximates the sampling distribution of the mean.
ggplot(data.frame(bootstrap_mean = boot_means)) +
  aes(x = bootstrap_mean) +
  geom_density()
```

Two details of @fig-bootstrap_mean-density are important to notice. First, the
distribution is centered at roughly `r round(mean(sample1), 2)`, the mean of
the original sample. This is always the case for a bootstrap sampling
distribution, and a major difference from the actual sampling distribution,
which will typically be centered near the population parameter.

Second, notice that the spread of the bootstrap sampling distribution is about
the same as the spread of the sampling distribution in
@fig-sample_mean-density. In fact, the standard deviation of the bootstrap
estimates is:

```{r}
sd(boot_means)
```

This is nearly `r round(sd(m), 2)`, the standard deviation of the sampling
distribution. Most of the time, the scale of the bootstrap sampling
distribution is relatively similar to the scale of the actual sampling
distribution.

The takeaway is that the bootstrap sampling distribution is useful for
estimating standard error, even though its center (and sometimes its shape) can
differ from the actual sampling distribution.

:::{.callout-warning}
The bootstrap is known to perform poorly in a few specific situations:

* When the size $n$ of the original sample is small (say $n < 30$). In this
  case, the methods in @sec-central-limit-theorem are usually a better choice.
* When the original sample is a poor representative of the population. This is
  a difficult case, because the best solution is to collect more and better
  data.
* When the estimator is designed to estimate an extreme value, such as the
  population minimum or maximum. In this case, there are often sophisticated
  methods tailored specifically to the estimator that perform better.
:::


## Bootstrap Percentile Intervals {#sec-bootstrap-percentile-intervals}

We can account for uncertainty about an estimate by instead presenting an
interval of estimates. We call this a **confidence interval**. The interval
should be narrow when we're relatively confident (small standard error) and
wide when we're uncertain (large standard error). @ims explain confidence
intervals with a helpful analogy: a single estimate is like fishing with a
spear, whereas a confidence interval is like fishing with a net. With the
confidence interval, we have a better chance of estimating correctly.

There are many ways to construct confidence intervals for an estimate. A
relatively simple way is to use percentiles of the bootstrap sampling
distribution. This is called a **bootstrap percentile interval**. For example,
if we take the 2.5th and 97.5th percentiles of the bootstrap means, we get:

```{r}
quantile(boot_means, c(0.025, 0.975))
```

We estimate that the population mean is somewhere in this interval, although as
with any estimate, this might be wrong. We call this a 95% bootstrap percentile
interval because it covers the central 95% of the bootstrap sampling
distribution. For reasons beyond the scope of this reader, there's also a 95%
chance that the interval covers the population mean.

<!--
For the sampling distribution of means in @fig-sample_mean-density, the 2.5th
percentile is about `r round(quantile(m, 0.025), 2)` and the 97.5th percentile
is about `r round(quantile(m, 0.975), 2)`. So a confidence interval for the
sample mean is:

$$
(-0.27,\ 0.27).
$$

We say this is a 95% percent confidence interval. The interval covers
approximately 95% of the plausible estimates. For reasons that go beyond the
scope of this reader, there's also an approximately 95% chance that it covers
the population mean. This confidence interval happens to be centered on 0, but
there's no reason a confidence interval has to be. We'll revisit confidence
intervals in @sec-bootstrap-percentile-intervals.
-->

:::{.callout-warning}
In situations where the bootstap performs poorly (see
@sec-bootstrap-algorithm), the bootstrap percentile intervals will usually be
too narrow. In other words, they tend to understate uncertainty and have less
than a 95% chance of covering the population parameter.
:::

:::{.callout-note title="See Also"}
For another introduction to the bootstrap and confidence intervals, see the
[Confidence intervals with bootstrapping][ims-boot] chapter, as well as
subsequent chapters, of _Introduction to Modern Statistics_ [@ims].

[ims-boot]: https://openintrostat.github.io/ims/foundations-bootstrapping.html

To learn more about the bootstrap, including other ways to bootstrap and to use
the bootstrap to construct confidence intervals, see _Bootstrap Methods and
Their Application_ [@davison1997].

In R, the [boot][] package provides functions to carry out a bootstrap and
compute confidence intervals.

[boot]: https://doi.org/10.32614/CRAN.package.boot

The [infer][] package provides functions for many methods in applied
statistics, with an interface inspired by and compatible with the
[Tidyverse][]. The bootstrap is among the methods the package supports.

[infer]: https://infer.netlify.app/
[Tidyverse]: https://tidyverse.org/
:::


## The Central Limit Theorem {#sec-central-limit-theorem}

The bootstrap is only possible because modern computers can resample a dataset
in a few seconds, but statisticians have needed to quantify uncertainty since
long before computers were invented. To do this, they developed mathematical
models of distributions and formulas for computing confidence intervals. These
methods are still widely used. Compared to the bootstrap, they have different
strengths and weaknesses, so it's useful to be familiar with both.

Before we can compute confidence intervals, we need to define some statistical
concepts. A **random variable** is an unrealized number or category that will
be drawn randomly from a specific distribution. Think of a random variable as
an observation that hasn't been collected yet. We might know how the population
is distributed, but we don't know the value of the observation. The statistical
convention is to write random variables in uppercase, like $X$ or $Y$, and
non-random values in lowercase, like $x$ and $y$.

The **normal distribution** is the centerpiece of many classical methods. Its
density curve is bell-shaped and extends to infinity in both the negative and
positive directions:

```{r}
#| echo: false
#| label: fig-standard-normal
#| fig-cap: >
#|     Density plot of the standard normal distribution. The mean is 0 and the
#|     standard deviation is 1.
ggplot() + xlim(-5, 5) + geom_function(fun = dnorm) + ylab("density")
```

The normal distribution is really a family of distributions parameterized by
their mean $\mu$ and variance $\sigma^2$. We denote these by $\mathcal{N}(\mu,
\sigma^2)$. All of them have the characteristic bell-shaped density. We call
the normal distribution with mean 0 and variance 1, denoted $\mathcal{N}(0,
1)$, the **standard normal distribution**.

Normal distributions are common because sums of random variables tend to be
nearly normal, especially if the number of terms in the sum is large. The
sample mean is an example of this. Suppose $X_1, X_2, \dots, X_n$ are
identically distributed random variables from a population with mean $\mu$ and
standard deviation $\sigma$. Suppose also that the value of each one is drawn
independently of the others. Then the sample mean $\bar{X}$, which estimates
the population mean $\mu$, is unbiased: its mean is $\mu$. Its standard error
is:

$$
\textrm{SE} = \frac{\sigma}{\sqrt{n}}.
$$

As the number of observations $n$ increases, the distribution of $\bar{X}$
becomes closer and closer to a normal distribution. Specifically,

$$
\bar{X} \to \mathcal{N}\left(\mu, \frac{\sigma^2}{n}\right)
\quad as \quad
n \to \infty.
$$

This result is known as the **Central Limit Theorem** (CLT) and is one of the
foundations of statistics. The CLT is important because it tells us that for
large samples, the sampling distribution of the mean is approximately normal.
We can use this to compute confidence intervals.

We can rearrange the CLT mathematically to write it in terms of a standard
normal distribution:

$$
\frac{\bar{X} - \mu} {\sigma / \sqrt{n}} \to \mathcal{N} (0, 1)
\quad as \quad
n \to \infty.
$$

Notice that the left-hand side is just the standard score
(@sec-standardization) of the sample mean.

:::{.callout-note}
There are other versions of the CLT that remove some of the assumptions. The
most important assumptions are that the random variables are independent and
that their standard deviations are finite.
:::

Most of the time, we don't know the population standard deviation $\sigma$. We
can replace it with the sample standard deviation $S$ to estimate the standard
error of the sample mean:

$$
\textrm{SE} = \frac{\sigma}{\sqrt{n}} \approx \frac{S}{\sqrt{n}}.
$$

For large samples (say $n \geq 30$), we can also replace $\sigma$ with $S$ in
the CLT. For small samples, $S$ tends to underestimate, so the standard score
is larger and doesn't have a standard normal distribution. The density is still
bell-shaped, but the tails are notably heavier. We call this **Student's t
distribution** (or the t distribution).

```{r}
#| echo: false
#| label: fig-t5
#| fig-cap: >
#|     Density plot of the standard normal distribution (solid) and Student's
#|     t distribution with 5 degrees of freedom (dotted).
ggplot() +
  xlim(-5, 5) +
  geom_function(fun = dt, args = list(df = 5), linetype = "dotted") +
  geom_function(fun = dnorm) +
  ylab("density")
```

Like the normal distribution, Student's t distribution is really a family of
distributions. The t distribution has just one parameter: **normality**, a
positive integer that controls how heavy the tails are. As normality increases,
the tails become lighter, and the t distribution becomes more and more like a
standard normal distribution. We denote the t distribution by by $t_v$, where
$v$ is normality. The normality parameter is also called **degrees of
freedom**.

After replacing $\sigma$ with $S$, the standard score of the sample mean has a
t distribution with normality $n - 1$. We can write this in mathematical
notation as:

$$
\frac{\bar{X} - \mu} {S / \sqrt{n}} \sim t_{n - 1}.
$$

The symbol $\sim$ means "is distributed as."

The distribution of the sample mean (and of its standard score) matters because
we can use it to construct a confidence interval. Specifically, because we know
the distribution, we can find a quantile $q$ that bounds the standard score on
both sides with probability $1 - \alpha$. That is,

$$
-q \leq \frac{\bar{X} - \mu}{S / \sqrt{n}} \leq q.
$$

We can rearrange the inequality to isolate the population mean $\mu$, since its
value is unknown to us:

$$
\bar{X} - q \cdot\frac{S}{\sqrt{n}}
\leq \mu
\leq \bar{X} + q \cdot\frac{S}{\sqrt{n}}.
$$

:::{.callout-note title="Note: Deriving the Confidence Interval" collapse="true"}
First, we multiply by $-S / \sqrt{n}$:

$$
q \cdot\frac{S}{\sqrt{n}}
\geq \mu - \bar{X}
\geq -q \cdot\frac{S}{\sqrt{n}}
$$

Written from smallest to largest, this is:

$$
-q \cdot\frac{S}{\sqrt{n}}
\leq \mu - \bar{X}
\leq q \cdot\frac{S}{\sqrt{n}}
$$

Finally, we add $\bar{X}$:

$$
\bar{X} - q \cdot\frac{S}{\sqrt{n}}
\leq \mu
\leq \bar{X} + q \cdot\frac{S}{\sqrt{n}}.
$$
:::

The endpoints of this inequality are random; the population mean $\mu$ is
non-random. To make this $1 - \alpha$ confidence interval concrete, we replace
the random values with their estimates:

$$
\left(
  \bar{x} - q \cdot \frac{s}{\sqrt{n}},
  \ \bar{x} + q \cdot \frac{s}{\sqrt{n}}
\right).
$$

In this case, $q$ is the $1 - \alpha / 2$ quantile of $t_{n - 1}$.

In R, the `qt` function computes quantiles of the t distribution. So the code
to construct a 95% confidence interval for the height differences example is:

```{r}
n = length(sample1)
std_err = sd(sample1) / sqrt(n)

# For a 0.95 = 1 - 0.05 confidence interval.
ci = mean(sample1) + qt(c(0.025, 0.975), n - 1) * std_err
round(ci, 4)
```

This t distribution confidence interval is similar to but a bit wider than the
bootstrap percentile interval (0.0112, 0.5188).

Compared to bootstrap percentile intervals, t distribution confidence intervals
have several advantages. They're cheaper to compute since no resampling is
needed. They also work relatively well even when the sample size is small.

The main limitation of t distribution confidence intervals is that they're
exclusively for the sample mean. The sampling distribution of a different
statistic will not necessarily be a t distribution. There are [some
methods][delta] to generalize the CLT to other statistics, but the bootstrap is
more flexible overall.

[delta]: https://en.wikipedia.org/wiki/Delta_method

:::{.callout-tip}
Knowing several different ways to compute confidence intervals (or any other
estimate) is a great way to check work. Compute them both ways and check
whether they agree. If they don't, go back and look for mistakes in the
computations.
:::

<!--
:::{.callout-tip title="Tip: Bootstrap t Intervals"}
:::
-->

:::{.callout-note title="See Also"}
[_Introduction to Modern Statistics_][ims] [@ims] introduces both the bootstrap
and classical methods based on the CLT.

[ims]: https://openintrostat.github.io/ims/

[_OpenIntro Statistics_][openintro] [@openintro] is an introductory statistics
textbook that exclusively uses classical methods based on the CLT. It provides
a more detailed look at these methods without letting rigor get in the way of
intuition.

[openintro]: https://www.openintro.org/book/os/
:::


